# Nature of Experimentation and Principles of Experimental Design (Data types and working with data) {#basic3}

**Duration:** 3-hour lab

## Learning Outcomes

Students should be able to:

1.  Demonstrate how to change data types/structures in R
2.  Demonstrate an ability to manipulate data frames to reflect a proper experimental design

## Data files for this lab

-   lab2.RData (containing your data frame from the previous lab)

-   infection.txt [Click here to download](https://o365cmu-my.sharepoint.com/:t:/g/personal/pimonrat_t_cmu_ac_th/IQB-_sr_C95ARLNJdJRhRVh9AU6i4_Tab5tHAHzKt2WAhKo?e=4tK81y)

-   tabledata.txt [Click here to download](https://o365cmu-my.sharepoint.com/:t:/g/personal/pimonrat_t_cmu_ac_th/IQA8dUUXG09RRbU3kmIkVyKBAVBv8zemwvnpLvocNx3y8CQ?e=8B2aic)

## Introduction

This worksheet will provide you with an understanding of data types and how to manipulate data frames. Read the text for explanations and input the text in the box into R (copy and paste into R is fine.)

## Nature of Experimentation and Data

In experiments, you deal with many types of data. Make sure you know what your dependent and independent variables are. For example, your treatment (independent variable) may be categorical treatments like types of fertilizers, species of trees, etc. In addition, your measurements (dependent variable) may be categorical (such as death or alive, colors, etc.), count (such as number of leaves, number of species, etc.), and/or continuous values (such as weight, length, etc.). Once it comes to analyzing your data, you should know the R data types you work with so you will be able to correctly specify your data types in R. Here are six data types:

1.  **Integer**: Holds whole numbers (no decimal point), e.g. 3, 65, 99
2.  **Character**: Holds letters or symbols, e.g. 'a', 'Thailand'
3.  **Numeric**: Stores decimal or fraction values, e.g. 0.24, 100, 10.345
4.  **Logical**: Holds true or false values
5.  **Complex**: The mathematical concept of an imaginary number e.g. 2+6i. We rarely use this type
6.  **Raw**: ray bytes of data e.g. 62 6f 62 We rarely use this in practice

## Data Structure in R

R has six data structures:

1.  **Vector**: holds one type of data (homogenous). The data can be any of the six types above. Use `c()` function to create a vector
2.  **List**: holds multiple types of data (heterogenous). Use `list()` function to create a list
3.  **Matrix**: two-dimensional, homogeneous data structure. A matrix contains a number of rows and columns. Use `matrix()` function to create a matrix
4.  **Data frame**: two-dimensional, heterogeneous data structure. Each column has the same data type. Use `data.frame()` function to create a data frame or import data from files
5.  **Array**: three-dimensional, homogeneous data structure. Use `array()` function to create a matrix
6.  **Factor**: A vector that stores categorical data. Use `factor()` function to create a factor. For example, if you have three treatments -- control, fertilizer a, and fertilizer b, your factor would specify your factor in R as treatment \<- `factor("control", "a", "b")`

**After importing the data into R, make sure that each variable has appropriate data type and/or structure**

## Loading R Workspace and Inspect Data Structures

```{r eval=F}
load("lab2.RData")
str(dat.no.na)
```

### Checking Data Types/Structure of Each Variable of the lab2 data set

| Variable  | Current data type/structure | Appropriate data type/structure |
|-----------|-----------------------------|---------------------------------|
| count     |                             |                                 |
| sample    |                             |                                 |
| Orobanche |                             |                                 |
| extract   |                             |                                 |
| percent   |                             |                                 |

## Changing Data Types/Structure

Orobanche and extract are character data type. In practice, they are a factor type. In our case, the types of column variable can be changed from the character type to the factor using a function called `as.factor()`

```{r eval=F}
dat.no.na$Orobanche <- as.factor(dat.no.na$Orobanche)
dat.no.na$extract <- as.factor(dat.no.na$extract)
str(dat.no.na)
```

Now check to see if all variables have appropriate data type/structure.

**After importing the data into R, make sure that each variable has appropriate data type and/or structure**

## Removing Some Columns

```{r eval=F}
new.dat <- within(dat.no.na, rm(sample)) # 1 build a new data frame named 'new.data'
new.dat
```

We have created a new data frame called 'new.dat' which is a subset of the existing data frame 'dat.no.na'.

The other way is to make changes to your existing data frame. With this option, your data frame will be completely modified, and you will not be able to undo the changes. Here the 2nd column, 'sample' will be removed from the 'dat.no.na' data frame.

```{r eval=F}
dat.no.na <- dat.no.na[, -2]
str(dat.no.na)
```

See Chapter 4 in Crawley 2012 [@crawley2012] (PDF provided in class) for more commands you can do with a data frame and try it in R.

## Ordering Your Data

We use `order()` function to order the rows by a variable.

```{r eval=F}
## order by the percent
dat.no.na[order(dat.no.na$percent), ] 

## order first by the extract, then the percent
dat.no.na[order(dat.no.na$extract, dat.no.na$percent), ] 
```

## Changing Name of Variable (Column) and Row

We use the `colnames()` function to change a name. The code asks to change the name of the 'percent' column to 'final'

```{r eval=F}
colnames(dat.no.na)[which(names(dat.no.na) == "percent")] <- "final "
head(dat.no.na)
```

Changing row names is a bit tricky. Remind yourself that each row should have a unique name (no repetition). Let's try changing row names to be 'sample' followed by a row number

```{r eval=F}
rownames(new.dat) <- paste("sample",1:nrow(new.dat))
new.dat
```

## Removing a Data Frame

To remove an entire data frame, use `rm()` function.

```{r eval=F}
rm(new.dat)
new.dat
```

When you call 'new.dat' you will get Error: object 'new.data' not found. This means that the data frame is already removed.

## Subsetting Data Frame

Sometimes we would like to only make a smaller data frame from an existing one. In this example, we only want to use data from only one extract, bean. We will start from the original data set, the 'dat' data frame. The codes below show two ways to subset the data frame.

```{r eval=F}
dat.bean <- subset(dat, dat$extract == 'bean') # 1
str(dat.bean)

dat.bean2 <- dat[which(dat$extract == 'bean'), ] # 2

# Ask if the two are identical
identical(dat.bean, dat.bean2) 

# Remove unused levels of factors
dat.bean <- droplevels(dat.bean) 
```

## Other Types of Data Sets That You May Work With

We may also record data of each individual in each row then we would like to summarize by counting or grouping data. We will use cross tabulation to handle this. Let's first import the data.

```{r eval=F}
infection <- read.table('infection.txt', header = T)
infection
head(infection)
str(infection)
summary(infection)
```

This is a dataset of individual patients who had or had no diseases. Each row has data of each patient. We would like to create a data frame containing counts according to variables of interest. In R the function `table()` will create a summary table of count.

```{r eval=F}
# We summarize infected data by gender
table(infection$infected, infection$sex) 

# add sum of cases
addmargins(table(infection$infected, infection$sex))
```

In contrast, we also can create individual case data from cross tabulation data. See the example below.

```{r eval=F}
count.table <- read.table('tabledata.txt', header = T)
count.table
```

In this data frame you will see that this is a count dataset. You would like to make a new data frame of each individual case, i.e. you count 12 healthy young males and you would like to make each row representing each male. At the end, you would like to have 12 rows of healthy young male and 7 rows of healthy old male and so on. We will use the `lapply()` function.

```{r eval=F}
lapply(count.table, function(x) rep(x, count.table$count))

## see what type of this data in R
class(lapply(count.table, function(x) rep(x, count.table$count))) 
```

At this point it is not a data frame yet. It is a list. We will change it to be a data frame by

```{r eval=F}
expand.tab <- as.data.frame(lapply(count.table, function(x) rep(x, count.table$count)))
head(expand.tab)

# remove the first column
expand.tab <- expand.tab[ ,-1]
head(expand.tab)

# add individual ID to each case
expand.tab$id <- seq(1, length(expand.tab[, 1]))
class(expand.tab)
```

## Exercise

**Submission:** A PDF file containing your answers. If you are asked to work in R, your answers must contain **R codes** that you wrote and **the outputs** you get. Your codes and outputs can be text and/or screen capture.

**Full score:** 10

**Data for this exercise:** hw3 data.csv [Click here to download](https://o365cmu-my.sharepoint.com/:x:/g/personal/pimonrat_t_cmu_ac_th/IQC7o7YrBKY5Rr97OtI9pzAuAbxfptYkVzSEz6wY2O0dZrQ?e=EFJKEA)

This is a dataset from a seed storage experiment. Seeds were collected and stored. And then the seeds were sown to evaluate germinability. There are four treatments of storing seeds under dry (two levels) vs moist (two levels) conditions and an initial germination (baseline).

### Questions/Directions

**1.** In R, create a data frame in R named 'seed', and then import a file 'hw3 data.csv'. Show the structure of the data frame.

**2.** Change the data structure of 'treatment' and 'replicate' to be a factor data structure.

**3.** Subset the data frame to include only dry20, dry40, wet20 and wet40 treatment (omit baseline treatment) and name the new data frame 'seed.treat'. Show the structure of the new data.

**4.** From the 'seed.treat' data frame, create a summary table of the count number of germinated vs ungerminated seeds (germ.or.not column) and the treatments (treatment). Show the output.

**5.** Answer these questions (no R code needed)

5.1. How many seeds were used per treatment?

5.2. Which treatment has the most germinated seeds?

### Grading Rubric

| Question | Score 2                         | Score 1                                              | Score 0                                         |
|-------------------|------------------|------------------|------------------|
| 1        | Correct R code and output       | One of R code or output is not included or incorrect | R code and output are not included or incorrect |
| 2        | Correct R code and output       | One of R code or output is not included or incorrect | R code and output are not included or incorrect |
| 3        | Correct R code and output       | One of R code or output is not included or incorrect | R code and output are not included or incorrect |
| 4        | Correct R code and output       | One of R code or output is not included or incorrect | R code and output are not included or incorrect |
| 5        | Correctly answer both questions | One answer is included or correct                    | No answer                                       |

**End of Lab 3**

------------------------------------------------------------------------
