# Probability distributions {#distributions}

**Duration:** 2-hour lecture

## Learning Outcomes

Students should be able to:

1. Name some discrete probability distributions and give some examples of data
2. Name some continuous probability distributions and give some examples of data

## Introduction

Probability distributions are the basis for most statistics. Learning probability distributions is beneficial because it makes us understand the nature of data we have and allows us to select appropriate statistics for data analysis and make predictions. In this lesson, we will explore the various types of probability distributions – discrete and continuous ones. Let's start with continuous probability distributions.

## Continuous Probability Distribution

We will learn three examples of continuous probability distribution.

### Normal Distribution

Normal distribution is also called Gaussian distribution, Gauss distribution, and Laplace-Gauss distribution. A random variable, $X$, follows a normal distribution with two parameters - $\mu$ and $\sigma^2$ parameter:

$$X \sim N(\mu, \sigma^2)$$

The distribution is characterized by:

$$E(X) = \mu$$
$$Var(X) = \sigma^2$$
$$SD(X) = \sigma$$

#### Properties of Normal Distribution

- The range is from $-\infty, +\infty$
- Symmetric around the mean (mode and median)
- It is unimodal
- It follows a **68-95-99.7 rule** (Figure \@ref(fig:normal-distribution) meaning that:
  - 68% of the data points is within the area between $\mu - \sigma$ and $\mu + \sigma$ (1 standard deviation)
  - 95% of data points is within the area between $\mu - 2\sigma$ and $\mu + 2\sigma$ (2 standard deviations)
  - 99.7% of the data points is within the area between $\mu - 3\sigma$ and $\mu + 3\sigma$ (3 standard deviations)

```{r normal-distribution, fig.cap="The 68-95-99.7 rule of the normal distribution with mean of 0 and standard deviation of 1", echo = FALSE}
# Generate normal distribution
x <- seq(-4, 4, length.out = 1000)
y <- dnorm(x, mean = 0, sd = 1)

# Plot
plot(x, y, type = "l", lwd = 2, col = "blue",
     xlab = "Value", ylab = "Density",
     main = "Normal Distribution (μ = 0, σ = 1)")

# Add shaded regions for 68-95-99.7 rule
polygon(c(x[x >= -1 & x <= 1], 1, -1), 
        c(y[x >= -1 & x <= 1], 0, 0), 
        col = rgb(0, 0, 1, 0.2), border = NA)
polygon(c(x[x >= -2 & x <= 2], 2, -2), 
        c(y[x >= -2 & x <= 2], 0, 0), 
        col = rgb(0, 0, 1, 0.1), border = NA)
polygon(c(x[x >= -3 & x <= 3], 3, -3), 
        c(y[x >= -3 & x <= 3], 0, 0), 
        col = rgb(0, 0, 1, 0.05), border = NA)

# Add vertical lines
abline(v = c(-3, -2, -1, 0, 1, 2, 3), lty = 2, col = "gray")
legend("topright", legend = c("68%", "95%", "99.7%"), 
       fill = rgb(0, 0, 1, c(0.2, 0.1, 0.05)), bty = "n")
```

#### Examples of Data in Ecology and Environmental Science

1. **Tree height in a forest stand.** Tree height within a homogenous forest stand often follows a normal distribution, as environmental factors like soil, water, and sunlight are uniformly distributed. For example, @zhang2015 studied tree height and diameter in boreal forests.

2. **Body size of animal populations.** The body size of animals such as fish or birds in a controlled habitat tends to be normally distributed. For example, @bonner2011.

3. **Plant leaf size.** Leaf length or area within a specific plant species growing in a uniform environment typically exhibits a normal distribution. For example, @nicotra2010 studied leaf phenotypic plasticity in *Centaurea melitensis*.

4. **Water quality measurements in stable lakes.** In stable, well-mixed lakes, parameters such as dissolved oxygen concentrations or pH over time can follow a normal distribution [@wetzel2001].

5. **Rainfall distribution over short timeframes.** Daily amount of rainfall in a specific region over short periods (e.g., a month) often approximate a normal distribution, especially in regions with consistent weather patterns [@wilks2006].

### Continuous Uniform Distribution

Uniform distribution is also called the rectangular distribution (Figure \@ref(fig:uniform-distribution)). It represents equal probability of an event occurring over a range of values. A random variable, $X$, follows a uniform distribution with two parameters - $a$ and $b$ parameter:

$$X \sim \text{Unif}(a, b)$$

The distribution is characterized by:

$$E(X) = \frac{1}{2}(a + b)$$
$$Var(X) = \frac{1}{12}(b - a)^2$$

#### Properties of Uniform Distribution

- The range is from $-\infty, +\infty$
- Symmetric around the mean (mean and median are the same) (Figure \@ref(fig:uniform-distribution))
- All outcomes in the range of distribution have an equal probability of occurring
- The probability of occurring is zero outside the interval $a$ and $b$

```{r uniform-distribution, fig.cap="An example of uniform distribution with a = 0, b = 2", echo = FALSE}
# Generate uniform distribution
x <- seq(-0.5, 2.5, length.out = 1000)
y <- dunif(x, min = 0, max = 2)

# Plot
plot(x, y, type = "l", lwd = 2, col = "red",
     xlab = "Value", ylab = "Density",
     main = "Uniform Distribution (a = 0, b = 2)",
     ylim = c(0, 0.6))
abline(v = c(0, 2), lty = 2, col = "gray")
```

#### Examples of Data in Ecology and Environmental Science

1. **Seed dispersion in designed experimental plots.** When seeds are scattered evenly over an experimental plot, their locations follow a uniform distribution within the plot boundaries [@harper1977].

2. **Sampling time for environmental monitoring.** Environmental data such as air quality or water quality are often sampled randomly within a fixed time frame such as at 8 am – 6 pm. The sampling times can follow a uniform distribution [@ott1978].

### Lognormal Distribution

A probability distribution of a random variable whose logarithm is normally distributed. If $X$ has a log-normal distribution, then $\ln(X)$ has a normal distribution. A random variable, $X$, follows a lognormal distribution with two parameters – $\mu$ (the mean of the logarithm of the variable) and $\sigma^2$ (the variance of the logarithm of the variable) parameter:

$$X \sim \text{Lognormal}(\mu, \sigma^2)$$

The distribution is characterized by:

$$E(X) = \exp(\mu + \sigma^2/2)$$
$$Var(X) = [\exp(\sigma^2) - 1] \cdot \exp(2\mu + \sigma^2)$$

#### Properties of Log-Normal Distribution

- The log-normal distribution is only for positive data ($X > 0$) (Figure \@ref(fig:lognormal-distribution))
- It is unimodal and asymmetrical. A peak is near the lower bound and a long tail is toward higher values (right skewed)

```{r lognormal-distribution, fig.cap="Lognormal distributions with different means (dashed lines)", echo = FALSE}
# Generate lognormal distributions
x <- seq(0.01, 10, length.out = 1000)
y1 <- dlnorm(x, meanlog = 0, sdlog = 0.5)
y2 <- dlnorm(x, meanlog = 1, sdlog = 0.5)
y3 <- dlnorm(x, meanlog = 0, sdlog = 1)

# Plot
plot(x, y1, type = "l", lwd = 2, col = "blue",
     xlab = "Value", ylab = "Density",
     main = "Lognormal Distributions",
     ylim = c(0, max(y1)))
lines(x, y2, lwd = 2, col = "red")
lines(x, y3, lwd = 2, col = "green")

# Add mean lines
abline(v = exp(0 + 0.5^2/2), lty = 2, col = "blue")
abline(v = exp(1 + 0.5^2/2), lty = 2, col = "red")
abline(v = exp(0 + 1^2/2), lty = 2, col = "green")

legend("topright", 
       legend = c("μ = 0, σ = 0.5", "μ = 1, σ = 0.5", "μ = 0, σ = 1"),
       col = c("blue", "red", "green"), lwd = 2, bty = "n")
```

#### Examples of Log-Normal Distribution in Ecology and Environmental Science

1. **Body sizes of animals or plants within a species.** Tree diameters [@enquist2001] often follow a log-normal distribution due to multiplicative growth processes.

2. **Concentrations of pollutants in the environment.** The levels of pollutants, such as heavy metals in soil or air pollution particulates, often follow a lognormal distribution because of the accumulation and dilution processes involved [@ott1990].

3. **Species abundance in ecological communities.** The number of individuals per species in an ecosystem often follows a log-normal distribution. It shows that some species are common while some species are rare [@preston1948].

## Discrete Probability Distribution

We will learn two examples of discrete probability distribution.

### Poisson Distribution

The Poisson distribution models discrete events that occur in a fixed interval of time and space. The assumptions of Poisson distribution are:

1. The rate of occurrence is constant
2. The occurrence of events is independent
3. Two events cannot occur at the same time

A random variable, $X$, follows a Poisson distribution with one parameter - $\lambda$ parameter (lambda):

$$X \sim \text{Pois}(\lambda)$$

The distribution is characterized by:

$$E(X) = \lambda$$
$$Var(X) = \lambda$$

#### Properties of Poisson Distribution

- The Poisson distribution is for natural numbers starting from 0 (Figure \@ref(fig:poisson-distribution))
- It is unimodal. The skewness depends on the $\lambda$ parameter
- The mean and variance of Poisson distribution are equal

```{r poisson-distribution, fig.cap="Poisson distributions with different values of lambda. Dashed lines show the means", echo = FALSE}
# Generate Poisson distributions
par(mfrow = c(1, 1))
b <- 0:12

plot(b, dpois(b, 1),type ='o',
     main = "Different lambda", lwd = 1.5,
     xlab = "Value", ylab = "Probability")
lines(b,dpois(b, 2),type ='o', col = 2, lwd = 1.5)
lines(b,dpois(b, 3),type ='o', col = 3, lwd = 1.5)

legend("topright", legend = c("1","2","3", "mean = lambda"), col = c(1:3,1),
       lwd = 1.5, lty = c(1, 1, 1, 2))

for(i in 1:3){
  abline(v = i, col = i, lty = 2, lwd = 1.5)
  text(i + 0.2, 0.35,
       label = i, col = i)
}
```

#### Examples of Poisson Distribution in Ecology and Environmental Science

1. **Number of animals observed per area** [@buckland1993]. The count of animals, such as birds or insects, observed in a fixed survey area during a specific time follows a Poisson distribution.

2. **Number of trees or seedlings in a surveyed plot** [@khamyong2018].

3. **Counts of pollinators' visits to flowers.** The number of pollinators visiting individual flowers in a set amount of time is often Poisson-distributed [@ricketts2004].

4. **Rainfall-induced landslides in a region** [@guzzetti2007]. The number of landslides after rainfall events in a fixed region during a particular time interval often follows Poisson distribution.

### Binomial Distribution

The binomial distribution models the number of successes in a sequence of independent trials. The assumptions of binomial distribution are:

1. The number of trials is fixed
2. The outcome is binary, meaning there are two options i.e., success and failure
3. For each trial, the probability of success is constant

A random variable, $X$, follows a binomial distribution with two parameters – $n$ and $p$ parameter. Parameter $n$ represents the number of trials and $p$ represents the probability of success:

$$X \sim \text{Bin}(n, p)$$

The distribution is characterized by:

$$E(X) = np$$
$$Var(X) = np(1-p)$$

#### Properties of Binomial Distribution

- The number of successes can be from 0 to $n$ (Figure \@ref(fig: binomial-distribution))
- It is unimodal. The skewness depends on the $p$ parameter
- For large number of trials, $n$, the distribution approaches a normal distribution
- For the probability of success is 0.5, the distribution is symmetric

```{r binomial-distribution, fig.cap="Binomial distributions with same probability of success at p = 0.5 and different number of trials (n). Dashed lines show the means", echo = FALSE}

# Generate binomial distributions
par(mar = c(5, 4, 1, 1))
a <- seq(0, 20)
par(mfrow = c(1, 1))

## vary size parameter
size <- c(1, 5, 20)
plot(a, dbinom(a, size[1], 0.5), type = 'o',
     main = "Different number of trials (p = 0.5)",
     xlim = c(0, 20), lwd = 1.5, xlab = "Value", ylab = "Probability")

for(i in 2:3){
  lines(a,dbinom(a, size[i], 0.5), type ='o', col = i,
        lwd = 1.5)
}

legend("topright",legend = c("1","5","20", "mean = np"), col = c(1:3, "black"),
       lty = c(1, 1, 1, 2), lwd = 1.5)

for(i in 1:3){
  abline(v = 0.5 * size[i], col = i, lty = 2, lwd = 1.5)
  text((0.5 * size[i]) + 0.6, 0.4,
       label = 0.5 * size[i], col = i)
}

```

#### Examples of Binomial Distribution in Ecology and Environmental Science

1. **Seed germination data** (e.g., @tiansawat2013, @yiamthaisong2024). There are two outcomes for each seed – germinate or does not germinate.

2. **Seedling survival (success or failure)** (e.g., @naruangsri2023). Dead or alive seedlings are recorded and can be modeled using a binomial response.

3. **Occurrence of some plant characteristics.** Often, we are interested in the presence and absence of some species traits, for example, light dependency of seed germination (e.g., @tiansawat2013). We can model the light dependency of different species (need light or do not need light) as a binomial response.

## Exercises

Fill in this following table to complete information of probability distributions. Then do more research and give two other examples of probability distributions.

| Probability Distribution | Discrete or Continuous | Range of Data | Example 1 | Example 2 |
|-------------------------|------------------------|---------------|-----------|-----------|
| Normal | | | | |
| Uniform | | | | |
| Lognormal | | | | |
| Binomial | | | | |
| Poisson | | | | |
| | | | | |
| | | | | |

## References

::: {#refs}
:::

