# Random Variables {#lab4randomvar}

**Duration:** 3-hour lab

## Learning Outcomes

Students should be able to:

1. Generate plots of random variables and describe the shape of the distribution

## Data Files for This Lab

None to download. A data set is generated in the worksheet

## Introduction

This worksheet will provide you with an understanding of random variables and probability distributions.
We will use a normal distribution in this lab. The workflow of this lab is inspired by the exercise sheet from the class the instructor took when she was in a graduate program [@ib519]. The workflow is modified to be more targeting the normal distribution. Read the text for explanations and input the text in the box into R (copy and paste into R is fine.) 

## Normal Distribution

A normal distribution is one of the most important distributions [@navarro2024].
It is also called Gaussian distribution.
It is described using two parameters, the mean of the distribution $μ$ and the standard deviation of the distribution $σ$.

$$X ∼ Normal(μ, σ)$$

In R, the function for the normal distribution

| Function | Description |
|----------|-------------|
| dnorm() | probability density function |
| pnorm() | cumulative distribution function |
| qnorm() | calculates the value associated with a specified tail probability, inverse of "p" |
| rnorm() | simulates random numbers |

## Generating Random Numbers

Let's start with rnorm to generate random number from a normal distribution

```{r eval=F}
set.seed(123)
x <- rnorm(100, mean = 0, sd = 1)

## draw a histogram of x
par(mfrow = c(1, 1))
hist(x)
abline(v = mean(x), col = 'purple')
```

Let's see mean, median and standard deviation of this data

```{r eval=F}
mean(x)
median(x)
sd(x)

# Add mean and median into the plot
abline(v = mean(x), lty = 2, lwd = 1.5, col = "purple")
abline(v = median(x), lty = 2, lwd = 1.5, col = 'darkorange')
```

## Density Plot

Create a density plot.
See the y axis.
It is now showing probability.

```{r eval=F}
hist(x, probability = TRUE, axes = T, xlim = c(-3, 3))

## Add a density line
lines(density(x), col = "red", lwd = 2)

## Now draw a normal distribution line
## with mean and SD of x
x2 <- seq(min(x), max(x), length = 50)
f <- dnorm(x2, mean = mean(x2), sd = sd(x2))

lines(x2, f, col = "blue", lwd = 2)
```

## Overlaying Boxplot on Histogram

Let's overlay a boxplot onto a histogram.
Box plots are a good visualization of the data distribution.
Pay attention on the shape of the box plot and the histogram.

```{r eval=F}
hist(x, probability = TRUE, axes = T, xlim = c(-3, 3))
par(new = TRUE)
boxplot(x, horizontal = TRUE, axes = FALSE,
        lwd = 1, col = rgb(0, 1, 1, alpha = 0.15),
        ylim = c(-3, 3))
points(mean(x), 1, col = 'purple', pch = '*', cex = 2)

stripchart(x, method = "jitter", pch = 19, cex = 0.5,
           add = TRUE, col = "blue")
```

## Effect of Changing Mean and Standard Deviation

Next we will see when we change mean and sd and the effect on the shape of normal distribution.
Let's start by generating a data frame with a sequence of values

```{r eval=F}
a = seq(-5, 5, length.out = 100)
mean(a)

plot(a, dnorm(a), bty = 'n',
     type = 'l', main = 'When the mean is changed.')
abline(v = 0) ## add a line to indicate where the mean is
lines(a, dnorm(a, 2), col = 2) ## changing the mean
abline(v = 2, col = 2)
lines(a,dnorm(a, -1), col = 3) ## changing the mean
abline(v = -1, col = 3)

plot(a, dnorm(a), type = 'l', bty = 'n',
     main = 'When the SD is changed')
abline(v = 0) ## add a line to indicate where the mean is
lines(a,dnorm(a, sd = 2),col = 2) ## changing the sd
abline(v = 0, col = 2)
lines(a,dnorm(a, sd = 3),col = 3) ## changing the sd
abline(v = 0, col = 3)

plot(a, dnorm(a), bty = 'n',
     type ='l', main = 'When the mean and SD are changed')
abline(v = 0)
lines(a, dnorm(a, -2, 1.5), col = 2) ## changing the mean and standard dev
abline(v = -2, col = 2)
lines(a, dnorm(a, 1, 2), col = 3) ## changing the mean and standard dev
abline(v = 1, col = 3)
```

## Cumulative Density Function (CDF) vs Probability Density Function (PDF)

```{r eval=F}
par(mfrow = c(2, 1))

plot(a, dnorm(a), type = 'l', bty = 'n',
     main = 'Probability density function of x',
     ylim = c(0, 1))
abline(v = 0)

plot(a, pnorm(a), type = 'l', bty = 'n',
     main = 'Cumulative density function of x',
     ylim = c(0, 1))
abline(v = 0)
```

## Sample Size and the Shape of the Distribution

Now we will see the histograms of data with different sample sizes and compare to the actual density of the standard normal.

```{r eval=F}
par(mfrow = c(2, 2))
n = c(5, 30, 500, 10000)
for(i in 1:4){
  hist(rnorm(n[i]), main = n[i],
       probability = TRUE,
       breaks = 40, xlab = 'Value')
  lines(a, dnorm(a), col = 2)
}
```

The example above shows that as the number of random draws from a probability distribution increases the histogram of those draws provides a better approximation of the density itself.

Use the code below to determine the mean and standard deviation of random draws of small and large sample sizes.
Notice that when the sample size increases the sample mean and standard deviation get closer to the true mean of 0 and standard deviation of 1 (mean = 0, sd = 1).
The standard error gets smaller when the sample size increases.

```{r eval=F}
b <- rnorm(10)
mean(b)
sd(b) ## standard deviation
sd(b)/sqrt(10) ## standard error

c <- rnorm(10000)
mean(c)
sd(c) ## standard deviation
sd(c)/sqrt(10000) ## standard error

save.image("lab4.RData")
```

## Exercise

**Submission:** A PDF file containing your answers. If you are asked to work in R, your answers must contain **R codes** that you wrote and **the outputs** you get. Your codes and outputs can be text and/or screen capture.

**Full score:** 10

**Data for this exercise:** Run the following code.

```{r eval=F}
set.seed(123)
hw4 <- c(rnorm(300, 1, 2), rnorm(300, 10, 2.5))
```

### Questions/Directions

**1.** In R, plot a histogram of the hw4 and show the plot.

**2.** Explain in words the shape of the distribution

**3.** Use R to find the mean and standard deviation of this data?

**4.** In R, plot a box plot of the hw4 and show the plot.

**5.** Explain in words what the limitation of the box plot to represent the shape of the distribution

### Grading Rubric

| Question | Score 2 | Score 1 | Score 0 |
|----------|---------|---------|---------|
| 1 | Correct R code and output | One of R code or output is not included or incorrect | R code and output are not included or incorrect |
| 2 | Correctly describe the shape in full sentence | Correctly describe the shape in but not in full sentence | No answer |
| 3 | Correct R code and outputs | One of R code or output is not included or incorrect | R code and output are not included or incorrect |
| 4 | Correct R code and output | One of R code or output is not included or incorrect | R code and output are not included or incorrect |
| 5 | Correctly explain in full sentence | Correctly explain but not in full sentence | No answer |

**End of Lab 4**