# Probability Distributions {#lab5prob}

**Duration:** 3-hour lab

## Learning Outcomes

Students should be able to:

1.  Generate plots of some common distributions and describe the important properties of them

## Data file for this lab

None to download.
Datasets are generated in the worksheet.

## Introduction

This worksheet will guide you to learn two each of continuous and categorical probability distributions.
The workflow of this lab is inspired by the exercise sheet from the class the instructor took when she was in a graduate program [@deitze2009].
Read the text for explanations and input the text in the box into R (copy and paste into R is fine.)

In R, there are a lot of probability distributions by default.
Here is a list of some common distributions with the R function and the parameters of each of the distributions.

| Distribution      | R name  | Parameters (arguments in function) |
|-------------------|---------|------------------------------------|
| Beta              | beta    | shape1, shape2, ncp                |
| Binomial          | binom   | size, prob                         |
| Cauchy            | cauchy  | location, scale                    |
| Chi-squared       | chisq   | df, ncp                            |
| Exponential       | exp     | rate                               |
| F                 | f       | df1, df2, ncp                      |
| Gamma             | gamma   | shape, scale                       |
| Geometric         | geom    | prob                               |
| Hypergeometric    | hyper   | m, n, k                            |
| Log-normal        | lnorm   | meanlog, sdlog                     |
| Logistic          | logis   | location, scale                    |
| Negative binomial | nbinom  | size, prob                         |
| Normal            | norm    | mean, sd                           |
| Poisson           | pois    | lambda                             |
| Student's t       | t       | df, ncp                            |
| Uniform           | unif    | min, max                           |
| Weibull           | weibull | shape, scale                       |
| Wilcoxon          | wilcox  | m, n                               |

## Continuous Distribution

### Uniform Distribution

Two bounding parameters are maximum and minimum numbers [@mathworks].
Out of this bound, the probability is 0.
Run the codes to learn about how the shape of the distribution changes with changing the maximum and minimum numbers.

```{r eval=F}
x <- seq(-5, 5, by = 0.1)
par(mfrow = c(1, 1), mar = c(2, 4, 1, 1))

# Uniform distribution
plot(x, dunif(x)) ## points of data
lines(x, dunif(x)) ## use line option instead

# Change parameters of the distribution
# min and max
# Look at the changes in shape
plot(x, dunif(x), type = 'l')
lines(x, dunif(x, min = 0, max = 4), col = 'red')
lines(x, dunif(x, min = -3, max = -0.5), col = 'blue')

## CDF of uniform distribution
plot(x, punif(x), type = 'l')
lines(x, punif(x, 0, 4), col = 'red')
lines(x, punif(x, -3, -0.5), col = 'blue')

hist(runif(300,-3 , 3),
     breaks = 30, xlim = range(x),
     probability = TRUE)
lines(x, dunif(x, -3, 3), col = 'red')
```

### Log Normal Distribution

This is the probability distribution of the data whose logarithm has a normal distribution.
Look at the shape of the distribution when the parameters are changed.

```{r eval=F}
## changing the meanlog, while fixing sdlog = 1 (default)
y <- 10^seq(-2, 2, by = 0.01) ## original data
par(mfrow = c(2, 1))

plot(y, dlnorm(y, 0), type = 'l',
     xlim = c(0, 15),
     main = "Changing the mean")
lines(y, dlnorm(y, 1), col = 'red')
lines(y, dlnorm(y, 2), col ='blue')

## show means in the plot
abline(v = 0, col = 1, lty = 2)
abline(v = 1, col = 'red', lty = 2)
abline(v = 2, col = 'blue',lty = 2)

## on a log scale = take log of original data
plot(y, dlnorm(y, 0),type = 'l',
     log = 'x',main = "Plot in log scale")
lines(y, dlnorm(y, 1), col = 'red')
lines(y, dlnorm(y, 2), col = 'blue')
abline(v = exp(0), col = 1, lty = 2)
abline(v = exp(1), col = 'red', lty = 2)
abline(v = exp(2), col= 'blue', lty = 2)

## changing the variance
par(mfrow = c(1, 1))
plot(y,dlnorm(y, 2, 0.125),
     type = 'l', xlim = c(0, 20),
     ylim = c(0,0.6), main = "Changing the variance")
lines(y, dlnorm(y, 2, 0.25), col ='red')
lines(y, dlnorm(y, 2, 0.5), col = 'blue')
lines(y, dlnorm(y, 2, 1), col = 'brown')
lines(y, dlnorm(y, 2, 2), col = 'green')
lines(y, dlnorm(y, 2, 4), col = 'magenta')
abline(v = exp(2), col = 1, lty = 2)

## random sample
# draw 250 random numbers from
# a log normal distribution with
# meanlog = 2, sdlog = 1
par(mfrow = c(1, 1))
hist(rlnorm(250, 2, 1), breaks = 30,
     probability = TRUE, ylim = c(0, 0.1))
lines(y, dlnorm(y, 2, 1), col = "purple")
abline(v = 2, col = "purple", lty = 2)
```

## Discrete Distribution

### Binomial Distribution

The binomial arises naturally as the counts of the number of successes given a probability, $p$, and a sample size, $n$ [@mathworks2].

```{r eval=F}
a <- seq(0, 20)
par(mfrow = c(1, 1))

## vary size parameter
size <- c(1, 5, 20)
plot(a, dbinom(a, size[1], 0.5), type = 'o',
     main = "Different number of trials (p = 0.5)",
     xlim = c(0, 20), lwd = 1.5)

for(i in 2:3){
  lines(a,dbinom(a, size[i], 0.5), type ='o', col = i,
        lwd = 1.5)
}

legend("topright",legend = c("1","5","20", "mean = np"), col = c(1:3, "black"),
       lty = c(1, 1, 1, 2), lwd = 1.5)

for(i in 1:3){
  abline(v = 0.5 * size[i], col = i, lty = 2, lwd = 1.5)
  text((0.5 * size[i]) + 0.6, 0.4,
       label = 0.5 * size[i], col = i)
}

## vary probability
p <- c(0.1, 0.5, 0.8)
plot(a, dbinom(a, 10, p[1]), type='o', main = "Different probabilities (size = 10)",
     col = 1, lwd = 1.5)

for(i in 2:3){
  lines(a, dbinom(a, 10, p[i]),col = i,
        lwd = 1.5 ,type = 'o')
}

legend("topright", legend = c("0.1","0.5","0.8", "mean = np"), col = c(1:3, 1),
       lty = c(1, 1, 1, 2), lwd = 1.5)

for(i in 1:3){
  abline(v = p[i] * 10, col = i, lty = 2, lwd = 1.5)
  text((p[i] * 10) + 0.5, 0.35,
       label = p[i] * 10, col = i)
}

## random draws from a binomial distribution
par(mfrow = c(1, 1))
## we draw 100 times from the binomial distribution
## with parameter trial = 10, prob. of success
## = 0.5
hist(rbinom(100, 10, 0.5) + 0.0001, ## small amount added because
     probability = TRUE, ## of way R calculates breaks
     breaks = 0: 11, main = "Random draws from binomial")
lines(a, dbinom(a, 10, 0.5), type = 's',col = 2, lwd = 2)
```

### Poisson Distribution

The Poisson distribution is common for count data (integer).
The events occur in a fixed amount of time or the number of discoveries in a fixed space.
There is one parameter, rate (lambda: $λ$).

```{r eval=F}
par(mfrow = c(1, 1))
b <- 0:12

plot(b, dpois(b, 1),type ='o',
     main = "Different lambda", lwd = 1.5)
lines(b,dpois(b, 2),type ='o', col = 2, lwd = 1.5)
lines(b,dpois(b, 3),type ='o', col = 3, lwd = 1.5)

legend("topright", legend = c("1","2","3", "mean = lambda"), col = c(1:3,1),
       lwd = 1.5, lty = c(1, 1, 1, 2))

for(i in 1:3){
  abline(v = i, col = i, lty = 2, lwd = 1.5)
  text(i + 0.2, 0.35,
       label = i, col = i)
}
```

## Comparison of Some Distributions

Now we will see three different distributions, normal, Poisson and binomial distribution, with the same mean.
Notice the similarity of the shape of distribution

```{r eval=F}
b <- 10:90

## all distributions with the same mean at 50
plot(b, dpois(b, 50), type = 's', ylim = c(0, 0.08),
     lwd = 1.5, main = "Comparison of distributions")
lines(b + 0.5, dnorm(b, 50, sqrt(50)), col = 2,
      lwd = 1.5)
lines(b, dbinom(b, 100, 0.5), col = 3, type ='s',
      lwd = 1.5)

legend("topright",legend=c("Poisson", "Normal", "Binomial"),
       col = 1:3, lty = 1, lwd = 1.5)

plot(b, dbinom(b, 100, 0.5),type = 's',
     lwd = 1.5)
lines(b + 0.5,dnorm(b, 50, sqrt(25)), col = 2,
      lwd = 1.5)
lines(b, dpois(b, 50),col = 3,type = 's',
      lwd = 1.5)

legend("topright",legend = c("Poisson", "Normal", "Binomial"),
       col = 1:3, lty = 1, lwd = 1.5)
```

## Exercise

**Submission:** A PDF file containing your answers.
If you are asked to work in R, your answers must contain **R codes** that you wrote and **the outputs** you get.
Your codes and outputs can be text and/or screen capture.

**Full score:** 10

**Data for this exercise:** None

### Questions/Directions

**1.** Give 1 example of a data with the log-normal distribution and give a reference to your answer

**2.** For the binomial distribution plot, explain in words what happens to the shape of the distribution when $np$ (mean) gets large?

**3.** For the Poisson distribution plot, explain in words what happens to the shape of the distribution when a lambda gets large?

**4.** Would the normal distribution (continuous) be a good approximation to the Poisson distribution (discrete)?
Why? Give a reference to your answer.

**5.** Would the normal distribution (continuous) be a good approximation to the binomial distribution (discrete)?
Why? Give a reference to your answer.

### Grading Rubric

| Question | Score 2                                       | Score 1                                                                            | Score 0                               |
|-------------------|------------------|------------------|------------------|
| 1        | Correct answer and give a reference           | One of the answers is not included or incorrect                                    | Answers are not included or incorrect |
| 2        | Correctly describe the shape in full sentence | Correctly describe the shape in but not in full sentence or partial correct answer | Incorrect answer or no answer         |
| 3        | Correctly describe the shape in full sentence | Correctly describe the shape in but not in full sentence or partial correct answer | Incorrect answer or no answer         |
| 4        | Correct answer and give a reference           | One of the answers is not included or incorrect                                    | Answers are not included or incorrect |
| 5        | Correct answer and give a reference           | One of the answers is not included or incorrect                                    | Answers are not included or incorrect |

**End of Lab 5**

---
