# Statistical modeling in R {#models}

**Duration:** 3-hour lab

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##  Learning Outcome: 

Students should be able to

1. demonstrate steps to run models in R

## Data Files for This Lab

- forest.csv [Click here to download](https://o365cmu-my.sharepoint.com/:x:/g/personal/pimonrat_t_cmu_ac_th/EU5zaH4DmWtImUA94AxVqMkB-DIDKI41VqegoYQCqZtaFQ?e=mLezqJ)

- test.csv [Click here to download](https://o365cmu-my.sharepoint.com/:x:/g/personal/pimonrat_t_cmu_ac_th/EScjHlLXQaZAlP8znz1SKu8BWgtn6rygn7Ymyrpx0czuqg?e=CBpuDn)

- train.csv [Click here to download](https://o365cmu-my.sharepoint.com/:x:/g/personal/pimonrat_t_cmu_ac_th/EeoxfjMlWYRIrkCT2vJ5wUgBkGxsR3sZOAjLebeNlOTgWg?e=G4XXCx)

## Introduction

We will use a data set from a research study by @khamyong2018 on Doi Suthep's forest structure for our lab session. We will see the relationship between tree height and diameter.

## Download Additional Packages

We will use `ggplot2` [@ggplot2] and `MLmetrics` [@mlmetrics].

```{r echo = T, eval = FALSE}
if(!require(MLmetrics)){install.packages("MLmetrics")}
if(!require(ggplot2)){install.packages("ggplot2")}

library(MLmetrics)
library(ggplot2)
```

## The Data Set

We import the data set of tree height and diameter at breast height (dbh).
For each tree, we also have the data of the habitat of the tree (forest types and plot). The data contain 6,890 rows of data. There are three forest types 

- dry deciduous, 
- dry evergreen, and 
- mixed deciduous forest. 

Then we will divide the whole data set into a training and a testing data sets. The training data are 70% sampled from the whole data set, while the other 30% of the data is for validation.

```{r echo = T, eval = FALSE}
### Import data
tack.new <- read.csv('data/forest.csv', header = T)

### assign right types of data to the variables
tack.new$community <- as.factor(tack.new$community)
tack.new$species <- as.factor(tack.new$species)
tack.new$family <- as.factor(tack.new$species)

###########################################
### separate into train and test data sets
set.seed(1234)
samp.new <- sample(nrow(tack.new), 0.7 * nrow(tack.new))
#tack.train.new <-tack.new[samp.new, ]
#tack.test.new <- tack.new[-samp.new, ]
```

This time I already provide train and test data set for you already.
So, you will run this following code.

```{r echo = T, eval = FALSE}
tack.train.new <- read.csv("data/train.csv", header = T)
tack.test.new <- read.csv("data/test.csv", header = T)

tack.train.new$community <- as.factor(tack.train.new$community)
tack.test.new$community <- as.factor(tack.test.new$community)

###################################################
### Sort test dataset by dbh 
### so you can use this to draw predicted line
###################################################
tack.test.new <- tack.test.new[order(tack.test.new$dbh), ]

```

## Data Visualization

Now, let's visualize the data We will do this in two ways - using R graphic and use ggplot2 package

```{r echo = T, eval = FALSE}
## use base r graphics
plot(tack.train.new$dbh, tack.train.new$height,
     xlab = "Diameter at breast height (cm)",
     ylab = "Height (m)",
     pch = as.numeric(tack.train.new$community),
     col = as.numeric(tack.train.new$community))
legend('bottomright',
       legend = c(unique(tack.train.new$community)),
       pch = as.numeric(c(unique(tack.train.new$community))),
       col = as.numeric(c(unique(tack.train.new$community)))
)
```

Take a look at the plot and compare the look of the plot with the one drawn from the package ggplot2.

```{r echo = T, eval = F, include = FALSE}
## use ggplot2 package
library(ggplot2)

plot <- ggplot(tack.train.new, 
               aes(x = dbh, y = height,
                   color = community,
                   shape = community))
plot + 
  geom_point() +
  scale_shape_manual(values = c(3, 20, 6)) +
  geom_smooth(se = TRUE, fullrange = FALSE,
              aes(fill = community))
```

## Modeling Steps

Now, let's run four competing non-linear models. We will evaluate how well each model fit the train data and also validate the models using the test data. We use the `set.seed()` function to guarantee that the same random values are produced each time you run the code. We start from a log-log model.

```{r echo = T, eval = FALSE}
##################################################
### Run competing models
##################################################
set.seed(1234)

#### 1 log-log used nls
loglog <- nls(height ~ exp(a + b* log(dbh)), 
              data = tack.train.new,
              start = list(a = 0.5, b = 0.1))
summary(loglog)
qqnorm(resid(loglog))
abline(a = 0, b = 1)
```

The summary of the model shows the significant parameter estimates of $a = 0.62295$ and $b = 0.663086$. The qq plot shows the deviation from normal distribution of the model residuals.

Now, let's fit other three models and see model summary and qq plots.

```{r echo = T, eval = FALSE}
###### 2 hyperbolic tangent
hyt <- nls((height) ~ a *(1 - exp(-(dbh) * b))/(1 + exp(-(dbh) * b)),
           data = tack.train.new,
           start = list(a = 3.8, b = 0.4))
summary(hyt)
qqnorm(resid(hyt))
abline(a = 0, b = 1)

##### 3 weibull
weibull <- nls((height) ~ a * (1 - exp(-b*(dbh^c))),
               data = tack.train.new,
               start = list(a = 3, b = 0.1, c = 1))
summary(weibull)
qqnorm(resid(weibull))
abline(a = 0, b = 1)

#### 4 monomolecular
mono <- nls((height) ~ a * (1 - exp(b-c*dbh)),
            data = tack.train.new,
            start = list(a = 30, b = 0.1, c = 0.1))

summary(mono)
qqnorm(resid(mono))
abline(a = 0, b = 1)
```

## Model Predictions

We have all the models.
Then we use the model parameter estimates to predict the height based on dbh for each model.

```{r echo = T, eval = FALSE}
################################################
### prediction of models
################################################
### prediction of train data
### We will use this to evaluate the models
tack.train.new$pred.loglog <- (predict(loglog, list(dbh =tack.train.new$dbh)))
tack.train.new$pred.hyt <- (predict(hyt, list(dbh =tack.train.new$dbh)))
tack.train.new$pred.weibull <- (predict(weibull, list(dbh =tack.train.new$dbh)))
tack.train.new$pred.mono <- (predict(mono, list(dbh =tack.train.new$dbh)))
head(tack.train.new)

### prediction of test data
### We will use this for validate the models
head(tack.test.new)
tack.test.new$pred.loglog <- (predict(loglog, list(dbh =tack.test.new$dbh)))
tack.test.new$pred.hyt <- (predict(hyt, list(dbh =tack.test.new$dbh)))
tack.test.new$pred.weibull <- (predict(weibull, list(dbh =tack.test.new$dbh)))
tack.test.new$pred.mono <- (predict(mono, list(dbh =tack.test.new$dbh)))
```

## Model Evaluation

For model evaluation, we will calculate several number and compare them among models.
The best model will be the one with

-   the *lowest* AIC
-   the *lowest* normalized RMSE
-   the *lowest* MSE.

```{r echo = T, eval = FALSE}
############################################
## Model evaluation
## MSE - mean square errors
## RMSE - normalized square root of MSE
## and AIC - Akaike's information criteria

AIC(loglog, hyt, weibull, mono)

library(MLmetrics)

MSE.train.all <- c(MSE(tack.train.new$pred.loglog, tack.train.new$height),
                   MSE(tack.train.new$pred.hyt, tack.train.new$height),
                   MSE(tack.train.new$pred.weibull, tack.train.new$height),
                   MSE(tack.train.new$pred.mono, tack.train.new$height))
MSE.train.all

## Normalized RMSE of train data
sqrt(MSE.train.all)/mean(tack.train.new$height)
```

## Model Validation

Now let's use the test data set to validate the prediction of the models.
We also calculate MAPE which stands for mean absolute percentage error showing how much the prediction deviates from the observed height measurements.
So, we want the model with *low* MAPE.

```{r echo = T, eval = FALSE}
############################################
### Model validation
### MSE RMSE MAPE of test data

MSE.test.all <- c(MSE(tack.test.new$pred.loglog, tack.test.new$height),
                  MSE(tack.test.new$pred.hyt, tack.test.new$height),
                  MSE(tack.test.new$pred.weibull, tack.test.new$height),
                  MSE(tack.test.new$pred.mono, tack.test.new$height))

MSE.test.all
## Normalized RMSE of test data
sqrt(MSE.test.all)/mean(tack.test.new$height)

##### MAPE test data - mean absolute percent errors 
mape.test.all <- c(MAPE(tack.test.new$pred.loglog, tack.test.new$height),
                   MAPE(tack.test.new$pred.hyt, tack.test.new$height),
                   MAPE(tack.test.new$pred.weibull, tack.test.new$height),
                   MAPE(tack.test.new$pred.mono, tack.test.new$height))

mape.test.all * 100
```

We got the best fit model which is the monomolecular model with three parameters. Now, we use the model to see relationship separated by forest types. I have run the monomolecular models separately by the forest types and gotten the parameter estimates a, b, c. I will not show the code how to do this.

-   DDF: height = 23.9370 \* (1 - exp(-0.0516 - 0.0318 \* dbh))
-   DEF:height = 40.5867 \* (1 - exp(-0.0426 - 0.0194 \* dbh))
-   MDF: height = 21.6263 \* (1 - exp(-0.0038 - 0.0712 \* dbh))

Finally, let's plot a publishable graph.

```{r, echo = T, eval = F}
library(dplyr)
### I added a new column with predicted height based on dbh for each tree (depending on the forest types)
tack.test.new <- 
  tack.test.new %>%
  mutate(
    mono = case_when(
      community == "DDF" ~ 23.9370 * (1 - exp(-0.0516 - 0.0318 * dbh)),
      community == "DEF" ~ 40.5867 * (1 - exp(-0.0426 - 0.0194 * dbh)),
      community == "MDF" ~ 21.6263 * (1 - exp(-0.0038 - 0.0712 * dbh))
    )
  )
```

## Data Visualization for Reports

Now use this code to plot.

```{r echo = T, , eval = F, warning = F}
library(ggplot2)
plot + 
  geom_point() +
  geom_line(data = tack.test.new, 
            aes(y = mono)) +
  xlab("Diameter at best height (cm)") +
  ylab("Tree height (m)") +
  theme_bw() +
  theme(axis.line = element_line(colour = "black"),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank(),
              panel.border = element_blank(),
              panel.background = element_blank(),
    legend.position = c(0.8, 0.2)) +
  scale_colour_hue(name = "Forest type",
                   breaks = c("DDF", "DEF", "MDF"),
                   labels = c("Dry dipterocarp",
                              "Dry evergreen",
                              "Mixed deciduous")) +
  scale_shape_manual(name = "Forest type",
                     breaks = c("DDF", "DEF", "MDF"),
                     values = c(3, 20, 6),
                     labels = c("Dry dipterocarp",
                                "Dry evergreen",
                                "Mixed deciduous"))
```

## Exercise

**Submission:** A PDF file containing your answers.
Your answers must contain **R codes** that you wrote and **the outputs** you get.
Your codes and outputs can be text and/or screen capture.

**Full score:** 10 Submission:

**Data file for this exercise:** forest.csv

### Questions/directions

**1.** In R, run one more model using the train data set.
The model equation is ln(H) = a \* DBH\^b.
Then, show the summary of the model.

```{r echo = T, eval = FALSE}
power.log <- nls(log(height) ~ a * dbh^b, data = , start = list(a = 0.5, b = 0.1))
```

**2.** See the AIC of the model

**3.** In words, compare the AIC of the power model and the monomoloecular model above.
Which model is better?

**4.** Validate the model by create additional column of predicted height based on the power model.

**5.** Calculate the MAPE of the power model and compare with the MAPE of the monomolecular model.
Which model is preferable?

### Grading rubric

| Question | Score = 2 | Score = 1 | Score = 0 |
|-------------------|------------------|------------------|------------------|
| 1 | Run the code and show the summary | Run the code but do not show the summary | R code and output are not included or incorrect |
| 2 | Correct R code and show the AIC | Partially correct | R code and output are not included or incorrect |
| 3 | Correctly answer | Partially correct answer | No answer |
| 4 | Run the code and show the column | Run the code but do not show the column | R code and output are not included or incorrect |
| 5 | Correct R code and show the MAPE | Partially correct | R code and output are not included or incorrect |

**End of Lab 9**