# Regression

**Duration:** 4-hour lecture

## Learning outcomes

Students should be able to:

- Give some examples of types of regression analyses
- Determine the assumptions of linear regression
- Identify if data are suitable for regression analysis
- Determine the steps for doing regression analysis
- Determine link functions to be used for different data types
- Identify the problems with fitting generalized linear models and adjust the models

## Introduction

Regression analysis is a fundamental statistical technique used to explore and model the relationship between variables. The primary goal is to determine how one or more explanatory (independent) variables influence a response (dependent) variable. Beyond identifying relationships, regression models are widely applied for making predictions.

For regression analysis, both response and explanatory variables are *continuous*. It allows researchers to:

1. Quantify the strength and form of the relationship between variables.
2. Estimate parameters that describe this relationship.
3. Use the model for prediction.

We will explore different types of regression models, with emphasis on linear and nonlinear regression. At its core, a regression model expresses the response variable, $Y$, as a function of one or more explanatory variables $X$, plus an error term. Here is an example of equation for multiple linear regression when there is more than one explanatory variable.

$$Y = a + b_1 X_1 + b_2 X_2 + \cdots + b_k X_k + \varepsilon$$

Where:

- $\varepsilon$ = random error that cannot be explained by the model
- $a$, $b$ = parameter estimate

## Different types of regression

There is not a single fixed number of regression types, but statisticians usually group them into several major categories depending on data structure, assumptions, the form of the relationship, and the research question.

### 1. Linear regression

Linear regression is the simplest and most common approach, where the relationship between explanatory and response variables is assumed to follow a straight line. Some subtypes of linear regression are:

- **Simple linear regression** – there is one explanatory variable
- **Multiple linear regression** – there are two or more explanatory variables
- **Polynomial** - this is considered a linear model, even though the curve it produces is not a straight line. For example, below the relationship between $X$ and $Y$ is not linear in shape but each parameter enters the equation in a linear way.

$$Y = a + b_1 X + b_2 X^2 + \varepsilon$$

- **Piecewise regression** – it fits separate linear models to different ranges of data, allowing researchers to capture relationships that change at certain thresholds or breakpoints.
- **Robust regression** - This is useful when a dataset contains outliers or influential points. Robust regression reduces the impact of extreme values, producing more reliable estimates in the presence of noisy or imperfect data.

### 2. Nonlinear regression

Nonlinear regression is used when the relationship between explanatory and response variables cannot be adequately described by a straight line. Instead of assuming linearity, the model specifies a nonlinear function, such as exponential, logarithmic, or logistic forms.

### 3. Generalized linear regression or generalized linear models (GLMs)

This extends the framework of linear regression to situations where the response variable does not follow a normal distribution. GLMs allow for different types of response distributions (e.g., binomial for binary outcomes, Poisson for count data).

### 4. Other approaches such as nonparametric regression

The nonparametric regression does not assume a specific functional form. It is flexible but less interpretable.

In this course, the focus will be on simple linear regression, nonlinear regression and GLMs.

## Linear regression

The simplest form of regression is linear regression, where the relationship between $Y$ and $X$ is assumed to be linear. For a simple linear regression (a single explanatory variable), the model is:

$$Y = a + bX + \varepsilon$$

Where:

- $Y$ is the response variable
- $X$ is the explanatory variable
- $a$ is the intercept (value of $Y$ when $X = 0$)
- $b$ is the slope (change in $Y$ for a one-unit change in $X$)
- $\varepsilon$ is the random error term

## Assumptions of linear regression

For the linear regression model to be valid, several assumptions must be met. Violations of these assumptions can lead to biased estimates or invalid conclusions.

1. **Linearity** - The relationship between explanatory variables and the response is linear.
2. **Independence** - Observations are independent of one another.
3. **Homoscedasticity** - Errors have constant variance ($\sigma^2$).
4. **Normality** - Errors are normally distributed with mean zero.
5. **No errors in X**
6. **No missing value**

### Example: examine the relationship between ground vs aerial based seedling-height survey

@changsalak2022 compared software tools to help with seedling measurement in a restoration project. They aimed to compare the seedling height from ground measurement with the seedling height from 3D models of aerial photographs. The $Y$ was seedling height of the models, and the $X$ was height from the ground measurement (ground truth data) (Figure \@ref(fig:fig-seedling-height)).

**Biological hypothesis**: The researchers expected that if the aerial photography is a good estimate of seedling height, the slope coefficient should be close to 1 and the $R^2$ should be close to 1.

```{r fig-seedling-height, echo=FALSE, fig.cap="Plotting seedling height from ground measurements against the height derived from the 3D model of aerial photography (Figure is modified from [@changsalak2022])", out.width="80%"}

# Load required library
library(ggplot2)

# Create sample data (approximating the plot)
# You should replace this with your actual data
ground_Height <- c(0.3, 0.5, 0.6, 0.7, 0.75, 0.8, 0.85, 1.0, 1.05, 1.1, 1.15, 1.2, 1.3, 1.6, 1.9)
Dronedeploy_Height <- c(0.28, 0.52, 0.72, 0.62, 0.8, 0.72, 0.65, 1.15, 0.9, 1.15, 1.28, 1.1, 1.05, 1.6, 1.6)

# Create a data frame
data <- data.frame(ground_Height, Dronedeploy_Height)

# Fit linear regression through the origin (intercept = 0)
model <- lm(Dronedeploy_Height ~ 0 + ground_Height, data = data)

# Get the slope
slope <- coef(model)[1]

# Calculate fitted values and residuals
data$fitted <- predict(model, data)
data$residuals <- data$Dronedeploy_Height - data$fitted

# Calculate fitted values and residuals
data$fitted <- predict(model, data)
data$residuals <- data$Dronedeploy_Height - data$fitted

# Get R-squared value
r_squared <- summary(model)$r.squared

# Create equation text
equation_text <- paste0("y = ", round(slope, 2), "x")
r2_text <- paste0("R² = ", round(r_squared, 3))

# Create the plot
ggplot(data, aes(x = ground_Height, y = Dronedeploy_Height)) +
  
  # Add points on top
  geom_point(color = "#3182bd", size = 4, alpha = 0.8) +
  
  labs(x = "\nGround truth height (m)",
       y = "Height derived from Dronedeploy (m)\n") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_line(color = "gray95"),
    panel.background = element_rect(fill = "gray98", color = NA),
    plot.background = element_rect(fill = "gray98", color = NA)
  ) +
  scale_x_continuous(limits = c(0, 2.1), breaks = seq(0, 2, 0.5)) +
  scale_y_continuous(limits = c(0, 2.1), breaks = seq(0, 2, 0.5)) +
  coord_fixed(ratio = 1)  # Equal aspect ratio

# Alternative: to save the plot
# ggsave("dronedeploy_regression.png", width = 8, height = 6, dpi = 300)
```

Linear regression attempts to fit a straight line through a cloud of data points. Many lines are generated through the data points. Each line has its own intercept and slope coefficient estimates. The best-fitting line (black line in Figure \@ref(fig:fig-residuals)) is the line that minimizes the sum of squared differences between observed values and predicted values (least squares method).

It is important to see that there are residuals (errors) between the actual data points and the fitted line (Figure \@ref(fig:fig-residuals)). The residual means the vertical distance between observed points and the fitted line.

```{r fig-residuals, echo=FALSE, fig.cap="Residuals of the regression line. Residuals provide a measure of model accuracy. If residuals are large or show patterns, the linear model may be inappropriate. (Figure is modified from [@changsalak2022])", out.width="80%"}
# Create the plot
ggplot(data, aes(x = ground_Height, y = Dronedeploy_Height)) +
  # Add red residual lines first
  geom_segment(aes(x = ground_Height, y = Dronedeploy_Height,
                   xend = ground_Height, yend = fitted),
               color = "red", linewidth = 0.5, alpha = 0.7) +
  # Add regression line
  geom_abline(intercept = 0, slope = slope, 
              color = "#2c3e50", linewidth = 0.8) +
  # Add points on top
  geom_point(color = "#3182bd", size = 4, alpha = 0.8) +
  # Add equation and R-squared text
  annotate("text", x = 0.3, y = 1.9, 
           label = equation_text, 
           hjust = 0, size = 4.5, fontface = "bold") +
  annotate("text", x = 0.3, y = 1.75, 
           label = r2_text, 
           hjust = 0, size = 4.5, fontface = "bold") +
  labs(x = "\nGround truth height (m)",
       y = "Height derived from Dronedeploy (m)\n") +
  theme_minimal() +
  theme(
    plot.title = element_text(size = 16, face = "bold", hjust = 0),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    panel.grid.major = element_line(color = "gray90"),
    panel.grid.minor = element_line(color = "gray95"),
    panel.background = element_rect(fill = "gray98", color = NA),
    plot.background = element_rect(fill = "gray98", color = NA)
  ) +
  scale_x_continuous(limits = c(0, 2.1), breaks = seq(0, 2, 0.5)) +
  scale_y_continuous(limits = c(0, 2.1), breaks = seq(0, 2, 0.5)) +
  coord_fixed(ratio = 1)  # Equal aspect ratio

# Print regression statistics
cat("Regression through the origin:\n")
cat("Slope:", round(slope, 3), "\n")
cat("R-squared:", round(summary(model)$r.squared, 3), "\n")

```

By plotting ground height against height derived from one of the models, we observe a positive linear trend. After using linear regression, the linear equation ($n = 15$) is $Y = 0.94X$ and the $R^2 = 0.98$. It seems like the height from 3D model can be used to estimate the height of seedlings well.

## Nonlinear regression

ADD content here: thinking about my data - Tree Khamyong paper?

## Generalized linear regression or generalized linear models (GLMs)

ADD content here: Aom data??



## Exercise: Lesson 10 Regression

1. Give three names of types of regression analyses

2. What are the assumptions of linear regression?

3. Incomplete - Is this data suitable for regression analysis?

| Data                                | Answer |
|-------------------------------------|--------|
| (to be added)                       |        |
| ?                                   |        |
| All continuous                      |        |
| Mixed continuous and categorical    |        |
| Mixed continuous and categorical    |        |

4. Determine the steps for doing regression analysis

5. Determine link functions to be used for different data types

| Response variables | Link function |
|--------------------|---------------|
| Proportion         |               |
| Continuous         |               |
| Binary             |               |
| Count              |               |

6. What is overdispersion? Why is it a problem when fitting generalized linear models? What to do if you encounter overdispersion?

---

## References
