# Regression analyses in R {#regression}

**Duration:** 3-hour lab

## Learning Outcome

Students should be able to

1.  demonstrate steps to run regression models in R

## Data Files for This Lab

-   seed3spp.csv [Click here to download](https://o365cmu-my.sharepoint.com/:x:/g/personal/pimonrat_t_cmu_ac_th/Eb8jMoWnwP9BvJnAaEIfWYkBu-kTlKaWh6tD-Icn579-wA?e=3Te2Up)

## Introduction

We will use a data set from a research study by @yiamthaisong2024 on Doi Suthep's seed morphology for our lab session. We will explore the relationships between seed traits.

## Download Additional Packages

We will use `ggplot2` [@ggplot2], `GGally` [@ggally], and `gvlma` [@gvlma].

We will also download additional packages to use in this lab session.

```{r package installing, echo = T, eval = F}
if(!require(GGally)){install.packages("GGally")}
if(!require(ggplot2)){install.packages("ggplot2")}
if(!require(gvlma)){install.packages("gvlma")}

library(ggplot2)
library(GGally)
library(gvlma)
```

## The Data Set

We import the data set of seed trait data of three tree species from Doi Suthep-Pui national park.
For each species, we collected trait data from 20 seeds.
The data contain 60 rows and 11 columns.
Let's look at the data set and see what trait data we have.

```{r import, echo = T, eval = T, results ='hide'}
### Import data
seed <- read.csv("data/seed3spp.csv", header = T)

### assign right types of data to the variables
seed$genus <- as.factor(seed$genus)
seed$epithet <- as.factor(seed$epithet)

### see the data
seed

summary(seed[,7:11])
```

## Data Visualization

Now, let's visualize the data and see the correlations among the variables.
Remember this first step is EDA and we should **always** do EDA before we run further analyses.

We will plot some graphs using `ggplot2` and `GGally` packages.

```{r explore, echo = T, eval = F, results='hide'}
## use GGally package
library(ggplot2)
library(GGally)

## use data from column 7:11 - volume to moisture
ggpairs(seed,          
        columns = 4:11) # selected columns
```

We can also plot the graph separately by species to see the differences among species.

```{r explore2, echo = T, eval = F, results='hide'}
ggpairs(seed,                 # Data frame
        columns = 7:11,        # Columns
        aes(color = genus,  # Color by species 
            alpha = 0.5))     # Transparency
```

Let's plot the two variables that we are interested in.

```{r plot, echo = T, eval = F, results='hide'}
library(ggplot2)
plot <- ggplot(seed, 
               aes(x = volume_mm3,
                   y = dry_weight_g * 1000,
                   color = genus,
                   shape = genus))

plot +
  geom_point() +
  xlab("\nSeed volume (cubic millimeter)") +
  ylab("Seed dry mass (milligram)\n")
```

## Linear Regression Model

A linear regression model aims to model a dependent variable as a function of one or more independent variable(s) [@regref]. Here, let's run linear regression models between volume (cubic mm) (dependent variable) and dry weight (mg) (independent variable).
We will evaluate how well each model fit and get the linear equation.

$$Y = β₀ + β₁X + ε $$

-   $Y$ is the seed dry weight
-   $X$ is the seed volume
-   $β₀$ is an intercept and
-   $β₁$ is a slope
-   $ε$ is an error term

We use the `set.seed()` function to guarantee that the same random values are produced each time you run the code.

```{r reg.model1, echo = T, results = 'hide'}
##################################################
### Run a regression model
### and run a function to get
### overall p value
##################################################
set.seed(1234)
model <- lm(dry_weight_g * 1000 ~ volume_mm3,
            data = seed)

summary(model)

overall_p <- function(my_model) {
    f <- summary(my_model)$fstatistic
    p <- pf(f[1],f[2],f[3],lower.tail=F)
    attributes(p) <- NULL
    return(p)
}
```

## Summary of Models

The summary of the model shows the significant parameter estimates of intercept = `r summary(model)$coefficients["(Intercept)", "Estimate"]` and slope = `r summary(model)$coefficients["volume_mm3", "Estimate"]`.
This means that the linear equation is

Dry weight * 1000 g = `r round(summary(model)$coefficients["(Intercept)", "Estimate"], digit = 2)` + `r round(summary(model)$coefficients["volume_mm3", "Estimate"], digit = 2)` \* (`r model$terms[[3]]`) + ε

Look at the model summary.
You will see $R^2^$ = `r round(summary(model)$r.squared, digit = 3)`.
This means that about `r round(summary(model)$r.squared, digit = 3) * 100` of variance in the dependent variable is predictable from the independent variable (`r model$terms[[3]]`).
This means the linear model fits the data well.

In addition, the F-statistics indicates whether the model with (dry weight * 1000 g) fit the data better than null model (model without `r model$terms[[3]]`).
Here we got F = `r round(summary(model)$fstatistic[["value"]], digit = 2)` and p value = `r overall_p(model)`, meaning the model with `r model$terms[[3]]` is significant.

## Model Evaluation

Now, we will evaluate the model to see if the model met the assumptions.

First, let's check if the residuals are normally distributed with mean of 0.

```{r model.eval1, echo = F, eval = T, results = 'hide', fig.keep = 1}

hist(model$residuals)
abline(v = mean(model$residuals))
qqnorm(model$residuals)
qqline(model$residuals, lwd = 2)
shapiro.test(model$residuals)
```

The plot shows the mean of residuals is about 0 but the distribution of the residuals is not quite normal (see the result of Shapiro test; p value = `r round(shapiro.test(model$residuals)$p.value, digit = 3)`).

Second, let's see a set of four diagnostic plots.
We can check for heteroscedasticity using a fitted value vs residual plot.
If you notice a speaker phone shape of points, it means heteroscedasticity.

```{r eval.plot, echo = F, eval = T}
par(mfrow=c(2,2)) 
plot(model)
```

See the upper left plot.
We seems to have a heteroscedasticity issue here.

Also, see the Q-Q residuals plot. The residuals are not quite normally distributed.
The bottom right plot, Residuals vs Leverage shows the regression is influenced by each data point.

From the plots, some data points (with written number) influence the regression and affect the quality of the regression.

In addition, there is a package that can help you check the assumptions of linear regression. Let's use `gvlma()` function from `gvlma` package.

```{r gvlma, echo = F, eval = F, results = 'asis'}
library(gvlma)
gvlma::gvlma(model)
```

The results of `gvlma` means

-   Global Stat: overall metric indicates whether the model passes or fails.
    Here, the model failed.

-   Skewness: measuring the distribution.
    Here, we passed.

-   Kurtosis: measuring the distribution, outliers, influential data.
    Here, we passed.

-   Link function: whether we mis-specified model.
    Here, we passed.

-   Heteroscedasticity: looking for equal variance in the residuals.
    Here, the model failed.
    The fitted vs residual plot agreed with this.

## Further Investigation

Look at our data closely.

-   What factor do you think might effect the relationship of volume vs dry weight?
-   Do you think different species will have the same volume and dry weight relationship?

Let's take into account the species in the linear model.
The regression model with more than one independent variable is call a multiple regression model.

```{r reg.model2, echo = T, results = 'hide'}
##################################################
### Run a multiple regression model 
### (x = volume and species)
##################################################
set.seed(1234)
model2 <- lm(dry_weight_g * 1000 ~ volume_mm3 * genus,
            data = seed) # with a interaction effect

summary(model2)

final.model <- step(model2) # step for model selection

AIC(model, final.model) # compare AIC of the model

anova(final.model)

```

Look at the output of `anova(final.model)`.
You will see that the two factors are significant in the model.

```{r summary, echo = F, eval = F, results = 'asis'}
summary(final.model)
```

## Exercise

**Submission:** A PDF file containing your answers.
If you are asked to use R, your answers must contain **R codes** that you wrote and **the outputs** you get. Your codes and outputs can be text and/or screen capture.

**Full score:** 10

**Data for this exercise:** seed3spp.csv (you already downloaded them.)

### Questions/Directions

**1.** Select two other variables that you would like to run a linear regression.
In R, plot a scatter plot of the two variables.

**2.** In R, run a linear model of the two selected variables, and show the summary of the model.

**3.** In R, evaluate the model and show the output of the evaluation.

**4.** Describe in words, how well the model fit the data.

**5.** Write out the model equation and describe in words what the model means,

### Grading Rubric

| Question | Score = 2                                                    | Score = 1                                                          | Score = 0                                                                               |
|-------------------|------------------|------------------|------------------|
| 1        | Correct R code and a plot with proper axis labels            | One of R code or output is not included or incorrect               | R code and output are not included or incorrect                                         |
| 2        | Correctly run the model and include the summary of the model | One of R code or output is not included or incorrect               | R code and output are not included or incorrect                                         |
| 3        | Correct R code and the outputs of the model evaluation       | One of R code or output is not included or incorrect               | R code and output are not included or incorrect                                         |
| 4        | Correctly describe the model evaluation                      | Partially correct description of the model evaluation              | Incorrect description of the model evaluation                                           |
| 5        | Correctly provide the equation and describe the meaning      | Partially correctly describe the model equation and/or the meaning | Incorrectly describe the model equation and/or the meaning or do not include the answer |

**Eng of Lab 10**