---
output:
  html_document: default
  pdf_document: default
---
# Analysis of Variance and Analysis of Covariance

**Duration:** 2-hour lecture

## Learning outcomes

Students should be able to:

-   Describe common terminology relating to ANOVA
-   Determine steps for performing ANOVA and ANCOVA
-   Identify when to use ANCOVA versus ANOVA
-   Explain the purpose of covariates in experimental design
-   Interpret ANOVA and ANCOVA results and post-hoc tests

## Introduction

Analysis of Variance (ANOVA) is a fundamental statistical technique used to test differences between two or more group means. ANOVA extends the t-test to more than two groups. The test determines whether observed differences among group means are statistically significant by partitioning total variability into meaningful sources of variation.

Analysis of Covariance (ANCOVA) is also a statistical techniques that combines ANOVA and regression to evaluate the influence of categorical and continuous predictors on a response variable. ANCOVA incorporates one or more covariates, allowing adjustment for potential confounding variables and increasing the precision of treatment comparisons.

### ANOVA vs. ANCOVA vs. Regression

There are some key different features among the tests (Table \@ref(tab:differences)). A covariate is a continuous (sometimes categorical) variable that is not the main factor of interest in a research study. However, the covariate is included in the analysis because it may influence the dependent variable. By accounting for a covariate, researchers can reduce error variation and obtain a clearer estimate of the treatment of interest.

| Feature               | ANOVA               | ANCOVA                   | Regression                        |
|-----------------|-----------------|-----------------|--------------------|
| Explanatory variables | Categorical only    | Categorical + Continuous | Any combination                   |
| Purpose               | Compare group means | Compare adjusted means   | Predict and explain relationships |
| Covariates            | None                | Yes (continuous)         | Can be included                   |

: (\#tab:differences) Key differences among ANOVA, ANCOVA, and regression

For example, suppose researchers are comparing tree growth rates among different forest restoration treatments (e.g., natural regeneration vs. active planting). In this case, soil moisture or initial seedling height could be included as a covariate. These variables are not the main treatment, but they strongly affect growth.

## What is ANOVA?

It is conceptually similar to regression analysis, with one key difference that the explanatory variable is **categorical** (e.g., treatment groups, fertilizer types) (Table \@ref(tab:differences)). ANOVA tells you whether means differ, but **not which specific means** are different from one another. To identify specific differenceswhich treatments are different post-hoc tests (e.g. Tukey's HSD) are needed.

## The Basic Concept of ANOVA

ANOVA works by partitioning the total variation in the data into:

1.  **Variation among groups** (due to treatment effects)
2.  **Variation within groups** (random error)

If the variation among groups is **substantially larger** than the variation within groups, we conclude that the groups have different means.

## One-Way ANOVA

### Hypotheses

For one-way ANOVA with k groups:

**Null Hypothesis (H₀):** μ₁ = μ₂ = μ₃ = ... = μₖ

-   All population means are equal.
-   Interpretation: There is no treatment effect.

**Alternative Hypothesis (H₁):** Not all means are equal.

-   At least one population mean is different from the others. This does NOT mean all means are different; some may be equal
-   Interpretation:There is a treatment effect.

## Example: Plant Growth with Two Fertilizers

Researcher want to compare plant height under three fertilizer treatments (Fertilizer A and B and Control - no fertilizer). The researhcers starts with seedlings with the same sizes (Table \@ref(tab:example-data)).

**Research design**

-   **Explanatory variable**: Fertilizer type (A, B, and no fertilizer)
-   **Response variable**: Plant height (cm)
-   **Sample size**: 7 plants per treatment (21 plants total)

```{r example-data, echo=FALSE, results = 'hide', message=FALSE, warning=FALSE}
# Create example dataset
set.seed(213708)
fertilizer <- rep(c("A", "B", "No fertilizer"), each = 7)
height <- c(
  # Treatment A (lower heights)
  6, 3, 4, 5, 3, 6, 4,
  # Treatment B (higher heights)
  8, 5, 6, 7, 8, 6, 8,
  # control (lower heights)
  3, 2, 4, 5, 3, 3, 2
)

plant_data <- data.frame(
  Plant_ID = 1:21,
  Fertilizer = factor(fertilizer),
  Height = height
)

# Display the data
#print(plant_data)

# Summary statistics
library(dplyr)
summary_plant <- plant_data %>%
  group_by(Fertilizer) %>%
  summarise(
    n = n(),
    Mean = round(mean(Height), digit = 1),
    SD = round(sd(Height), digit = 1),
    SE = round(SD / sqrt(n), digit = 1),
    Min = min(Height),
    Max = max(Height),
    .groups = 'drop'
  )

```


```{r plant-data-tab, echo=FALSE, message=FALSE, warning=FALSE, tab.cap="The data of fertilizer experiment for ANOVA", results='asis'}
### Write the data into a table
library(knitr)
library(kableExtra)

plant_data %>%
  kbl(align = "c",
      col.names = c("Plant ID", "Fertilizer", "Height (cm)"),
      caption = "Plant height under different fertilizer treatments") %>%
  kable_styling(full_width = F) %>%
  column_spec(1, bold = TRUE)
```

### Visualizing the Data

It is always a good practice to explore the data before testing. The research should make the summary of the data separated by the factor of interest (Fertilizer) in a table (Table \@ref(tab:summary-data)) and/or a boxplot (Figure \@ref(fig:visualize-groups)).

```{r summary-data, echo=FALSE, message=FALSE, warning=FALSE, tab.cap="Summary of experiment data for ANOVA", results='asis'}
summary_plant%>%
  kbl(align = "c",
      col.names = c("Fertilizer", "n", "Mean height (cm)", 
                    "SD (cm)", "SE (cm)",
                    "Mininum height (cm)",
                    "Maximum height (cm)"),
      caption = "Summary of plant height separated by fertilizer treatments") %>% 
  kable_styling(full_width = F) %>%
  column_spec(1, bold = TRUE)
```


```{r visualize-groups, fig.height=5, message=FALSE, warning=FALSE, echo=FALSE, fig.cap="Boxplot shows plant height under different treatments"}

if(!require("ggthemes", character.only = TRUE)) {
  install.packages("ggthemes", dependencies = TRUE)}

library(ggplot2)
library(ggthemes)

# Boxplot
ggplot(plant_data, aes(x = Fertilizer, y = Height, fill = Fertilizer)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.1, size = 2) +
  geom_hline(yintercept = mean(plant_data$Height), 
             linetype = "dashed", color = "purple", size = 1) +
  annotate("text", x = 1.5, y = mean(plant_data$Height) + 0.3, 
           label = "Overall Mean", color = "purple") +
  labs(
       x = "Fertilizer Treatment",
       y = "Plant height (cm)") +
  theme_bw() +
  theme(panel.border = element_blank(), legend.position = "none",
  axis.line.x = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
  axis.line.y = element_line(linewidth = 0.5, linetype = "solid", colour = "black"),
  axis.text = element_text(size = 14), axis.title = element_text(size = 14))

```

### Understanding Sums of Squares

ANOVA partitions the total variation into components:

#### Total Sum of Squares (SSTotal)

SSTotal measures total variation of all observations from the overall mean:

$$SS{Total} = \sum_{i=1}^{n}(y_i - \bar{\bar{y}})^2$$

where:

- $y_i$ = value of each data point
- $\bar{\bar{y}}$ = overall mean (grand mean)

#### Sum of Squared Error (SSE)

SSE is also called "within-group variation" or "residual variation":

$$SSE = \sum_{i=1}^{n}(y_i - \bar{y}_A)^2 + \sum_{i=1}^{n}(y_i - \bar{y}_B)^2$$

This represents variation of individuals within each treatment group.

#### Sum of Squares Between (SSB)

SSB is also called "among-group variation" or "treatment variation." It measures variation of group means from the overall mean, weighted by sample size.

#### The Partition

$$SS{Total} = SSB + SSE$$

```{r calculate-ss, message=FALSE, echo=FALSE, warning=FALSE, results='hide'}
# Calculate sums of squares manually
overall_mean <- mean(plant_data$Height)
group_means <- plant_data %>%
  group_by(Fertilizer) %>%
  summarise(group_mean = mean(Height))

# SSTotal
ss_total <- sum((plant_data$Height - overall_mean)^2)

# SSE (within groups)
sse <- plant_data %>%
  left_join(group_means, by = "Fertilizer") %>%
  mutate(sq_dev = (Height - group_mean)^2) %>%
  summarise(SSE = sum(sq_dev)) %>%
  pull(SSE)

# SSB (between groups)
ssb <- sum(7 * (group_means$group_mean - overall_mean)^2)

# Verify partition
cat("SSTotal =", round(ss_total, digits = 2), "\n")
cat("SSB =", round(ssb, digits = 2), "\n")
cat("SSE =", round(sse, digits = 2), "\n")
cat("SSB + SSE =", round(ssb + sse, digits = 2), "\n")
```

In this fertilizer experiment, 
- SSB is `r round(ssb, digits = 2)`.
- SSE is `r round(sse, digits = 2)`.
- SSTotal is `r round(ssb, digits = 2)` + `r round(sse, digits = 2)`, equal to `r round(ss_total, digits = 2)`.

### Degrees of Freedom

The concept of degrees of freedom (df) was introduced by Johann Carl Friedrich Gauss in 1821. It represents the **number of values in a calculation that are free to vary**.

For our fertilizer example:

- **df for SSB** = k - 1 = 3 - 1 = 2 (where k = number of treatments)
- **df for SSE** = k(n - 1) = 3(7 - 1) = 18 (where n = replicates per treatment)
- **df for SSTotal** = N - 1 = 21 - 1 = 20 (where N = total observations)

### The F-Distribution and F-Statistic

The F-distribution is used to test whether two variances are equal. The F-statistic is the ratio:

$$F = \frac{\text{Variance among groups}}{\text{Variance within groups}} = \frac{MSB}{MSE}$$

where:

- $MSB = \frac{SSB}{df_SSB}$ (Mean Square Between)
- $MSE = \frac{SSE}{df_SSE}$ (Mean Square Error)

**Interpretation:**

- If F is close to 1, sample variances are equal (no treatment effect)
- If F is large, variation among groups exceeds variation within groups (treatment effect likely)

### The ANOVA Table

The standard format for presenting ANOVA results is in Table \@ref(tab:anova-tab-temp).

| Source of Variation | Sum Square | df | Mean Square (MS) | F | P-value |
|---------------------|----|----|------------------|---|------------|
| Treatment (Between) | SSB | k-1 | MSB = SSB/(k-1) | F = MSB/MSE | From F-table |
| Error (Within) | SSE | k(n-1) | MSE = SSE/[k(n-1)] |  |  |
| Total | SSTotal | kn-1 |  |  |  |

: (\#tab:anova-tab-temp) ANOVA table

ANOVA is a hypothesis testing, so p-value is interpreted accordingly. In common practice, the significance level is set at $\alpha$ = 0.05.

- If the P-value $\geq$ 0.05, the null hypothesis is not rejected.
- If the P-value $<$ 0.05, the null hypothesis is rejected.

```{r run-anova, echo = FALSE, results='hide', message=FALSE, echo=FALSE, warning=FALSE}
if(!require("gridExtra", character.only = TRUE)) {
  install.packages("gridExtra", dependencies = TRUE)
  library("gridExtra", character.only = TRUE)
}

# Fit the model
anova_model <- aov(Height ~ Fertilizer, data = plant_data)

# Display ANOVA table
#summary(anova_model)

# More detailed output
anova_results <- anova(anova_model)
#print(anova_results)


if(!require("sfsmisc", character.only = TRUE)) {
  install.packages("sfsmisc", dependencies = TRUE)}
if(!require("formatdown", character.only = TRUE)) {
  install.packages("formatdown", dependencies = TRUE)}

library(sfsmisc)
library(formatdown)
```

Table \@ref(tab:anova-tab-fertilizer) shows the result of the fertilizer experiment. If the P-value $<$ 0.05, the null hypothesis is rejected. There are at least one mean of the plant growth differ from other means.

| Source of Variation | Sum Square | df | Mean Square (MS) | F | P-value |
|---------------------|----|----|------------------|---|------------|
| Treatment (Between) | `r round(anova_results$"Sum Sq"[1], digit = 1)` | `r round(anova_results$"Df"[1], digit = 1)` | `r round(anova_results$"Mean Sq"[1], digit = 1)` | `r round(anova_results$"F value"[1], digit = 1)` | `r format_sci(anova_results$"Pr(>F)"[1], digits = 2)` |
| Error (Within) | `r round(anova_results$"Sum Sq"[2], digit = 1)` | `r round(anova_results$"Df"[2], digit = 1)` | `r round(anova_results$"Mean Sq"[2], digit = 1)` |  |  |
| Total | `r round(anova_results$"Sum Sq"[1] + anova_results$"Sum Sq"[2] , digit = 1)` | `r anova_results$"Df"[1] + anova_results$"Df"[2]` |  |  |  |

: (\#tab:anova-tab-fertilizer) ANOVA table of the fertilizer experiment

Before reporting the results, researchers must check if the assumptions of ANOVA are met.

### Assumptions of ANOVA
It is important that before interpreting ANOVA results, the following assumptions must be verified [@field2012discovering]

#### Random Sampling
Data should be collected through random sampling.

#### Homogeneity of Variance (Homoscedasticity)
Variances should be equal across groups.

**Check BEFORE running ANOVA:**

There are several tests for homogeneity of variance such as Levene's test below. It is a hypothesis testing with a null hypothesis of equal variances. For visual check, boxplots should have similar spreads (see Figure (\@ref(fig:visualize-groups))).

```{r levene-test, , echo = FALSE, message=FALSE, warning=FALSE, results='hide'}
# Levene's test for homogeneity of variance
library(car)
levene <- leveneTest(Height ~ Fertilizer, data = plant_data)

# Visual check
#ggplot(plant_data, aes(x = Fertilizer, y = Height)) +
#  geom_boxplot() +
#  labs(title = "Check for Equal Variances",
#       subtitle = "Boxplots should have similar spreads") +
#  theme_minimal()
```

In our case, the result of the Levene's Test (P-value  = `r round(levene$"Pr(>F)"[1], digit = 2)`) indicates that the null hypothesis is not rejected. The variances among groups are homogeneous (Table \@ref(tab:levene-tab)).

|  | df | F | P-value |
|-----|----|---|------------|
| group |`r levene$Df`|`r round(levene$"F value"[1], digit = 2)`|`r format_sci(levene$"Pr(>F)"[1], digits = 2)`|

: (\#tab:levene-tab) `r attr(levene, "heading")`

#### Independence of Errors
Residuals should be independent of each other.

#### Normality of Residuals
Residuals should follow a normal distribution.

**Check AFTER running ANOVA:**

```{r shapiro-fertilizer, message=FALSE, warning=FALSE, echo = FALSE}
# Shapiro-Wilk test

# Extract residuals
residuals <- residuals(anova_model)

shapiro <- shapiro.test(residuals)
```

This assumption is checked using visualization (Figure \@ref(fig:check-normality)) and Shapiro-Wilk normality test .

```{r check-normality, message=FALSE, warning=FALSE, echo = FALSE, fig.cap="Q-Q plot (left) and histogram (right) of the residuals of the ANOVA model"}


# Q-Q plot
par(mfrow = c(1, 2))
qqnorm(residuals)
qqline(residuals, col = "red")

# Histogram
hist(residuals, breaks = 10, main = "Histogram of Residuals",
     xlab = "Residuals", col = "lightblue")
```

For the Shapiro test, the P-value is `r shapiro$p.value`; the null hypothesis is not rejected. The Shapiro test indicates that the residuals are normally distributed.

## Effect Size: Eta-Squared (η²)

Effect size measures the magnitude of the treatment effect:

$$\eta^2 = \frac{SSE}{SSTotal}$$
For our example, the η² is `r round(sse/ss_total, digit = 2)` for the fertilizer treatment. This means that 34% of the variance in the outcome can be explained by the treatment. 

## Post-Hoc Tests

After running the test and checking all the assumptions, the ANOVA test we ran is valid. The conclusion is that not all plant height means are equal. There are the effects of fertilizer treatment on plant growth.

When ANOVA shows a significant F-statistic, we know that **at least one group mean differs** from the others. However, ANOVA does not tell us:

- Which specific groups are different?
- How many pairwise differences exist?
- The magnitude of these differences?

The next step is to use a post-hoc test.

Post-hoc tests are conducted **after** finding significant ANOVA. The tests compare all possible pairs or specific subsets of treatment.

### Common Post-Hoc Tests

There are many post-hoc tests to choose from (Table \@ref(tab:post-hoc-test)). It is good to keep in mind the features of each test and the purpose of the comparisons. In terms of conservativeness, that the test is conservative means that the method requires stronger evidence to reject the null hypothesis. The method keeps Type I error (false positives) very low, but reducing statistical power to detect real differences.

| Test | Use When | Conservativeness | Special Features |
|------|----------|------------------|------------------|
| **Tukey HSD** | All pairwise comparisons | Moderate | Most popular, good power |
| **Bonferroni** | Few planned comparisons | Conservative | Simple, very strict |
| **Holm** | Multiple comparisons | Less conservative than Bonferroni | Sequential method |
| **Scheffe** | Complex contrasts | Very conservative | Can test any contrast |
| **Dunnett** | Comparing to control | Moderate | Only vs. control group |
| **Fisher LSD** | Protected by significant ANOVA | Liberal | Use cautiously |
| **Games-Howell** | Unequal variances | Moderate | No homogeneity assumption |

: (\#tab:post-hoc-test) Comparisons of Post-Hoc Methods

## Example: Post-hoc Test of Plant Growth Example 

Let's use the fertilizer experiment with three treatments to demonstrate two of the post-hoc tests:

### Tukey's Honestly Significant Difference (HSD)

Tukey's HSD compares all possible pairs of means (All pairwise comparisons among groups) using the Studentized Range Distribution.

- Tukey's HSD works best with **Equal sample sizes** (but it is robust to minor imbalances).
- The test assumes **Homogeneity of variance**.
- Tukey's HSD is the most common post-hoc test.

From the fertilizer experiment, the mean of plant height from Fertilizer A did not significantly from that of the control group. Other pairs of comparison shows significantly differences (Table \@ref(tab:tukey-hsd-tab)).

```{r tukey-hsd, message=FALSE, echo = FALSE, results='hide', warning=FALSE}
# Perform Tukey's HSD test
tukey_results <- TukeyHSD(anova_model, conf.level = 0.95)

print(tukey_results)

# Access the table
tukey_table <- as.data.frame(tukey_results$Fertilizer)
tukey_table$Comparison <- rownames(tukey_table)
tukey_table <- tukey_table[, c("Comparison", "diff", "lwr", "upr", "p adj")]
colnames(tukey_table) <- c("Comparison", "Mean_Diff", "Lower_CI", "Upper_CI", "p_adj")

# Add significance indicator
tukey_table$Significant <- ifelse(tukey_table$p_adj < 0.05, "***", 
                                  ifelse(tukey_table$p_adj < 0.10, "*", "ns"))

```


```{r tukey-hsd-tab, message=FALSE, echo = FALSE, tab.cap= "Tukey's HSD results comparing three treatment pairs", warning=FALSE, results='asis'} 
kbl(tukey_table, digit = 2, row.names = FALSE,
    caption = "Tukey's HSD results comparing three treatment pairs") %>%
kable_styling(bootstrap_options = "striped", full_width = F)
```

The results are visualized with letters to indicate the significant difference (Figure \@ref(fig:visualize-hsd)). The treatments that share the same letters are not significantly different from each other.

```{r visualize-hsd, echo=FALSE, message=FALSE, warning=FALSE, fig.height = 5, fig.cap= "Visualization of Tukey's HSD results"}

if(!require("ggthemes", character.only = TRUE)) {
  install.packages("ggthemes", dependencies = TRUE)}
library(ggplot2)
library(ggthemes)

add_letter <- c(by(plant_data$Height, plant_data$Fertilizer, mean))

# Boxplot
ggplot(plant_data, aes(x = Fertilizer, y = Height, fill = Fertilizer)) +
  geom_boxplot(alpha = 0.6) +
#  geom_jitter(width = 0.1, size = 2) +
#  geom_hline(yintercept = mean(plant_data$Height), 
#             linetype = "dashed", color = "purple", size = 1) +
  annotate("text", x = 1:3, y = add_letter + 2.5, 
           label = c("b", "a", "b"), color = "black", size = 6, fontface = "bold") +
  stat_summary(fun.y = mean, geom = "point", shape = 18, size = 3, color = "darkred") +
  geom_point(x = 2.8, y = 9, shape = 18, size = 3, color = "darkred") +
  annotate("text", x = 3.1, y = 9, 
           label = "=  Mean", color = "black", size = 5) +
  annotate("rect", xmin = 2.6, xmax = 3.5, ymin = 8.5, ymax = 9.5,
           fill = NA, alpha = 0.3, color = "black") +
  labs(
       x = "Fertilizer Treatment",
       y = "Plant height (cm)") +
  ylim(0, 9.5) +
  theme_bw() +
  theme(panel.border = element_blank(), legend.position = "none",
  axis.line.x = element_line(size = 0.5, linetype = "solid", colour = "black"),
  axis.line.y = element_line(size = 0.5, linetype = "solid", colour = "black"),
  axis.text = element_text(size = 14), axis.title = element_text(size = 14))
```

### Dunnett's Test

Dunnett's Test is used to compare treatments to a control group. The test is used when the researchers are not interested in comparing treatments to each other. It is more powerful than Tukey's HSD for this specific purpose. In our example, Dunnet's test is used to compare fertilizers A and B to "No fertilizer" control. It does NOT compare A vs B. It shows that only Fertilizer B give significant taller plants than the control treatment (Table \@ref(tab:dunnett-tab)).

```{r dunnett, message=FALSE, results='hide', warning=FALSE, echo=FALSE}
library(DescTools)

# Dunnett's test (comparing all to No fertilizer control)
dunnett_results <- DunnettTest(
  Height ~ Fertilizer,
  data = plant_data,
  control = "No fertilizer"  # Specify control group
)

#print(dunnett_results)

#cat("\nInterpretation:\n")
#cat("This test compares Fertilizer A and Fertilizer B to the control (No fertilizer)\n")
#cat("It does NOT compare A vs B\n")
```


```{r dunnett-tab, message=FALSE, echo = FALSE, tab.cap= "Dunnett's results comparing each treatments with the control treatment", warning=FALSE, results='asis'} 
kbl(dunnett_results$`No fertilizer`, digit = 2, row.names = TRUE,
    caption = "Dunnett's results comparing each treatments with the control treatment") %>%
kable_styling(bootstrap_options = "striped", full_width = F)
```

## Reporting ANOVA Results
The following paragraph presents the written report for ANOVA.

> "An analysis of variance (ANOVA) was conducted to evaluate the effect of fertilizer treatment (Control, Fertilizer A, and B) on final plant height. There was a significant effect of fertilizer treatment on final plant height (F(2, 18) = `r round(anova_results$"F value"[1], digit = 2)`, P < 0.001, η² = `r round(sse/ss_total, digit = 2)`). 
>
> Post-hoc pairwise comparisons using the Tukey's HSD revealed that Fertilizer B resulted in significantly taller plants (Mean = `r summary_plant$Mean[2]` cm, SE = `r summary_plant$SE[2]`) than both Fertilizer A (Mean = `r summary_plant$Mean[1]` cm, SE = `r summary_plant$SE[1]`, p < 0.001) and the control with no fertilizer (Mean = `r summary_plant$Mean[3]` cm, SE = `r summary_plant$SE[3]`, p < 0.001). Fertilizer A did not produced significantly taller plants than the control (P > 0.05). Fertilizer B is recommended to be used for increase plant height."

## What is ANCOVA?
ANCOVA is a statistical technique that combines ANOVA and regression. It allows us to compare group means (like ANOVA) and control for the effect of continuous variables (covariates).

### Benefits of ANCOVA

Sometimes we cannot randomly assign subjects to groups, or we want to control for pre-existing differences. Even with randomization, groups may differ at baseline on important variables. By accounting for variation due to covariates, we reduce error variance and increase our ability to detect treatment effects. 

## Example: Fertilizer Experiment with Initial Seedling Size

Researchers want to compare three fertilizer treatments (Organic, Synthetic, and Control - no fertilizer) on final plant height (Table \@ref(tab:ancova-data-tab)). However, seedlings have different initial sizes when the experiment begins, which could confound our results.

**Research Design:**
-   **Independent Variable (Treatment):** Fertilizer type (3 levels)
-   **Dependent Variable (Outcome):** Final plant height (cm) after 8 weeks
-   **Covariate:** Initial seedling height (cm) at the start of the experiment

**Why use ANCOVA here?**
Even with random assignment, seedlings naturally vary in initial size. By controlling for initial height, we can more accurately assess the effect of fertilizer treatments on growth.

```{r create-data, echo =FALSE, message=FALSE, warning=FALSE, results='hide'}
library(dplyr)
library(ggplot2)

set.seed(213708)

# Number of seedlings per treatment
n_per_group <- 15

# Create the dataset
fertilizer_data <- data.frame(
  Seedling_ID = 1:(3 * n_per_group),
  Fertilizer = factor(rep(c("Control", "Organic", "Synthetic"), each = n_per_group),
                      levels = c("Control", "Organic", "Synthetic")),
  Initial_Height = c(
    rnorm(n_per_group, 8, 2),    # Control - similar initial heights
    rnorm(n_per_group, 8.5, 2),  # Organic - slightly taller initially
    rnorm(n_per_group, 7.8, 2)   # Synthetic - slightly shorter initially
  )
)

# Generate final heights based on initial height + treatment effect + growth rate
fertilizer_data <- fertilizer_data %>%
  mutate(
    Treatment_Effect = case_when(
      Fertilizer == "Control" ~ 12,      # Baseline growth
      Fertilizer == "Organic" ~ 18,      # Moderate enhancement
      Fertilizer == "Synthetic" ~ 24     # Strong enhancement
    ),
    # Final height = Initial height + Treatment effect + random growth variation
    Final_Height = Initial_Height + Treatment_Effect + rnorm(3 * n_per_group, 0, 2.5)
  )

fertilizer_data$Initial_Height <- round(fertilizer_data$Initial_Height, 1)

fertilizer_data$Final_Height <- round(fertilizer_data$Final_Height, 1)
# Display first few rows
#cat("Fertilizer Experiment Dataset\n")
#cat("============================\n\n")
#head(fertilizer_data, 12)

# Summary statistics
#cat("\n\nSummary Statistics by Fertilizer Treatment:\n")
summary_stats <- fertilizer_data %>%
  group_by(Fertilizer) %>%
  summarise(
    n = n(),
    Initial_Mean = round(mean(Initial_Height), 1),
    Initial_SD = round(sd(Initial_Height), 1),
    Final_Mean = round(mean(Final_Height), 1),
    Final_SD = round(sd(Final_Height), 2),
    Growth = round(Final_Mean - Initial_Mean, 1),
    .groups = 'drop'
  )

#print(summary_stats)
```

```{r ancova-data-tab, echo=FALSE, message=FALSE, warning=FALSE, tab.cap="The data of fertilizer experiment for ANCOVA", results='asis'}
### Write the data into a table
library(knitr)
library(kableExtra)

fertilizer_data[,-4] %>%
  kbl(align = "c",
      col.names = c("Seedling ID", "Fertilizer", "Initial Height (cm)",
                    "Final Height (cm)"), 
      caption = "Initial and Final height under different fertilizer treatments") %>% 
  kable_styling(full_width = F) %>%
  column_spec(1, bold = TRUE)

```

### Visualizing the Data

First, let's explore the data before testing. The research should make the summary of the data separated by the factor of interest (Fertilizer) in Table \@ref(tab:summary-data-ancova). In addition, let's plot some graph showing the relationship between initial and final height of plants (Figure \@ref(fig:visualize-data))

```{r summary-data-ancova, echo=FALSE, message=FALSE, warning=FALSE, tab.cap="Summary of fertilizer experiment data by treatment", results='asis'}
summary_stats %>%
  kbl(align = "c",
      col.names = c("Fertilizer type", "n", "Mean initial height (cm)", 
                    "Initial height SD (cm)", "Mean final height (cm)", "Final height SD (cm)", "Absolute growth (cm)"),
      caption = "Summary of plant height separated by fertilizer treatments") %>% 
  kable_styling(full_width = F) %>%
  column_spec(1, bold = TRUE)
```

```{r visualize-data, fig.height=8, echo =FALSE, message=FALSE, warning=FALSE, fig.cap="Scatter plot (A.) shows the relationship between initial and final plant height and boxplots of initial (B.) and final heights (C.)"}

library(patchwork)

# Scatter plot: Initial vs Final Height by Fertilizer
p1 <- ggplot(fertilizer_data, aes(x = Initial_Height, y = Final_Height, color = Fertilizer)) +
  geom_point(size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", se = FALSE, size = 1.2) +
  labs(title = "A.",
       #subtitle = "Colored by fertilizer treatment",
       x = "Initial Seedling Height (cm)",
       y = "Final Plant Height (cm)") +
  theme_classic() +
  theme(legend.position = "none", 
        axis.text = element_text(size = 14), axis.title = element_text(size = 14))

# Boxplot: Initial heights
p2 <- ggplot(fertilizer_data, aes(x = Fertilizer, y = Initial_Height, fill = Fertilizer)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.1, alpha = 0.5) +
  labs(title = "B.",
       #subtitle = "Shows baseline differences",
       x = "Fertilizer Treatment",
       y = "Initial Height (cm)") +
  theme_classic() +
  theme(legend.position = "none", 
        axis.text = element_text(size = 14), axis.title = element_text(size = 14))
  

# Boxplot: Final heights
p3 <- ggplot(fertilizer_data, aes(x = Fertilizer, y = Final_Height, fill = Fertilizer)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.1, alpha = 0.5) +
  labs(title = "C.",
       #subtitle = "Outcome variable",
       x = "Fertilizer Treatment",
       y = "Final Height (cm)") +
  theme_classic() +
  theme(legend.position = "none", 
        axis.text = element_text(size = 14), axis.title = element_text(size = 14))

# Combine plots
p1 / (p2 + p3)
```

## Comparison: ANOVA vs. ANCOVA

Let's compare what happens when we ignore vs. control for initial seedling height.

### Analysis 1: ANOVA (Ignoring Initial Height)

Table \@ref(tab:anova-only-tab) presents the result of ANOVA that did not take into the account of initial height. The conclusion is that at least one mean differs from the others.

```{r anova-only, echo =FALSE, message=FALSE, warning=FALSE}
# Traditional ANOVA - ignoring initial height
anova_model_cov <- aov(Final_Height ~ Fertilizer, data = fertilizer_data)
#summary(anova_model_cov)

# Extract key statistics
anova_results_cov <- anova(anova_model_cov)

#cat("\nANOVA Results (Without Covariate):\n")
#cat("F-statistic:", round(anova_f, 3), "\n")
#cat("p-value:", round(anova_p, 4), "\n")
#cat("\nThis analysis ignores initial seedling size differences.\n")
```

```{r anova-only-tab, message=FALSE, echo = FALSE, include=FALSE, tab.cap= "ANOVA results that ignore initial height", warning=FALSE, results='asis'} 
kbl(anova_results_cov, digit = 2, row.names = TRUE) %>%
kable_styling(bootstrap_options = "striped", full_width = F)
```

| Source of Variation | Sum Square | df | Mean Square (MS) | F | P-value |
|---------------------|----|----|------------------|---|------------|
| Treatment (Between) | `r round(anova_results_cov$"Sum Sq"[1], digit = 1)` | `r round(anova_results_cov$"Df"[1], digit = 1)` | `r round(anova_results_cov$"Mean Sq"[1], digit = 1)` | `r round(anova_results_cov$"F value"[1], digit = 1)` | `r format_sci(anova_results_cov$"Pr(>F)"[1], digits = 2)` |
| Error (Within) | `r round(anova_results_cov$"Sum Sq"[2], digit = 1)` | `r round(anova_results_cov$"Df"[2], digit = 1)` | `r round(anova_results_cov$"Mean Sq"[2], digit = 1)` |  |  |
| Total | `r round(anova_results_cov$"Sum Sq"[1] + anova_results_cov$"Sum Sq"[2] , digit = 1)` | `r anova_results_cov$"Df"[1] + anova_results_cov$"Df"[2]` |  |  |  |

: (#tab:anova-only-tab) ANOVA Results of Fertilizer Experiment (This analysis ignores the effect of initial height)

### Analysis 2: ANCOVA (Controlling for Initial Height)

Table \@ref(tab:ancova-tab) presents the result of ANCOVA that take into the account of initial height. The conclusion is that at least one mean differs from the others. There is also significant effect on initial height on final height of the plants. In comparison, ANCOVA increases statistical power.

```{r ancova, echo =FALSE, message=FALSE, warning=FALSE, results='hide'}
# ANCOVA with initial height as covariate
ancova_model <- aov(Final_Height ~ Initial_Height + Fertilizer, data = fertilizer_data)
summary(ancova_model)

# Extract key statistics
ancova_results <- anova(ancova_model)
#ancova_f <- ancova_results[[1]]["Fertilizer", "F value"]
#ancova_p <- ancova_results[[1]]["Fertilizer", "Pr(>F)"]

#cat("\nANCOVA Results (With Covariate):\n")
#cat("F-statistic:", round(ancova_f, 3), "\n")
#cat("p-value:", round(ancova_p, 4), "\n")

# Effect of covariate
#covariate_f <- ancova_results[[1]]["Initial_Height", "F value"]
#covariate_p <- ancova_results[[1]]["Initial_Height", "Pr(>F)"]
#cat("\nCovariate (Initial Height) Effect:\n")
#cat("F-statistic:", round(covariate_f, 3), "\n")
#cat("p-value:", round(covariate_p, 4), "\n")
#cat("\nThis shows that initial height is a significant predictor of final height.\n")
```

| Source of Variation | Sum Square | df | Mean Square (MS) | F | P-value |
|---------------------|----|----|------------------|---|------------|
| Initial Height (covariate) | `r round(ancova_results$"Sum Sq"[1], digit = 1)` | `r round(ancova_results$"Df"[1], digit = 1)` | `r round(ancova_results$"Mean Sq"[1], digit = 1)` | `r round(ancova_results$"F value"[1], digit = 1)` | `r format_sci(ancova_results$"Pr(>F)"[1], digits = 2)` |
| Treatment (Between Fertilizer) | `r round(ancova_results$"Sum Sq"[2], digit = 1)` | `r round(ancova_results$"Df"[2], digit = 1)` | `r round(ancova_results$"Mean Sq"[2], digit = 1)` | `r round(ancova_results$"F value"[2], digit = 1)` | `r format_sci(ancova_results$"Pr(>F)"[2], digits = 2)` |
| Error | `r round(ancova_results$"Sum Sq"[3], digit = 1)` | `r round(ancova_results$"Df"[3], digit = 1)` | `r round(ancova_results$"Mean Sq"[3], digit = 1)` |  |  |
| Total | `r round(sum(ancova_results$"Sum Sq") , digit = 1)` | `r sum(ancova_results$"Df")`|  |  |  |

: (\#tab:ancova-tab) ANCOVA results of the Fertilizer Treatment

```{r comparison, echo =FALSE, message=FALSE, warning=FALSE, tab.cap="Comparisons fo F statistics and P-value of ANOVA and ANCOVA", include=FALSE, results='asis'}
comparison <- data.frame(
  Analysis = c("ANOVA", "ANCOVA"),
  F_statistic = c(anova_results$"F value"[1], ancova_results$"F value"[2]),
  P_value = format(c(anova_results$"Pr(>F)"[1], ancova_results$"Pr(>F)"[2]),
  Significant = c(anova_results$"Pr(>F)"[1] < 0.05, ancova_results$"Pr(>F)"[2]) < 0.05)
)

#print(comparison)

#cat("\n*** Notice how ANCOVA increases statistical power! ***\n")

kbl(comparison, digit = 2, row.names = TRUE) %>%
kable_styling(bootstrap_options = "striped", full_width = F)
```


## Understanding Adjusted Means

### What Are Adjusted Means?

In ANCOVA, we calculate **adjusted means** - the expected group means if all groups had the same average covariate value. When taking the initial height into account, the adjusted mean of seedlings in the organic fertilizer treatment is slightly lower, while the seedling height mean of the synthetic fertilizer treatment is slightly higher than unadjusted means (Figure \@ref(fig:plot-adjusted-means)).

```{r adjusted-means, results='hide', echo=FALSE, warning=FALSE, message=FALSE}
library(emmeans)

# Calculate adjusted means
adjusted_means <- emmeans(ancova_model, ~ Fertilizer)
#print(adjusted_means)

# Compare with unadjusted means
unadjusted_means <- fertilizer_data %>%
  group_by(Fertilizer) %>%
  summarise(Unadjusted_Mean = mean(Final_Height))

# Combine for comparison
adjusted_df <- as.data.frame(adjusted_means)

comparison_means <- unadjusted_means %>%
  left_join(
    adjusted_df,
    by = "Fertilizer") %>%
  rename(adjusted_mean = emmean)

#print(comparison_means)
```

```{r plot-adjusted-means, echo=FALSE, warning=FALSE, message=FALSE, fig.height=6, fig.cap= "Comparisons of unadjusted and adjusted means" ,}
# Reshape for plotting
#if(!require("ggpattern", character.only = TRUE)) {
#  install.packages("ggpattern", dependencies = TRUE)}
library(tidyr)
#library(ggpattern)

comparison_long <- comparison_means %>%
  pivot_longer(cols = c(Unadjusted_Mean, adjusted_mean),
               names_to = "Type",
               values_to = "Mean")

ggplot(comparison_long, aes(x = Fertilizer, y = Mean, fill = Type)) +
  geom_col(position = "dodge", alpha = 0.7, color = "black") +
  geom_text(aes(label = round(Mean, 1)), 
            position = position_dodge(width = 0.9),
            vjust = -0.5, size = 3.5) +
  scale_fill_manual(values = c("Unadjusted_Mean" = "grey", "adjusted_mean" = "white")) +
  labs(title = "Adjusted means control for initial seedling height differences",
       #subtitle = "",
       y = "Final Plant Height (cm)",
       x = "Fertilizer Treatment",
       fill = "Mean Type") +
  theme_classic() +
  theme(legend.position = "bottom", 
        axis.text = element_text(size = 14), axis.title = element_text(size = 14))
  
```

**Interpretation:** The adjusted means show what the final heights would be if all seedlings started at the same initial height. This gives us a fairer comparison of the fertilizer treatments' true effects.

## Assumptions of ANCOVA

ANCOVA has all the assumptions of ANOVA, **plus additional ones**:

### Linear Relationship

The relationship between the covariate (initial height) and dependent variable (final height) should be linear.

If the relationship is linear, the blue and red lines should be similar (Figure \@ref(fig:check-linearity)), and residuals should show no pattern (Figure \@ref(fig:check-residuals)).

```{r check-linearity, echo = FALSE, message=FALSE, warning=FALSE, fig.cap="Linearity Check: Initial vs. Final Height"}
# Scatter plot with linear fit
ggplot(fertilizer_data, aes(x = Initial_Height, y = Final_Height)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = FALSE, color = "blue", size = 1.2) +
  geom_smooth(method = "loess", se = FALSE, color = "red", linetype = "dashed") +
  labs(title = "Blue line = linear fit, Red line = loess (flexible) fit",
       x = "Initial Seedling Height (cm)",
       y = "Final Plant Height (cm)") +
  theme_classic() +
  theme(legend.position = "bottom", 
        axis.text = element_text(size = 14), axis.title = element_text(size = 14))
```

```{r check-residuals, echo = FALSE, message=FALSE, warning=FALSE, fig.cap="Residuals and covariate (Initial height) plot"}
# Check residuals vs. covariate
plot(fertilizer_data$Initial_Height, residuals(ancova_model),
     xlab = "Initial Height (cm)", ylab = "Residuals",
     pch = 16, col = rgb(0, 0, 1, 0.5))
abline(h = 0, col = "red", lty = 2, lwd = 2)
```

### Homogeneity of Regression Slopes

**Critical assumption:** The relationship between initial height and final height should be the same across all fertilizer treatments (parallel slopes) (Figure \@ref(fig:check-slopes)).

```{r check-slopes, fig.height=6, fig.cap="Homogeneity of Slopes Test", echo=FALSE, message=FALSE, warning=FALSE}
# Visual check: separate regression lines by group
ggplot(fertilizer_data, aes(x = Initial_Height, y = Final_Height, color = Fertilizer)) +
  geom_point(alpha = 0.6, size = 2) +
  geom_smooth(method = "lm", se = TRUE, alpha = 0.2, size = 1.2) +
  labs(title = "Lines should be roughly parallel (similar slopes)",
       x = "Initial Seedling Height (cm)",
       y = "Final Plant Height (cm)") +
  theme_classic() +
  theme(legend.position = "bottom",
        axis.text = element_text(size = 14), axis.title = element_text(size = 14))

# Statistical test for homogeneity of slopes
# Test interaction between Fertilizer and Initial_Height
interaction_model <- aov(Final_Height ~ Initial_Height * Fertilizer, data = fertilizer_data)
#summary(interaction_model)

# Check the interaction term
interaction_p <- summary(interaction_model)[[1]]["Initial_Height:Fertilizer", "Pr(>F)"]

#summary(interaction_model)[[1]]["Initial_Height:Fertilizer", "F value"]
#cat("\nHomogeneity of Slopes Test:\n")
#cat("Interaction p-value:", round(interaction_p, 4), "\n")

#if (interaction_p < 0.05) {
#  cat("\n⚠ WARNING: Slopes are NOT homogeneous! ANCOVA may not be appropriate.\n")
#  cat("The relationship between initial and final height differs across fertilizer treatments.\n")
#} else {
#  cat("\n✓ Good: Slopes are homogeneous. ANCOVA assumptions are met.\n")
#  cat("The relationship between initial and final height is similar across all treatments.\n")
#}
```

**Why this matters:** If slopes differ, it means the fertilizer treatments affect not just overall growth, but how much plants respond to their initial size. In that case, ANCOVA's assumption is violated.

### Independence of Covariate and Treatment

The covariate should not be affected by the treatment. This means:

- The covariate should be measured **before** treatment
- The covariate should not be influenced by group assignment

In our fertilizer example:
- Initial seedling height was measured **before** fertilizer treatments were applied
- Random assignment ensures fertilizer type doesn't affect initial height
- The covariate (initial size) is a pre-existing characteristic

**Counter-example where this would be violated:**
If we measured "leaf chlorophyll content" after one week of fertilizer treatment and used it as a covariate, this would be wrong because the fertilizer already influenced chlorophyll levels.

### Standard ANOVA Assumptions

Researchers also should check for:
- **Normality of residuals**
- **Homogeneity of variance**
- **Independence of observations**

```{r check-standard-assumptions, include=FALSE, echo=FALSE}
# Normality check
par(mfrow = c(1, 2))
qqnorm(residuals(ancova_model), main = "Q-Q Plot of Residuals")
qqline(residuals(ancova_model), col = "red", lwd = 2)
hist(residuals(ancova_model), breaks = 15, main = "Histogram of Residuals",
     xlab = "Residuals", col = "lightblue", border = "white")
par(mfrow = c(1, 1))

# Shapiro-Wilk test for normality
shapiro_test <- shapiro.test(residuals(ancova_model))
cat("Shapiro-Wilk Test for Normality:\n")
cat("W =", round(shapiro_test$statistic, 4), ", p-value =", round(shapiro_test$p.value, 4), "\n")
if (shapiro_test$p.value > 0.05) {
  cat("✓ Residuals are normally distributed (p > 0.05)\n\n")
} else {
  cat("⚠ Residuals may not be normally distributed (p < 0.05)\n\n")
}

# Homogeneity of variance (Levene's test)
library(car)
levene_test <- leveneTest(Final_Height ~ Fertilizer, data = fertilizer_data)
cat("Levene's Test for Homogeneity of Variance:\n")
print(levene_test)
if (levene_test$`Pr(>F)`[1] > 0.05) {
  cat("\n✓ Variances are equal across groups (p > 0.05)\n")
} else {
  cat("\n⚠ Variances may not be equal across groups (p < 0.05)\n")
}

# Comprehensive diagnostic plots
par(mfrow = c(2, 2))
plot(ancova_model)
par(mfrow = c(1, 1))
```

## Post-Hoc Comparisons with ANCOVA and Visualizing Post-Hoc Results

When ANCOVA shows significant group differences, we use a Tukeys's HSD post-hoc test to identify which groups differ (Figure \@ref(fig:plot-posthoc)).

```{r posthoc-ancova, echo=FALSE, results='hide', warning=FALSE, message=FALSE}
# Pairwise comparisons of adjusted means
pairwise_results <- pairs(adjusted_means)
print(pairwise_results)

# Confidence intervals
confint(pairwise_results)

# Compact letter display
library(multcomp)
cld(adjusted_means, Letters = letters)
```

```{r plot-posthoc, fig.height=6, fig.cap="Adjusted Means with Post-Hoc Comparisons. Groups with different letters are significantly different.", echo=FALSE, warning=FALSE, message=FALSE}
#if(!require("mulcompView", character.only = TRUE)) {
#  install.packages("mulcompView", dependencies = TRUE, 
#                   repos = "http://cran.us.r-project.org")}
#library(multcompView)

# Plot adjusted means with error bars and significance letters
adj_means_df <- as.data.frame(adjusted_means)

# Get compact letter display
cld_result <- cld(adjusted_means, Letters = letters)
cld_df <- as.data.frame(cld_result)

ggplot(cld_df, aes(x = Fertilizer, y = emmean, fill = Fertilizer)) +
  geom_col(alpha = 0.7) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), 
                width = 0.2, size = 1) +
  geom_text(aes(label = .group, y = upper.CL + 2), 
            size = 5, fontface = "bold") +
  labs(title = "Error bars show 95% confidence intervals",
       y = "Adjusted Mean of Plant Height") +
 theme_classic() +
  theme(legend.position = "bottom", 
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 14),
        axis.text = element_text(size = 14), axis.title = element_text(size = 14))
```

### Effect Size for ANCOVA - Partial Eta-Squared

For ANCOVA, we use **partial eta-squared** which shows the proportion of variance in the outcome explained by each effect, after controlling for other effects (Table \@ref(tab:effect-size-ancova)).

The interpretation of effect size
Small: η² = 0.01
Medium: η² = 0.06
Large: η² = 0.14

```{r effect-size-ancova, tab.cap="Partial eta-squared of ANCOVA model of the fertilizer experiment", echo =FALSE, warning=FALSE, message=FALSE, results='asis'}
library(effectsize)

# Calculate effect sizes
partial_eta <- eta_squared(ancova_model, partial = TRUE)

kbl(eta_squared(ancova_model, partial = TRUE), digit = 2, row.names = FALSE,
    caption = "Partial eta-squared of ANCOVA model of the fertilizer experiment") %>%
kable_styling(bootstrap_options = "striped", full_width = F)

# Interpretation
#cat("\nEffect Size Interpretation:\n")
#cat("Small: η² = 0.01\n")
#cat("Medium: η² = 0.06\n")
#cat("Large: η² = 0.14\n")
```

## Multiple Covariates

ANCOVA can include multiple covariates to further improve precision. For example, with the fertilizer experiment, adding one more covariate of leaf number to reflect seedling vigor at the beginning of the experiment will give us a more precise estimate of fertilizer effects.

## When NOT to Use ANCOVA

ANCOVA is **not appropriate** when:

1. **Slopes are not homogeneous** (interaction between covariate and treatment)
   - Solution: Consider separate regression analyses for each group
   
2. **Covariate is affected by treatment**
   - Solution: Use the covariate as an outcome variable or use different design
   
3. **Relationship is non-linear**
   - Solution: Transform variables or use polynomial terms
   
4. **Covariate has measurement error**
   - Solution: Consider structural equation modeling or reliability correction

## Reporting ANCOVA Results of the Fertilizer Experiment

> "An analysis of covariance (ANCOVA) was conducted to evaluate the effect of fertilizer treatment (Control, Organic, Synthetic) on final plant height after 8 weeks, while controlling for initial seedling height. The assumption of homogeneity of regression slopes was tested and satisfied (F(2, 39) = `r round(summary(interaction_model)[[1]]["Initial_Height:Fertilizer", "F value"], digit = 2)`, P = `r round(summary(interaction_model)[[1]]["Initial_Height:Fertilizer", "Pr(>F)"], digit = 2)`), indicating that the relationship between initial and final height was similar across all fertilizer treatments. 
>
> The covariate, initial seedling height, was significantly related to final plant height (F(1, 40) = `r round(ancova_results$"F value"[1], digit = 2)`), P < 0.01, partial η² = `r round(partial_eta[1,2], digit = 2)`), demonstrating that larger seedlings at the start of the experiment grew to be larger plants. After controlling for initial seedling height, there was a significant effect of fertilizer treatment on final plant height (F(2, 40) = `r round(ancova_results$"F value"[2], digit = 2)`, P < 0.001, partial η² = `r round(partial_eta[2,2], digit = 2)`). 
>
> Post-hoc pairwise comparisons using the Bonferroni correction revealed that synthetic fertilizer (Adjusted Mean = `r round(comparison_means$adjusted_mean[3], digit = 1)` cm, SE = `r round(comparison_means$SE[3], digit = 1)`) resulted in significantly taller plants than both organic fertilizer (Adjusted Mean = `r round(comparison_means$adjusted_mean[2], digit = 1)` cm, SE = `r round(comparison_means$SE[2], digit = 1)`), p < 0.001) and the control with no fertilizer (Adjusted Mean = `r round(comparison_means$adjusted_mean[1], digit = 1)` cm, SE = `r round(comparison_means$SE[1], digit = 1)`), p < 0.001). Organic fertilizer also produced significantly taller plants than the control (P < 0.001), demonstrating that both fertilizer types enhanced plant growth beyond the baseline."

In summay, ANOVA and ANCOVA are common techniques to determine differences in mean among categorical explanatory variables. The techniques have assumptions that researchers must check before and after running the analysis. 

## Exercise

**Scenario A:** (For questions 1 and 2) 
You have collected data on seedling growth under four different light conditions (Full sun, Partial shade, Full shade, and Artificial light). You measured the final height of 10 seedlings per treatment.

1. State the null and alternative hypotheses in words.

2. Arrange the following steps in the correct order for performing a one-way ANOVA:

A. Conduct post-hoc tests if ANOVA is significant  
B. Check assumptions (normality, homogeneity of variance)  
C. State the research question and hypotheses
D. Calculate and interpret effect size (η²)  
E. Run the ANOVA and examine the F-statistic and p-value  
F. Visualize the data (boxplots, means plot)
G. Report results with appropriate statistics  

**Correct order for question 2:** ___ → ___ → ___ → ___ → ___ → ___ →

3. For each research scenario, determine whether ANOVA or ANCOVA is more appropriate. Explain your reasoning and identify any covariates.

**Scenario B:**
A researcher compares plant yield across four irrigation methods. The fields have different baseline soil moisture levels (measured before irrigation started) and different initial nutrient levels (also measured beforehand).

- **Best analysis:** ☐ ANOVA ☐ ANCOVA
- **Reasoning:**
- **Covariates (if ANCOVA):**

**Scenario C:**
Testing whether three different pesticides affect insect populations. One week after application, insect counts are recorded. The researcher also measures plant health at that same time point.

- **Best analysis:** ☐ ANOVA ☐ ANCOVA
- **Can plant health (measured at week 1) be used as a covariate?** ☐ Yes ☐ No
- **Reasoning:**

4. List three benefits of using a covariate in research studies.

5. Interpreting the outputs of a study 

**Scenario D:** A study compared three fertilizer treatment - Control (no fertilizer), Organic, and Synthetic - on tomato yield (kg), controlling for initial soil nitrogen (ppm).

**ANCOVA Results:**
```
                      Df Sum Sq Mean Sq F value   Pr(>F)
Soil_Nitrogen          1  842.3   842.3  125.41  < 0.001
Fertilizer             2  256.8   128.4   19.12  < 0.001
Residuals             36  241.7     6.7
```

**Adjusted Means:**
```
Fertilizer    Adjusted Mean (kg)    SE
Control            15.2             0.8
Organic            21.5             0.7
Synthetic          24.8             0.8
```

**Questions:**

5.1) Is soil nitrogen significantly related to yield? What does this tell you about the importance of controlling for it?

5.2) Is there a significant effect of fertilizer type?

5.3) Calculate partial η² for the fertilizer effect: Formula: SS_fertilizer / (SS_fertilizer + SS_residuals)

5.4 The unadjusted means were: 
- Control = 14.8 kg,
- Organic = 22.1 kg, and
- Synthetic = 24.3 kg. 
Why do the adjusted and unadjusted means differ?

---

## Reference
