# Analysis of Variance and Analysis of Covariance

**Duration:** 2-hour lecture

## Learning outcomes

Students should be able to:

-   Describe common terminology relating to ANOVA
-   Determine steps for performing ANOVA
-   Identify when to use ANCOVA versus ANOVA
-   Explain the purpose of covariates in experimental design
-   Interpret ANOVA and ANCOVA results and post-hoc tests

## Introduction

Analysis of Variance (ANOVA) is a fundamental statistical technique used to test differences between two or more group means. ANOVA extends the t-test to more than two groups. The test determines whether observed differences among group means are statistically significant by partitioning total variability into meaningful sources of variation.

Analysis of Covariance (ANCOVA) is also a statistical techniques that combines ANOVA and regression to evaluate the influence of categorical and continuous predictors on a response variable. ANCOVA incorporates one or more covariates, allowing adjustment for potential confounding variables and increasing the precision of treatment comparisons.

### ANOVA vs. ANCOVA vs. Regression

There are some key different features among the tests (Table \@ref(tab:differences)). A covariate is a continuous (sometimes categorical) variable that is not the main factor of interest in a research study. However, the covariate is included in the analysis because it may influence the dependent variable. By accounting for a covariate, researchers can reduce error variation and obtain a clearer estimate of the treatment of interest.

| Feature               | ANOVA               | ANCOVA                   | Regression                        |
|-----------------|-----------------|-----------------|--------------------|
| Explanatory variables | Categorical only    | Categorical + Continuous | Any combination                   |
| Purpose               | Compare group means | Compare adjusted means   | Predict and explain relationships |
| Covariates            | None                | Yes (continuous)         | Can be included                   |

: (#tab:differences) Key differences among ANOVA, ANCOVA, and regression

For example, suppose researchers are comparing tree growth rates among different forest restoration treatments (e.g., natural regeneration vs. active planting). In this case, soil moisture or initial seedling height could be included as a covariate. These variables are not the main treatment, but they strongly affect growth.

## What is ANOVA?

It is conceptually similar to regression analysis, with one key difference that the explanatory variable is **categorical** (e.g., treatment groups, fertilizer types) (Table \@ref(tab:differences)). ANOVA tells you whether means differ, but **not which specific means** are different from one another. To identify specific differenceswhich treatments are different post-hoc tests (e.g. Tukey's HSD) are needed.

## The Basic Concept of ANOVA

ANOVA works by partitioning the total variation in the data into:

1.  **Variation among groups** (due to treatment effects)
2.  **Variation within groups** (random error)

If the variation among groups is **substantially larger** than the variation within groups, we conclude that the groups have different means.

## One-Way ANOVA

### Hypotheses

For one-way ANOVA with k groups:

**Null Hypothesis (H₀):** μ₁ = μ₂ = μ₃ = ... = μₖ

-   All population means are equal.
-   Interpretation: There is no treatment effect.

**Alternative Hypothesis (H₁):** Not all means are equal.

-   At least one population mean is different from the others. This does NOT mean all means are different; some may be equal
-   Interpretation:There is a treatment effect.

## Example: Plant Growth with Two Fertilizers

Let's work through an example with:

-   **Explanatory variable**: Fertilizer type (A and B)
-   **Response variable**: Plant height (cm)
-   **Sample size**: 7 plants per treatment (14 plants total)

```{r example-data}
# Create example dataset
set.seed(213708)
fertilizer <- rep(c("A", "B", "No fertilizer"), each = 7)
height <- c(
  # Treatment A (lower heights)
  6, 3, 4, 5, 3, 6, 4,
  # Treatment B (higher heights)
  8, 5, 6, 7, 8, 6, 8,
  # control (lower heights)
  3, 2, 4, 5, 3, 3, 2
)

plant_data <- data.frame(
  Plant_ID = 1:21,
  Fertilizer = factor(fertilizer),
  Height = height
)

# Display the data
print(plant_data)

# Summary statistics
library(dplyr)
plant_data %>%
  group_by(Fertilizer) %>%
  summarise(
    n = n(),
    Mean = mean(Height),
    SD = sd(Height),
    Min = min(Height),
    Max = max(Height)
  )
```

### Visualizing the Data

It is always a good practice to explore the data before testing.

```{r summary_stats, fig.height=5, echo = FALSE}
summary_stats <- plant_data %>%
  group_by(Fertilizer) %>%
  summarise(
    n = n(),
    Mean = mean(Height),
    SD = sd(Height),
    SE = SD / sqrt(n),
    .groups = 'drop'
  )
```

```{r visualize-groups, fig.height=5}

if(!require("ggthemes", character.only = TRUE)) {
  install.packages("ggthemes", dependencies = TRUE)}
library(ggplot2)
library(ggthemes)

# Boxplot
ggplot(plant_data, aes(x = Fertilizer, y = Height, fill = Fertilizer)) +
  geom_boxplot(alpha = 0.6) +
  geom_jitter(width = 0.1, size = 2) +
  geom_hline(yintercept = mean(plant_data$Height), 
             linetype = "dashed", color = "purple", size = 1) +
  annotate("text", x = 1.5, y = mean(plant_data$Height) + 0.3, 
           label = "Overall Mean", color = "purple") +
  labs(
       x = "Fertilizer Treatment",
       y = "Plant height (cm)") +
  theme_bw() +
  theme(panel.border = element_blank(), legend.position = "none",
  axis.line.x = element_line(size = 0.5, linetype = "solid", colour = "black"),
  axis.line.y = element_line(size = 0.5, linetype = "solid", colour = "black"),
  axis.text = element_text(size = 14), axis.title = element_text(size = 14))

```
### Understanding Sums of Squares

ANOVA partitions the total variation into components:

#### Total Sum of Squares (SSTotal)

SSTotal measures total variation of all observations from the overall mean:

$$SS_{Total} = \sum_{i=1}^{n}(y_i - \bar{\bar{y}})^2$$

where:

- $y_i$ = value of each data point
- $\bar{\bar{y}}$ = overall mean (grand mean)

#### Sum of Squared Error (SSE)

SSE is also called "within-group variation" or "residual variation":

$$SSE = \sum_{i=1}^{n}(y_i - \bar{y}_A)^2 + \sum_{i=1}^{n}(y_i - \bar{y}_B)^2$$

This represents variation of individuals within each treatment group.

#### Sum of Squares Between (SSB)

SSB is also called "among-group variation" or "treatment variation." It measures variation of group means from the overall mean, weighted by sample size.

#### The Partition

$$SS_{Total} = SS_B + SS_E$$

```{r calculate-ss}
# Calculate sums of squares manually
overall_mean <- mean(plant_data$Height)
group_means <- plant_data %>%
  group_by(Fertilizer) %>%
  summarise(group_mean = mean(Height))

# SSTotal
ss_total <- sum((plant_data$Height - overall_mean)^2)

# SSE (within groups)
sse <- plant_data %>%
  left_join(group_means, by = "Fertilizer") %>%
  mutate(sq_dev = (Height - group_mean)^2) %>%
  summarise(SSE = sum(sq_dev)) %>%
  pull(SSE)

# SSB (between groups)
ssb <- sum(7 * (group_means$group_mean - overall_mean)^2)

# Verify partition
cat("SSTotal =", round(ss_total, digits = 2), "\n")
cat("SSB =", round(ssb, digits = 2), "\n")
cat("SSE =", round(sse, digits = 2), "\n")
cat("SSB + SSE =", round(ssb + sse, digits = 2), "\n")
```

### Degrees of Freedom

The concept of degrees of freedom (df) was introduced by Johann Carl Friedrich Gauss in 1821. It represents the **number of values in a calculation that are free to vary**.

For our fertilizer example:

- **df for SSB** = k - 1 = 3 - 1 = 2 (where k = number of treatments)
- **df for SSE** = k(n - 1) = 3(7 - 1) = 18 (where n = replicates per treatment)
- **df for SSTotal** = N - 1 = 21 - 1 = 20 (where N = total observations)

### The F-Distribution and F-Statistic

The F-distribution is used to test whether two variances are equal. The F-statistic is the ratio:

$$F = \frac{\text{Variance among groups}}{\text{Variance within groups}} = \frac{MSB}{MSE}$$

where:

- $MSB = \frac{SSB}{df_SSB}$ (Mean Square Between)
- $MSE = \frac{SSE}{df_SSE}$ (Mean Square Error)

**Interpretation:**

- If F is close to 1, sample variances are equal (no treatment effect)
- If F is large, variation among groups exceeds variation within groups (treatment effect likely)

### The ANOVA Table

The standard format for presenting ANOVA results:

| Source of Variation | SS | df | Mean Square (MS) | F | Critical F |
|---------------------|----|----|------------------|---|------------|
| Treatment (Between) | SSB | k-1 | MSB = SSB/(k-1) | F = MSB/MSE | From F-table |
| Error (Within) | SSE | k(n-1) | MSE = SSE/[k(n-1)] | - | - |
| Total | SSTotal | kn-1 | - | - | - |

## Running ANOVA in R

```{r run-anova}
# Fit the model
anova_model <- aov(Height ~ Fertilizer, data = plant_data)

# Display ANOVA table
summary(anova_model)

# More detailed output
anova_results <- anova(anova_model)
print(anova_results)

# Extract key values
f_statistic <- anova_results$`F value`[1]
p_value <- anova_results$`Pr(>F)`[1]
df_treatment <- anova_results$Df[1]
df_error <- anova_results$Df[2]

cat("\nF-statistic:", round(f_statistic, 3), "\n")
cat("p-value:", round(p_value, 4), "\n")
cat("Degrees of freedom: Treatment =", df_treatment, ", Error =", df_error, "\n")
```

## Assumptions of ANOVA
It is important that before interpreting ANOVA results, these assumptions must be verified:

### 1. Random Sampling
Data should be collected through random sampling.

### 2. Homogeneity of Variance (Homoscedasticity)
Variances should be equal across groups.

**Check BEFORE running ANOVA:**

There are several tests for homogenerity of variance such as Levene's test below. It is a hypothesis testing with a null hypothesis of equal variances. For visual check, boxplots should have similar spreads (see Figure (\@ref(fig:visualize-group))).

```{r levene-test}
# Levene's test for homogeneity of variance
library(car)
leveneTest(Height ~ Fertilizer, data = plant_data)

# Visual check
#ggplot(plant_data, aes(x = Fertilizer, y = Height)) +
#  geom_boxplot() +
#  labs(title = "Check for Equal Variances",
#       subtitle = "Boxplots should have similar spreads") +
#  theme_minimal()
```

In our case, the result of the Levene's Test (P value  = 0.7) indicates that the null hypothesis is not rejected. The variances among groups are homogenous.

### 3. Independence of Errors
Residuals should be independent of each other.

### 4. Normality of Residuals
Residuals should follow a normal distribution.

**Check AFTER running ANOVA:**

This assumption is checked using visualization and Shapiro-Wilk noremality test.

```{r check-normality}
# Extract residuals
residuals <- residuals(anova_model)

# Q-Q plot
par(mfrow = c(1, 2))
qqnorm(residuals)
qqline(residuals, col = "red")

# Histogram
hist(residuals, breaks = 10, main = "Histogram of Residuals",
     xlab = "Residuals", col = "lightblue")

# Shapiro-Wilk test
shapiro.test(residuals)
```

The p-value is 0.30; the null hypothesis is not rejected. The Shapiro test indicates that the residuals are normally distributed.

```{r residual-plots, fig.height=8}
# Comprehensive diagnostic plots
par(mfrow = c(2, 2))
plot(anova_model)
par(mfrow = c(1, 1))
```

## Effect Size: Eta-Squared (η²)

Effect size measures the magnitude of the treatment effect:

$$\eta^2 = \frac{SS_{effect}}{SS_{Total}}$$

## The results of one-way ANOVA
After running the test and checking all the assumptions, the ANOVA test we ran is valid. The conclusion is that not all plant height means are equal. There are the effects of fertilizer treatment on plant growth.

When ANOVA shows a significant F-statistic, we know that **at least one group mean differs** from the others. However, ANOVA doesn't tell us:

- Which specific groups are different?
- How many pairwise differences exist?
- The magnitude of these differences?

The next step is to use post-hoc tests

## Post-Hoc Tests (A Posteriori)

- Conducted **after** finding significant ANOVA
- Exploratory
- Compare all possible pairs or specific subsets
- Require stronger error rate control

### Common Post-Hoc Tests

### Overview of Methods

There are many post-hoc tests to choose from (Table \@ref(tab: post-hoc-list)). It is good to keep in mind the features of each test and the purpose of the comparisons. In terms of conservativeness, that the test is conservative means that the method requires stronger evidence to reject the null hypothesis. The method keeps Type I error (false positives) very low, but reducing statistical power to detect real differences.

| Test | Use When | Conservativeness | Special Features |
|------|----------|------------------|------------------|
| **Tukey HSD** | All pairwise comparisons | Moderate | Most popular, good power |
| **Bonferroni** | Few planned comparisons | Conservative | Simple, very strict |
| **Holm** | Multiple comparisons | Less conservative than Bonferroni | Sequential method |
| **Scheffe** | Complex contrasts | Very conservative | Can test any contrast |
| **Dunnett** | Comparing to control | Moderate | Only vs. control group |
| **Fisher LSD** | Protected by significant ANOVA | Liberal | Use cautiously |
| **Games-Howell** | Unequal variances | Moderate | No homogeneity assumption |

: (#tab:post-hoc-test) Comparisons of Post-Hoc Methods

## Example: Post-hoc Test of Plant Growth Example 

Let's use the fertilizer experiment with three treatments to demonstrate two of the post-hoc tests:

### Tukey's Honestly Significant Difference (HSD)

Tukey's HSD compares all possible pairs of means (**All pairwise comparisons** among groups) using the Studentized Range Distribution.

- Tukey's HSD works best with **Equal sample sizes** (but it is robust to minor imbalances).
- The test assumes **Homogeneity of variance**.
- Tukey's HSD is the most post-hoc test

```{r tukey-hsd}
# Perform Tukey's HSD test
tukey_results <- TukeyHSD(anova_model, conf.level = 0.95)

print(tukey_results)

# Access the table
tukey_table <- as.data.frame(tukey_results$Fertilizer)
tukey_table$Comparison <- rownames(tukey_table)
tukey_table <- tukey_table[, c("Comparison", "diff", "lwr", "upr", "p adj")]
colnames(tukey_table) <- c("Comparison", "Mean_Diff", "Lower_CI", "Upper_CI", "p_adj")

# Add significance indicator
tukey_table$Significant <- ifelse(tukey_table$p_adj < 0.05, "***", 
                                  ifelse(tukey_table$p_adj < 0.10, "*", "ns"))

print(tukey_table)
```

#### Visualization of Tukey's HSD results

```{r visualize-hsd, fig.height = 5}

if(!require("ggthemes", character.only = TRUE)) {
  install.packages("ggthemes", dependencies = TRUE)}
library(ggplot2)
library(ggthemes)

add_letter <- c(by(plant_data$Height, plant_data$Fertilizer, mean))

# Boxplot
ggplot(plant_data, aes(x = Fertilizer, y = Height, fill = Fertilizer)) +
  geom_boxplot(alpha = 0.6) +
#  geom_jitter(width = 0.1, size = 2) +
#  geom_hline(yintercept = mean(plant_data$Height), 
#             linetype = "dashed", color = "purple", size = 1) +
  annotate("text", x = 1:3, y = add_letter + 2.5, 
           label = c("b", "a", "b"), color = "black", size = 6, fontface = "bold") +
  stat_summary(fun.y = mean, geom = "point", shape = 18, size = 3, color = "darkred") +
  geom_point(x = 2.8, y = 9, shape = 18, size = 3, color = "darkred") +
  annotate("text", x = 3.1, y = 9, 
           label = "=  Mean", color = "black", size = 5) +
  annotate("rect", xmin = 2.6, xmax = 3.5, ymin = 8.5, ymax = 9.5,
           fill = NA, alpha = 0.3, color = "black") +
  labs(
       x = "Fertilizer Treatment",
       y = "Plant height (cm)") +
  ylim(0, 9.5) +
  theme_bw() +
  theme(panel.border = element_blank(), legend.position = "none",
  axis.line.x = element_line(size = 0.5, linetype = "solid", colour = "black"),
  axis.line.y = element_line(size = 0.5, linetype = "solid", colour = "black"),
  axis.text = element_text(size = 14), axis.title = element_text(size = 14))
```

### Dunnett's Test

Dunnett's Test is used to compare treatments to a control group. The test is used when the researchers are not interested in comparing treatments to each other. It is more powerful than Tukey's HSD for this specific purpose. In our example, Dunnet's test is used to compare fertilizers A and B to "No fertilizer" control.

```{r dunnett}
library(DescTools)

# Dunnett's test (comparing all to No fertilizer control)
dunnett_results <- DunnettTest(
  Height ~ Fertilizer,
  data = plant_data,
  control = "No fertilizer"  # Specify control group
)

print(dunnett_results)

cat("\nInterpretation:\n")
cat("This test compares Fertilizer A and Fertilizer B to the control (No fertilizer)\n")
cat("It does NOT compare A vs B\n")
```

