# Generalized Linear Regression Models (GLMs) in R {#glm}

*Pimonrat Tiansawat (Chiang Mai University)*

Updated: `r format(Sys.Date(), '%B, %Y')`

**Duration:** 3-hour lab

## Learning outcomes

Students should be able to

1.  demonstrate steps to run GLMs in R
2.  Interpret the results of the models

## Data files for this lab

-   germination_data.csv [Click here to download](https://o365cmu-my.sharepoint.com/:x:/g/personal/pimonrat_t_cmu_ac_th/Edl3RIHvtItIl1gxzwBgOIABaHL9VsQ9ZAzDsssOSgO4Ww?e=Exbr5k)

## Introduction

We consider using GLMs when the dependent (response) variable is:

1.  Count data expressed as proportions
2.  Count data that are not proportions
3.  Binary data
4.  Data on time to death

In our lab today we will work with seed germination data (proportion data) of two tree species *Strychnos nux-vomica* and *Litsea glutinosa*. The germination is tested under two treatment group of dry and wet storage with an additional cold storage (-20 degree celcius) at the driest storage (dry to 5% of initial moisture content). Overall, there are eight treatments - control (baseline), and four dry storage, and three wet storage treatments.

The explanatory (independent) variables are categorical variable - species and treatments.

## Download additional packages

We will use `ggplot2` [@ggplot2] and `dplyr` [@dplyr].

```{r eval = F}
if(!require(ggplot2)){install.packages("ggplot2")}
if(!require(dplyr)){install.packages("dplyr")}

library(ggplot2)
library(dplyr)
```

## The Data Set

-   **genus** and **epithet** = two species
-   **initial.mc.percent** = initial moisture content (%)
-   **treatment.all** = all treatments
-   **treatment.group/group2** = grouping treatment - baseline, dry, wet storage
-   **rep** = three replicates of each treatment
-   **sow** = the number of seeds sown
-   **germinated** = the number of germinated seeds
-   **class** = two types of seed storage behaviours

```{r echo = T, eval=F}
germination <- read.csv('data/germination_data.csv', header = T)

names(germination) ## See column names

#specify the appropriate data types
germination$genus <- as.factor(germination$genus)
germination$epithet <- as.factor(germination$epithet)
germination$treatment.all <- as.factor(germination$treatment.all)
germination$treatment.group <- as.factor(germination$treatment.group)
germination$treatment.group2 <- as.factor(germination$treatment.group2)
germination$rep <- as.factor(germination$rep)
germination$class <- as.factor(germination$class)
```

## Data Visualization

Let's plot some graphs

```{r eval = F}
library(ggplot2)
library(dplyr)

# Calculate germination percent
data <- germination %>%
  mutate(
    species = paste(genus, epithet),
    germination_percent = (germinated / sow) * 100
  )

# Create the plot
# First, see all treatments
ggplot(data, aes(x = treatment.all, y = germination_percent, fill = class)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.6) +
  facet_wrap(~ species, ncol = 2, scales = "free_x") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text = element_text(face = "italic", size = 10),
    legend.position = "top",
    panel.grid.major.x = element_blank()
  ) +
  labs(
    title = "Seed Germination percents by Species and Treatment",
    x = "Treatment",
    y = "Germination (%)",
    fill = "Seed Class"
  ) +
  scale_fill_manual(values = c("orthodox" = "#3498db", "recalcitrant" = "#e74c3c")) +
  ylim(0, 100)

# Create a summary table
summary_stats <- data %>%
  group_by(species, class, treatment.all) %>%
  summarise(
    mean_germ_percent = mean(germination_percent),
    sd_germ_percent = sd(germination_percent),
    n = n(),
se_rate = sd(germination_rate) / sqrt(n()),
    .groups = "drop"
  )

summary_stats
```

Use a line graph to easily compare among treatments.

```{r eval = F}
summary_stats.group <- data %>%
  group_by(species, class, treatment.group2) %>%
  summarise(
    mean_germ_percent = mean(germination_percent),
    sd_germ_percent = sd(germination_percent),
    n = n(),
se_rate = sd(germination_rate) / sqrt(n()),
    .groups = "drop"
  )

ggplot(summary_stats.group, aes(x = treatment.group2, y = mean_germ_percent, 
                         color = class, group = class)) +
  geom_line(linewidth = 1.5, alpha = 0.7) +
  geom_point(size = 2, alpha = 0.9) +
  
  # Add ribbon for error
  geom_ribbon(
    aes(ymin = mean_germ_percent - se_rate, ymax = mean_germ_percent+ se_rate, fill = class),
    alpha = 0.2,
    color = NA
  ) +
  
  scale_color_manual(values = c("orthodox" = "#1982C4", "recalcitrant" = "#FF6B6B")) +
  scale_fill_manual(values = c("orthodox" = "#1982C4", "recalcitrant" = "#FF6B6B")) +
  
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    strip.text = element_text(face = "bold.italic"),
    legend.position = "top",
    panel.grid.minor = element_blank()
  ) +
  
  labs(
    title = "Germination Trends Across Treatments",
    x = "Treatment",
    y = "Mean Germination (%)",
    color = "Seed Class",
    fill = "Seed Class"
  )
```

We can also visual based on treatment groups - dry vs wet storage.

```{r eval = F}
# Compare among baseline germination, and germination after dry and wet storages

ggplot(data, aes(x = treatment.group2, y = germination_percent, fill = class)) +
  geom_boxplot(alpha = 0.7) +
  geom_jitter(width = 0.2, size = 2, alpha = 0.6) +
  facet_wrap(~ species, ncol = 2, scales = "free_x") +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),
    strip.text = element_text(face = "italic", size = 10),
    legend.position = "top",
    panel.grid.major.x = element_blank()
  ) +
  labs(
    title = "Seed Germination Percents by Species and Treatment Group",
    x = "Treatment group",
    y = "Germination (%)",
    fill = "Seed Class"
  ) +
  scale_fill_manual(values = c("orthodox" = "#3498db", "recalcitrant" = "#e74c3c")) +
  ylim(0, 100)
```

A line graph for easily comparison of treatment groups.

```{r eval = F}
ggplot(summary_stats, aes(x = treatment.group2, y = mean_germ_percent, 
                         color = class, group = class)) +
  geom_line(linewidth = 1.5, alpha = 0.7) +
  geom_point(size = 2, alpha = 0.9) +
  
  # Add ribbon for error
  geom_ribbon(
    aes(ymin = mean_germ_percent - se_rate, ymax = mean_germ_percent+ se_rate, fill = class),
    alpha = 0.2,
    color = NA
  ) +
  
  scale_color_manual(values = c("orthodox" = "#1982C4", "recalcitrant" = "#FF6B6B")) +
  scale_fill_manual(values = c("orthodox" = "#1982C4", "recalcitrant" = "#FF6B6B")) +
  
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 9),
    strip.text = element_text(face = "bold.italic"),
    legend.position = "top",
    panel.grid.minor = element_blank()
  ) +
  
  labs(
    title = "Germination Trends Across Treatments",
    x = "Treatment",
    y = "Mean Germination (%)",
    color = "Seed Class",
    fill = "Seed Class"
  )
```

**From the graphs**

The seeds of *Litsea* lost its germinability after dry storage, while the germination of *Strychnos*'s seeds is still high after dry storage. *Litsea*'s seed cannot be dried and stored in the freezer. Storing seeds in the cold conditions seems to reduced the seed germination of *Strychnos*.

## Modeling Steps [@crawley2012]

```{r eval=F}
y <- cbind(germinated, sow-germinated) 
# Create new vector of successful & unsuccessful germination

y

# See levels of explanatory variable
levels(germination$treatment.all)
levels(germination$treatment.group)
levels(germination$treatment.group2)
levels(germination$genus)

model <- glm(y ~ genus * treatment.group2, data = germination, family = binomial)

# 2 x 8 factorial analysis

summary(model)

```

## Model Evaluation

First, let's look at the ration of the residual deviance and the df in the summary. In this case, the ratio value is \> 1. This is show overdispersion.

```{r eval=F}
# See if there is an overdispersion
model$deviance/model$df.residual
# solve problem using quasibinomial

model <- glm(y ~ genus * treatment.group2, data = germination, family = quasibinomial) 

# Update model by removing interaction term
model2 <- update(model, ~ .-genus :treatment.group2) 

# See if models are different
anova(model, model2, test = "Chisq") 
```

## Model Summary

Let's see the summary

Now we use quasi binomial. We do not need to see the residual deviance anymore. The two models are significantly different. We cannot remove the interaction term.

```{r eval=F}
summary(model)

anova(model)
```

The output of `anova(model)` shows that the explanatory variables of species and treatment are a significant terms in the model. The interaction term is also significant. This mean that the effect of storage treatments depends on species.

One of our options would be analyze the effect of storage separately by species.

## Exercise

**Submission:** A PDF file containing your answers. If you are asked to use R, your answers must contain **R codes** that you wrote and **the outputs** you get. Your codes and outputs can be text and/or screen capture.

**Full score:** 10 Submission:

**Data for this exercise:** germination_data.csv (You already have this data set.)

### Questions/directions

**1.** In R, create a data frame of a subset data by choosing one of the species and show the data frame.

**2.** In R, run a GLM to test whether the germination of the chosen species is affected by the treatment. You may shoose to use all treatments or treatment groups as your explanatory variable.

**3.** In R, check for violation of model assumption and provide the solution if the assumption is violated.

**4.** In R, plot a graph to show the mean germination of each treatment.

**5.** Write in words, the interpretation of the model explaining the effects of the types of storage on seeds germination of the selected species? According to the results, will we be able to store the species in dry and cold conditions?

### Grading rubric

| Question | Score = 2                                     | Score = 1                                         | Score = 0                                           |
|-------------------|------------------|------------------|------------------|
| 1        | Correct R codes and show the data frame       | Incorrect R codes or incorrect data frame         | R code and data frame are not included or incorrect |
| 2        | Correct R code and show the model             | Partially correct R code and show the model       | R code and output are not included or incorrect     |
| 3        | Correct R code and show the model             | Partially correct R code and show the model       | R code and output are not included or incorrect     |
| 4        | Run the code and show the graph               | Run the code but R code or the graph is not shown | R code and output are not included or incorrect     |
| 5        | Correct interpretation and answer 2 questions | Partially correct                                 | Incorrect interpretation and/or do not answer       |

**End of Lab 11**
