---
output:
  html_document: default
  pdf_document: default
---
# Multivariate Analysis

**Duration:** 4-hour lecture

## Learning Outcomes

Students should be able to:

1.    Describe common terminology in multivariate analysis
2.    Distinguish between dependence and interdependence techniques
3.   Describe how Principal Component Analysis (PCA) works and interpret PCA outputs including eigenvalues, eigenvectors, and biplots
4.   Explain the purpose and logic of Linear Discriminant Analysis (LDA)
5.   Distinguish LDA from PCA and explain when each is appropriate
6.   List and check the assumptions of LDA

## Introduction

In ecology and biology, researchers often collect measurements on many variables simultaneously — such as morphological traits, environmental gradients, or species abundances. Analysing each variable separately can miss important patterns and relationships that exist across variables together. **Multivariate analysis** refers to a family of statistical techniques designed to analyse data sets in which multiple variables are measured on each observation simultaneously [@gotelli2013].

Multivariate methods allow researchers to:

-   Detect structure and patterns hidden in complex data
-   Identify natural groupings or categories among observations
-   Reduce the dimensionality of data while retaining most of the information
-   Generate hypotheses for further study

A classic example is the Iris dataset (@Anderson1935, published by @Fisher1936), which contains four morphological measurements — petal width, petal length, sepal width, and sepal length — for 150 specimens across three *Iris* species. No single variable cleanly separates the species, but multivariate analysis can reveal the underlying structure effectively.

## Two Main Categories of Multivariate Techniques

Multivariate techniques are broadly classified into two categories based on how variables are treated (Table \@ref(tab:mv-categories)).

**Interdependence techniques** treat all variables simultaneously, with no distinction between dependent and independent variables. The goal is to uncover structure within the data. Examples include:

-   Principal Component Analysis (PCA)
-   Cluster analysis
-   Multidimensional scaling
-   Correspondence analysis

**Dependence techniques** assume a distinction between dependent (response) and independent (explanatory) variables. The goal is to explain or predict the dependent variable(s) from the independent variable(s). Examples include:

-   Multiple Analysis of Variance (MANOVA)
-   Multivariate discriminant analysis
-   Canonical correlation analysis
-   Structural equation modelling

In this lesson, we present one example each of the **interdependence** and **dependence techniques**. The two examples covered in this lesson are considered an ordination technique.

```{r mv-categories, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(knitr)
library(kableExtra)

mv_table <- data.frame(
  Feature = c("Variable roles", "Primary goal", "Examples"),
  Interdependence = c("No distinction", 
                      "Uncover structure among all variables", 
                      "PCA, Cluster analysis, Multidimensional scaling"),
  Dependence = c("Dependent vs. independent", 
                 "Explain or predict y from x", 
                 "MANOVA, Discriminant analysis, Canonical correlation")
  )

mv_table %>%
  kbl(align = "l",
      col.names = c("Feature", "Interdependence", "Dependence"),
      caption = "Comparison of interdependence and dependence multivariate techniques") %>%
  kable_styling(bootstrap_options = "striped", full_width = TRUE) %>%
  column_spec(1, bold = TRUE)
```

## Ordination and Data Reduction

**Ordination** is the process of ordering or arranging sampling units along axes that represent major gradients of variation in the data [@gotelli2013]. Rather than working in a high-dimensional space with many variables, ordination reduces the data to two or three summary axes that capture the most important patterns.

The key benefits of ordination are:

-   **Noise reduction** — minor, random variation is filtered into lower-priority axes
-   **Redundancy reduction** — correlated variables are combined into composite axes
-   **Visualization** — complex multivariate patterns can be plotted in two dimensions

Formally, ordination summarises a dataset containing $n$ variables into a smaller number $p$ of composite (synthetic) variables, where $p < n$. Each composite variable is a linear combination of the original variables.

## Interdependence Category of Multivariate Analysis: Principal Component Analysis

**Principal Component Analysis (PCA)** is one of the most widely used exploratory multivariate techniques. It was introduced by Karl Pearson in 1901 [@Pearson1901] in his paper *"On Lines and Planes of Closest Fit to Systems of Points in Space"* and further developed by Harold Hotelling [@Hotelling1933].

The core idea of PCA is to find a small number of **linear combinations** of the original variables — called **principal components (PCs)** — that capture as much of the variation in the data as possible. Mathematically, if the original data have variables $X_1, X_2, \ldots, X_n$, PCA finds new composite variables:

$$C_1 = a_{11}X_1 + a_{12}X_2 + \cdots + a_{1n}X_n$$
$$C_2 = a_{21}X_1 + a_{22}X_2 + \cdots + a_{2n}X_n$$
$$\vdots$$

such that $C_1$ explains the maximum possible variance in the data, $C_2$ explains the maximum remaining variance orthogonal (perpendicular and independent) to $C_1$, and so on.

The coefficients $\{a_{11}, a_{12}, \ldots, a_{1n}\}$ are called **eigenvectors** (or loadings) of the covariance or correlation matrix, and the variance explained by each component is its associated **eigenvalue**.

### When to Use PCA

PCA is appropriate when:

-   You want to reduce many variables to a smaller number of composite variables while retaining most information
-   Variables have approximately linear relationships with one another
-   Sample size is adequate: Tabachnick and Fidell (1989) suggested a minimum of 5 observations per variable; Tabachnick and Fidell (1996) recommended at least 100 observations in total. When the correlation structure among variables is weak, a larger sample is needed.

## Key Mathematical Concepts

### Variance-Covariance Matrix

The **variance-covariance matrix** (often called simply the covariance matrix) summarises how each variable varies on its own and co-varies with every other variable. For variables $A$, $B$, and $C$, the matrix has the form shown in Table \@ref(tab:cov-matrix).

The **variance** of variable $A$ is:
$$\text{Var}(A) = \frac{\sum (A_i - \bar{A})^2}{N}$$

The **covariance** between variables $A$ and $B$ is:
$$\text{Cov}(A, B) = \frac{\sum (A_i - \bar{A})(B_i - \bar{B})}{N}$$

A positive covariance means the two variables tend to increase together; a negative covariance means one increases as the other decreases.

```{r cov-matrix, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
cov_df <- data.frame(
  ` ` = c("A", "B", "C"),
  A = c("Var(A)", "Cov(A,B)", "Cov(A,C)"),
  B = c("Cov(A,B)", "Var(B)", "Cov(B,C)"),
  C = c("Cov(A,C)", "Cov(B,C)", "Var(C)")
)

cov_df %>%
  kbl(align = "c",
      col.names = c("", "A", "B", "C"),
      caption = "Structure of a variance-covariance matrix for three variables A, B, and C") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE)
```

### Correlation Matrix

The **correlation matrix** standardises the covariances so all values lie between −1 and +1. The correlation between $A$ and $B$ is:

$$\text{corr}(A, B) = \frac{\text{Cov}(A, B)}{sd(A) \cdot sd(B)}$$

where $sd(A)$ and $sd(B)$ are the standard deviations of $A$ and $B$.

### Covariance Matrix vs. Correlation Matrix in PCA

The choice of which matrix to use for PCA has important implications (Table \@ref(tab:matrix-choice)):

-   **Covariance matrix**: use when all variables are measured in the **same units** and you want variables with larger variance to have more influence on the analysis.
-   **Correlation matrix**: use when variables are measured in **different units**, or when you want all variables to have equal influence. Using the correlation matrix is equivalent to standardising all variables to mean = 0 and SD = 1 before the analysis.

```{r matrix-choice, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
matrix_df <- data.frame(
  Feature = c("Variable units", "Effect of scale", "Variable influence", "Mean eigenvalue"),
  Covariance = c("Must be same units", 
                 "Variables with large variance dominate", 
                 "Unequal — proportional to variance",
                 "Not equal to 1.0"),
  Correlation = c("Can be different units", 
                  "All variables standardised first", 
                  "Equal — all variables contribute similarly",
                  "Equal to 1.0")
)

matrix_df %>%
  kbl(align = "l",
      col.names = c("Feature", "Covariance Matrix", "Correlation Matrix"),
      caption = "Comparison of using covariance versus correlation matrix in PCA") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  column_spec(1, bold = TRUE)
```

### Orthogonality

A key property of principal components is that they are **orthogonal** — they are perpendicular to one another in multivariate space. This means each principal component captures a completely independent (uncorrelated) dimension of variation. The first PC captures the direction of maximum variance, the second captures the next greatest variance in a direction perpendicular to the first, and so on (Figure \@ref(fig:pca-concept)).

```{r pca-concept, echo=FALSE, fig.cap="Conceptual diagram of the first two principal components (C1 and C2) fitted to a two-dimensional data cloud. C1 runs along the axis of greatest variation; C2 is perpendicular to C1.", fig.height=4.5, fig.width=5, message=FALSE, warning=FALSE}

set.seed(42)
n <- 60
x1 <- rnorm(n, 5, 0.5)
x2 <- 0.7 * x1 + rnorm(n, 0, 0.3)

par(mar = c(4, 4, 2, 2))
plot(x1, x2, pch = 1, col = "steelblue", cex = 1.2,
     xlab = "Variable X1", ylab = "Variable X2",
     xlim = c(3.5, 6.5), ylim = c(2.0, 5.5),
     main = "")

# Draw PC1 direction (red arrow)
arrows(4.6, 2.8, 6.0, 4.9, col = "red", lwd = 2.5, length = 0.15)
text(6.1, 4.8, expression(italic(C[1])), col = "red", cex = 1.1, font = 2)

# Draw PC2 direction (blue arrow)
arrows(5.5, 4.5, 4.2, 4.1, col = "steelblue4", lwd = 2.5, length = 0.15)
text(4.0, 4.3, expression(italic(C[2])), col = "steelblue4", cex = 1.1, font = 2)
```

## Steps in Performing PCA 

PCA proceeds as follows [@jaadi2025pca]:

1.  **Standardize the continuous variables** — subtract the mean and divide by the standard deviation (if using the correlation matrix).

2.  **Compute the covariance or correlation matrix** — this captures all pairwise relationships between variables.

3.  **Compute eigenvectors and eigenvalues** of the covariance/correlation matrix — eigenvectors define the directions of the new axes (principal components); eigenvalues indicate the amount of variance explained by each axis.

4.  **Create a feature vector** — decide how many principal components to retain based on their eigenvalues and variance explained.

5.  **Recast the data** along the principal component axes — project the original data onto the new axes to obtain PC scores for each observation.

## How Many Principal Components to Retain?

### The Scree Plot

A **scree plot** displays eigenvalues in decreasing order against component number. The analyst looks for an "elbow" — a sharp bend where eigenvalues begin to level off. Components before the elbow are retained; those in the flat tail are discarded as they explain relatively little additional variance (Figure \@ref(fig:screeplot-concept)).

### Kaiser's Rule

**Kaiser's rule** states: retain all principal components with eigenvalues **greater than 1.0** (when using the correlation matrix). An eigenvalue of 1.0 means the component explains as much variance as a single standardised variable — components below this threshold are considered to explain less than their "fair share."

### Cumulative Variance Explained

A complementary criterion is to retain enough components to account for a **cumulative variance of 50–70%** of the total. In ecological studies with complex data, even 50% cumulative variance may be acceptable.

```{r screeplot-concept, echo=FALSE, fig.cap="Example scree plot showing eigenvalues for 12 components. The elbow at component 2 (dashed line) suggests retaining the first two principal components. The horizontal dotted line marks eigenvalue = 1 (Kaiser's rule).", fig.height=4, fig.width=6, message=FALSE, warning=FALSE}

eigenvalues <- c(6.2, 1.3, 0.75, 0.6, 0.5, 0.45, 0.38, 0.3, 0.25, 0.22, 0.18, 0.15)
components <- 1:12

par(mar = c(4, 4, 2, 2))
plot(components, eigenvalues, type = "b", pch = 19, col = "darkblue",
     xlab = "Component Number", ylab = "Eigenvalue",
     ylim = c(0, 7), xaxt = "n")
axis(1, at = 1:12)
abline(v = 2, lty = 2, col = "red", lwd = 1.5)
abline(h = 1, lty = 3, col = "gray40", lwd = 1.5)
legend("topright", legend = c("Elbow (retain up to here)", "Kaiser threshold (eigenvalue = 1)"),
       lty = c(2, 3), col = c("red", "gray40"), bty = "n", cex = 1)
```

## Example: PCA of Seed Morphology Data

To illustrate PCA in practice, we use a dataset of seed morphological measurements from three tropical tree species collected in northern Thailand [@doisuthep2022final]:

-   *Cinnamomum caudatum* (Lauraceae)
-   *Erythrina subumbrans* (Fabaceae)
-   *Garcinia cowa* (Clusiaceae)

Twenty seeds were measured per species ($n = 60$ total). The measured variables are: seed width (mm), seed length (mm), seed thickness (mm), estimated volume (mm³), seed coat thickness (mm), wet weight (g), dry weight (g), and moisture content (%) [\@ref(tab:seed-data)].

### Loading and Exploring the Data

```{r load-data, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}

library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)

# Load data
seed <- read.csv("data/seed3spp.csv")

# Create species label
seed$species <- paste(seed$genus, seed$epithet)
seed$species <- trimws(seed$species)

seed$species.abb <- paste0(substr(seed$genus, 1, 1),". ",seed$epithet)

seed$species <- as.factor(seed$species)
seed$species.abb <- as.factor(seed$species.abb)
# Preview
seed.head <- head(seed)[, -c(1:2,13)]
seed.for.table <- seed.head %>%
  relocate(species)
```

```{r seed-data, echo=FALSE, message=FALSE, warning=FALSE}

seed.for.table %>% 
  kbl(align = "c",
      col.names = c("Species", "Seed ID", "Width (mm)", "Length (mm)",
                    "Thickness (mm³)" ,"Volume (cubic mm)", "Seed coat thickness (mm)", "Wet weight (g)", "Dry weight (g)", "Moisture content (%)"),
      caption = "First Part of a Data Set of Seed Morphology") %>%
  kable_styling(bootstrap_options = "striped", full_width = F) %>%
  column_spec(1, bold = FALSE, italic = TRUE)
```

Then, let's explore the data set and calculate mean traits for the three species.
```{r summary-seed, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# Summary statistics by species
sum_seed <- seed %>%
  group_by(species.abb) %>%
  summarise(
    n = n(),
    `Width (mm)` = round(mean(width_mm), 1),
    `Length (mm)` = round(mean(long_mm), 1),
    `Thickness (mm)` = round(mean(thickness_mm), 1),
    `Volume (mm³)` = round(mean(volume_mm3), 1),
    `Coat thick (mm)` = round(mean(coat_thick_mm), 1),
    `Wet wt (g)` = round(mean(wet_weight_g), 1),
    `Dry wt (g)` = round(mean(dry_weight_g), 1),
    `Moisture (%)` = round(mean(moisture_percent), 0),
    .groups = "drop"
  )

sum_seed %>%
  kbl(align = "c",
      caption = "Mean seed morphological traits for three tropical tree species (n = 20 seeds per species)") %>%
  kable_styling(bootstrap_options = "striped", full_width = TRUE) %>%
  column_spec(1, bold = FALSE, italic = TRUE)
```

### Visualizing the Raw Data

Before running PCA it is always good practice to visualize variation in individual traits across species (Figure \@ref(fig:boxplots-seed)). Here are examples of four seed traits

```{r boxplots-seed, echo=FALSE, fig.cap="Boxplots of selected seed traits (length, width, volume, and dry weight) by species. Note that Garcinia cowa seeds are substantially larger than the other two species on all traits.", fig.height=6, fig.width=9, message=FALSE, warning=FALSE}

if(!require("ggthemes", character.only = TRUE)) {
  install.packages("ggthemes", dependencies = TRUE)
}
library(ggthemes)
library(tidyr)

seed_long <- seed %>%
  select(species.abb, long_mm, width_mm, volume_mm3, dry_weight_g) %>%
  pivot_longer(cols = -species.abb, names_to = "trait", values_to = "value") %>%
  mutate(trait = recode(trait,
    long_mm = "Length (mm)",
    width_mm = "Width (mm)",
    volume_mm3 = "Volume (mm³)",
    dry_weight_g = "Dry weight (g)"
  ))

ggplot(seed_long, aes(x = species.abb, y = value, fill = species.abb)) +
  geom_boxplot(alpha = 0.6, width = 0.5) +
  geom_jitter(width = 0.1, size = 1.5, alpha = 0.6) +
  facet_wrap(~ trait, scales = "free_y") +
  labs(x = "Species", y = "Value") +
  theme_bw() +
  theme(
    legend.position = "none",
    legend.title = element_blank(),
    #axis.text.x = element_blank(),
    #axis.ticks.x = element_blank(),
    strip.text = element_text(size = 14),
    axis.text.x = element_text(size = 14, face = "italic", 
                               angle = 45, hjust = 1, vjust = 1),
    
    axis.text.y = element_text(size = 14),
    axis.title = element_text(size = 14))
    #legend.text = element_text(face = "italic", size = 10)
```

### Running PCA

We select the eight continuous morphological variables for PCA. Because the variables are measured in different units and span very different scales (e.g., mm vs. mm³ vs. g), we use the **correlation matrix**, which standardises all variables to mean = 0 and SD = 1 before analysis. In R, we use `prcomp` to run PCA.

```{r run-pca, echo=FALSE, message=FALSE, warning=FALSE, include = FALSE}
# Select numeric trait columns
pca_vars <- seed %>%
  select(width_mm, long_mm, thickness_mm, volume_mm3,
         coat_thick_mm, wet_weight_g, dry_weight_g, moisture_percent)

# Run PCA using the correlation matrix (scale = TRUE)
pca_result <- prcomp(pca_vars, center = TRUE, scale. = TRUE)

# View summary
summary(pca_result)
```

### Eigenvalues and Variance Explained

We can use the scree plot of eigenvalues to decide how many PC should be retained. Based on Kaiser's rule (eigenvalue ≥ 1), **`r sum(eigenvalues >= 1)` principal components** should be retained. These components together explain **`r cum_var[sum(eigenvalues >= 1)]`%** of the total variation in seed morphology (\@ref(fig:eigenvalues)). 

```{r eigenvalues, echo=FALSE, fig.cap="Scree plot of eigenvalues for the eight principal components from seed morphology PCA. The elbow appears after PC2, and the dashed line shows the Kaiser threshold (eigenvalue = 1).", fig.height=4, fig.width=6, message=FALSE, warning=FALSE}

eigenvalues <- pca_result$sdev^2
pct_var <- round(eigenvalues / sum(eigenvalues) * 100, 1)
cum_var <- cumsum(pct_var)

# Scree plot
par(mar = c(4, 4, 2, 2))
plot(1:length(eigenvalues), eigenvalues, type = "b", pch = 19,
     col = "steelblue", xlab = "Principal Component", ylab = "Eigenvalue",
     xaxt = "n")
axis(1, at = 1:length(eigenvalues))
abline(h = 1, lty = 2, col = "red", lwd = 1.5)
legend("topright", legend = "Kaiser threshold (eigenvalue = 1)",
       lty = 2, col = "red", bty = "n", cex =1)
```

The variance explained by each PC is shown in Table (\@ref(tab:varaince-table))

```{r variance-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
var_table <- data.frame(
  PC = paste0("PC", 1:length(eigenvalues)),
  Eigenvalue = round(eigenvalues, 3),
  `Variance Explained (%)` = pct_var,
  `Cumulative (%)` = cum_var
)

var_table %>%
  kbl(align = "c",
      col.names = c("Component", "Eigenvalue", "Variance Explained (%)", "Cumulative (%)"),
      caption = "Eigenvalues and variance explained by each principal component from the seed morphology PCA") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  row_spec(which(var_table$Eigenvalue >= 1), background = "#e8f4f8") %>%
  column_spec(1, bold = TRUE)
```

### Loadings (Eigenvectors)

**Loadings** show the correlation between each original variable and each principal component. Large positive or negative loadings (typically |loading| > 0.5) indicate that a variable is strongly associated with that component and helps to interpret what biological gradient the component represents.

```{r loadings-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
loadings_df <- as.data.frame(round(pca_result$rotation[, 1:3], 3))
loadings_df$Variable <- rownames(loadings_df)

loadings_df <- loadings_df %>%
  mutate(Variable = recode(Variable,
    width_mm = "Width (mm)",
    long_mm = "Length (mm)",
    thickness_mm = "Thickness (mm)",
    volume_mm3 = "Volume (mm³)",
    coat_thick_mm = "Coat thickness (mm)",
    wet_weight_g = "Wet weight (g)",
    dry_weight_g = "Dry weight (g)",
    moisture_percent = "Moisture content (%)"
  )) %>%
  select(Variable, PC1, PC2, PC3)

rownames(loadings_df) <- NULL
loadings_df %>%
  kbl(align = "c",
      col.names = c("Variable", "PC1", "PC2", "PC3"),
      caption = "PCA loadings for the first three principal components. Values close to ±1 indicate strong association with the component.") %>%
  kable_styling(bootstrap_options = "striped", full_width = FALSE) %>%
  column_spec(1, bold = TRUE)
```

**Interpretation of loadings**

-   **PC1** has large positive loadings for width, length, thickness, volume, wet weight, and dry weight. This component represents overall **seed size** — seeds scoring high on PC1 are large in all dimensions.
-   **PC2** has a large positive loading for moisture content and a contrasting loading for dry weight. This component represents a **water content gradient**, distinguishing seeds that retain relatively more or less moisture relative to their dry mass.

### PCA Biplot

A **biplot** displays the PC scores of observations (as points) and the loadings of variables (as arrows). Arrows pointing in the same direction indicate positively correlated variables; arrows pointing in opposite directions indicate negatively correlated variables; arrows that are roughly perpendicular are uncorrelated. The length of an arrow is proportional to how well the variable is represented in the reduced PCA space (Figure \@ref(fig:biplot)).

```{r biplot, echo=FALSE, fig.cap="PCA biplot of seed morphological data for three tropical tree species. Points represent individual seeds coloured by species; arrows show variable loadings on PC1 and PC2. The percentage of variance explained by each axis is shown in parentheses.", fig.height=6, fig.width=7, message=FALSE, warning=FALSE}

if(!require("ggrepel", character.only = TRUE)) {
  install.packages("ggrepel", dependencies = TRUE)
}
library(ggrepel)

# PC scores
scores <- as.data.frame(pca_result$x[, 1:2])
scores$species <- seed$species

# Loadings scaled for biplot
load_scale <- 3
loadings_plot <- as.data.frame(pca_result$rotation[, 1:2] * load_scale)
loadings_plot$variable <- c("Width", "Length", "Thickness", "Volume",
                             "Coat thick.", "Wet weight.", "Dry weight.", "Moisture")

# Variance labels
pct1 <- round(pct_var[1], 1)
pct2 <- round(pct_var[2], 1)

ggplot() +
  # Points
  geom_point(data = scores, aes(x = PC1, y = PC2, colour = species, shape = species),
             size = 2.5, alpha = 0.8) +
  # Arrows
  geom_segment(data = loadings_plot,
               aes(x = 0, y = 0, xend = PC1, yend = PC2),
               arrow = arrow(length = unit(0.2, "cm")),
               colour = "black", linewidth = 0.7) +
  # Variable labels
  geom_text_repel(data = loadings_plot,
                  aes(x = PC1, y = PC2, label = variable),
                  size = 3.5, colour = "black", fontface = "bold") +
  # Reference lines
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  labs(
    x = paste0("PC1 (", pct1, "% variance explained)"),
    y = paste0("PC2 (", pct2, "% variance explained)"),
    colour = "Species", shape = "Species"
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.title = element_blank(),
    legend.text = element_text(face = "italic", size = 10),
    axis.title = element_text(size = 11)
  )
```

**Interpretation of the biplot:**

-   The three species are well separated along **PC1** (the size axis). *Garcinia cowa* seeds (large, heavy) score high on PC1, while *Cinnamomum caudatum* seeds (small, light) score low. *Erythrina subumbrans* is intermediate.
-   Along **PC2**, *Erythrina subumbrans* tends to score higher, suggesting its seeds have relatively higher moisture content for their dry mass compared with the other two species.
-   Length, width, thickness, volume, wet weight, and dry weight arrows all point in roughly the same direction, confirming that these size-related traits are strongly positively correlated.
-   Moisture content points in a distinct direction from dry weight, indicating it is not strongly correlated with dry mass — moisture is a separate axis of variation.

### Reporting PCA Results

> "A principal component analysis was conducted on eight seed morphological traits (width, length, thickness, volume, coat thickness, wet weight, dry weight, and moisture content) for 60 seeds from three tropical tree species, using the correlation matrix. Based on Kaiser's rule (eigenvalue ≥ 1), the first `r sum(eigenvalues >= 1)` principal components were retained, explaining `r cum_var[sum(eigenvalues >= 1)]`% of the total variance. PC1 (explaining `r pct_var[1]`% of variance) had high positive loadings for all size-related traits and is interpreted as a **seed size** axis. PC2 (explaining `r pct_var[2]`% of variance) was strongly loaded by moisture content and is interpreted as a **moisture content** axis. The biplot showed clear separation among the three species, primarily along PC1, with *Garcinia cowa* occupying the high-size end, *Cinnamomum caudatum* the low-size end, and *Erythrina subumbrans* in an intermediate position with relatively higher moisture content."

## Summary of PCA

Table \@ref(tab:pca-summary) summarises the key decision points when performing PCA.

```{r pca-summary, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
sum_df <- data.frame(
  Step = c("Matrix choice", "Retention criterion 1", "Retention criterion 2", 
           "Interpreting components", "Visualization"),
  Decision = c("Correlation matrix if variables have different units or scales",
               "Kaiser's rule: retain PCs with eigenvalue > 1",
               "Retain enough PCs for cumulative variance > 50–70%",
               "Examine loadings; high |loading| (> 0.5) indicates strong association",
               "Biplot — simultaneously shows scores and loadings")
)

sum_df %>%
  kbl(align = "l",
      col.names = c("Step", "Guidance"),
      caption = "Key decision points in PCA") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  column_spec(1, bold = TRUE)
```

## Dependence Category of Multivariate Analysis: Linear Discriminant Analysis (LDA)

From the example of the seed morphological data set (Table \@ref(tab:seed-data)), PCA results suggest that the species occupy distinct regions of multivariate space — but PCA is an **unsupervised** technique. It finds axes of maximum overall variance without any knowledge of group membership.

A next question is: *given that we know which species each seed belongs to, can we find the combination of traits that best separates the groups — and can we use that combination to classify unknown specimens?* This is the domain of **Linear Discriminant Analysis (LDA)**, also known as Multivariate Discriminant Analysis or Fisher's Discriminant Analysis.

LDA is in the **dependence** category of multivariate techniques: 

-   **The dependent variable** is species,
-   **The independent (predictor) variables** are the morphological measurements are the .

LDA is both a dimension-reduction tool — like PCA — and a classification method, making it one of the most practically useful multivariate techniques in biology and ecology.

## LDA vs. PCA: Key Differences

Although both LDA and PCA produce linear combinations of variables and reduce dimensionality, they pursue fundamentally different goals (Table \@ref(tab:lda-vs-pca)).

```{r lda-vs-pca, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
library(knitr)
library(kableExtra)

comp_df <- data.frame(
  Feature = c("Category", "Uses group labels?", "Goal",
              "Axes maximise", "Number of axes", "Typical use"),
  PCA = c("Interdependence (unsupervised)",
          "No",
          "Summarise overall variation",
          "Total variance",
          "Up to p (number of variables)",
          "Explore data structure, reduce dimensions"),
  LDA = c("Dependence (supervised)",
          "Yes",
          "Separate known groups",
          "Between-group variance relative to within-group variance",
          "Up to g − 1 (number of groups minus one)",
          "Classify observations, confirm group separation")
)

comp_df %>%
  kbl(align = "l",
      col.names = c("Feature", "PCA", "LDA"),
      caption = "Comparison of Principal Component Analysis (PCA) and Linear Discriminant Analysis (LDA)") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  column_spec(1, bold = TRUE)
```

A useful way to remember the distinction: PCA asks *"where is the most variation?"*, while LDA asks *"where are the groups most different?"*.

## The Logic of LDA

LDA finds linear combinations of predictor variables — called **discriminant functions** — that maximise the ratio of **between-group variance** to **within-group variance**. A good discriminant function spreads the group means far apart while keeping the observations within each group tightly clustered.

For $p$ predictor variables and $g$ groups, the discriminant functions are:

$$D_k = b_{k1}X_1 + b_{k2}X_2 + \cdots + b_{kp}X_p \quad k = 1, 2, \ldots, g-1$$

where $b_{k1}, b_{k2}, \ldots, b_{kp}$ are the **discriminant coefficients** (analogous to eigenvectors in PCA). Each observation receives a **discriminant score** on each function, and classification is based on which group centroid (mean discriminant score) the observation is closest to.

### Between-Group and Within-Group Matrices

LDA decomposes total multivariate variance into two parts:

-   **Within-group scatter matrix ($\mathbf{W}$)**: summarizes variation of observations around their own group mean. This is "noise" — variance that does not help separate groups.
-   **Between-group scatter matrix ($\mathbf{B}$)**: summarises how far apart the group means are from the overall mean. This is "signal" — variance that helps separate groups.

LDA finds the axis that maximizes $\mathbf{B} / \mathbf{W}$, the ratio of between-group to within-group dispersion. This is solved as an **eigenvalue problem** on the matrix $\mathbf{W}^{-1}\mathbf{B}$: the eigenvectors give the discriminant function coefficients and the eigenvalues indicate how much group separation each function achieves.

### How Many Discriminant Functions?

The maximum number of discriminant functions is $\min(p,\; g-1)$, 
where $p$ is the number of predictor variables and $g$ is the number of groups. With three species and eight morphological traits, we can extract at most $3 - 1 = 2$ discriminant functions.

### Mahalanobis Distance and Classification

To classify a new observation, LDA computes its **Mahalanobis distance** to each group centroid. The Mahalanobis distance accounts for correlations among variables and differences in variance, making it more appropriate than ordinary Euclidean distance in multivariate settings. An observation is assigned to the group whose centroid it is closest to.

## Assumptions of LDA

LDA has the following assumptions (Table \@ref(tab:lda-assumptions)):

```{r lda-assumptions, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}

if(!require("knitr", character.only = TRUE)) {
  install.packages("knitr", dependencies = TRUE)}
if(!require("kableExtra", character.only = TRUE)) {
  install.packages("kableExtra", dependencies = TRUE)}
if(!require("ggplot2", character.only = TRUE)) {
  install.packages("ggplot2", dependencies = TRUE)}
if(!require("dplyr2", character.only = TRUE)) {
  install.packages("dplyr", dependencies = TRUE)}


library(ggplot2)
library(knitr)
library(kableExtra)
library(dplyr)

ass_df <- data.frame(
  Assumption = c("Multivariate normality",
                 "Homogeneity of covariance matrices",
                 "No multicollinearity",
                 "No outliers",
                 "Adequate sample size"),
  Description = c("Each group's predictor variables follow a multivariate normal distribution",
                  "The variance-covariance matrix is the same in all groups (homoscedasticity)",
                  "Predictors should not be perfectly correlated with one another",
                  "Extreme outliers can strongly influence discriminant function coefficients",
                  "At least 5 observations per predictor variable per group is recommended"),
  `How to check` = c("Univariate normality per group (Shapiro-Wilk); Q-Q plots",
                     "Box's M test; comparison of group covariance matrices",
                     "Correlation matrix; VIF values",
                     "Mahalanobis distance; scatter plots",
                     "Count observations and variables per group")
)

ass_df %>%
  kbl(align = "l",
      col.names = c("Assumption", "Description", "How to Check"),
      caption = "Assumptions of Linear Discriminant Analysis and how to check them") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  column_spec(1, bold = TRUE)
```

When the homogeneity of covariance assumption is violated, **Quadratic Discriminant Analysis (QDA)** can be used as an alternative — it allows each group to have its own covariance matrix, at the cost of requiring a larger sample size.

## Steps in Performing LDA

1.  **Explore the data** — visualize group separation with boxplots and a PCA or pairs plot.
2.  **Check assumptions** — normality per group, homogeneity of covariance matrices, multicollinearity.
3.  **Fit the LDA model** — compute discriminant functions from the training data.
4.  **Examine discriminant function coefficients** — identify which variables contribute most to group separation.
5.  **Plot discriminant scores** — visualize group separation in discriminant space.
6.  **Evaluate classification accuracy** — construct a confusion matrix; perform cross-validation (leave-one-out) to estimate realistic classification rates.
7.  **Report results** — state the percentage variance explained by each function, classification accuracy, and which variables best discriminated the groups.

## Example: LDA of Seed Morphology Data

We continue with the seed morphology dataset used in the PCA example, comprising eight morphological measurements on 20 seeds each from three tropical tree species: *Cinnamomum caudatum*, *Erythrina subumbrans*, and *Garcinia cowa*. The research question is now explicitly supervised: 

1. **Can we classify seeds to their correct species based on their morphological measurements**, and 
2. **Which traits are most informative for classification?**

### Loading and Preparing the Data

```{r load-data, echo=FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(knitr)
library(kableExtra)
library(MASS)       # for lda()

# Load data - data is there already.

# Select trait variables and species label
seed_traits <- seed %>% 
  select(species.abb, width_mm, long_mm, thickness_mm, volume_mm3,
         coat_thick_mm, wet_weight_g, dry_weight_g, moisture_percent)

# Quick summary
table(seed_traits$species.abb)
```

### Step 1: Explore the Data

Before fitting LDA, we visualize how well individual traits separate the three species using boxplots (Figure \@ref(fig:explore-boxplots)) and a pairs plot of selected traits (Figure \@ref(fig:pairs-plot)).

```{r explore-boxplots, echo=FALSE, fig.cap="Boxplots of all eight seed morphological traits by species. Traits with little overlap between boxes are better candidates for group discrimination.", fig.height=7, fig.width=10, message=FALSE, warning=FALSE}
library(tidyr)

if(!require("ggthemes")) install.packages("ggthemes", dependencies = TRUE)
library(ggthemes)

trait_labels <- c(
  width_mm      = "Width (mm)",
  long_mm       = "Length (mm)",
  thickness_mm  = "Thickness (mm)",
  volume_mm3    = "Volume (mm³)",
  coat_thick_mm = "Coat thickness (mm)",
  wet_weight_g  = "Wet weight (g)",
  dry_weight_g  = "Dry weight (g)",
  moisture_percent = "Moisture (%)"
)

seed_long <- seed_traits %>%
  pivot_longer(cols = -species, names_to = "trait", values_to = "value") %>%
  mutate(trait = recode(trait, !!!trait_labels))

ggplot(seed_long, aes(x = species, y = value, fill = species)) +
  geom_boxplot(alpha = 0.6, outlier.shape = 21) +
  geom_jitter(width = 0.12, size = 1.3, alpha = 0.5) +
  facet_wrap(~ trait, scales = "free_y", ncol = 4) +
  labs(x = NULL, y = "Value") +
  theme_bw() +
  theme(
    legend.position  = "bottom",
    legend.title     = element_blank(),
    legend.text      = element_text(face = "italic", size = 10),
    axis.text.x      = element_blank(),
    axis.ticks.x     = element_blank(),
    strip.text       = element_text(size = 10)
  )
```

```{r pairs-plot, echo=FALSE, fig.cap="Pairs plot of four selected seed traits coloured by species. Pairs of traits that show good group separation in bivariate scatter are strong discriminators.", fig.height=6, fig.width=7, message=FALSE, warning=FALSE}
species_cols <- c(
  "Cinnamomum caudatum"  = "#E69F00",
  "Erythrina subumbrans" = "#56B4E9",
  "Garcinia cowa"        = "#009E73"
)

pairs(
  seed_traits[, c("long_mm", "width_mm", "dry_weight_g", "moisture_percent")],
  col   = species_cols[as.character(seed_traits$species)],
  pch   = 19,
  cex   = 0.9,
  labels = c("Length\n(mm)", "Width\n(mm)", "Dry wt\n(g)", "Moisture\n(%)"),
  gap   = 0.3
)
legend("topright",
       legend = names(species_cols),
       col    = species_cols,
       pch    = 19,
       bty    = "n",
       cex    = 0.85,
       text.font = 3)
```

### Step 2: Check Assumptions

#### Normality

We test univariate normality of each trait within each species using the Shapiro-Wilk test. A non-significant result (p > 0.05) is consistent with normality.

```{r normality-test, echo=TRUE, message=FALSE, warning=FALSE, results='asis'}
traits <- c("width_mm","long_mm","thickness_mm","volume_mm3",
            "coat_thick_mm","wet_weight_g","dry_weight_g","moisture_percent")

norm_results <- lapply(unique(seed_traits$species), function(sp) {
  sub <- seed_traits %>% filter(species == sp)
  lapply(traits, function(tr) {
    sw <- shapiro.test(sub[[tr]])
    data.frame(Species = sp, Trait = tr,
               W = round(sw$statistic, 3),
               p_value = round(sw$p.value, 3))
  }) %>% bind_rows()
}) %>% bind_rows()

norm_results %>%
  mutate(Result = ifelse(p_value >= 0.05, "Normal", "Non-normal")) %>%
  kbl(align = "c",
      col.names = c("Species", "Trait", "W statistic", "p-value", "Decision"),
      caption = "Shapiro-Wilk normality test for each trait within each species. p > 0.05 is consistent with normality.") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  column_spec(1, italic = TRUE) %>%
  row_spec(which(norm_results$p_value < 0.05), background = "#fff3cd")
```

#### Homogeneity of Covariance Matrices

Box's M test evaluates whether the covariance matrices are equal across groups. The test is sensitive to departures from normality; a significant result warrants caution but does not necessarily invalidate LDA, especially with balanced group sizes.

```{r boxm-test, echo=TRUE, message=FALSE, warning=FALSE}
if(!require("biotools")) install.packages("biotools", dependencies = TRUE)
library(biotools)

boxm_result <- boxM(
  seed_traits[, traits],
  seed_traits$species
)
print(boxm_result)
```

#### Multicollinearity

We inspect the correlation matrix of the trait variables. Correlations above 0.90 indicate problematic multicollinearity.

```{r collinearity, echo=FALSE, fig.cap="Correlation matrix of seed morphological traits. Darker blue = strong positive correlation; white = no correlation. Highly correlated pairs (e.g., volume and dry weight) carry redundant information.", fig.height=5, fig.width=6, message=FALSE, warning=FALSE}
if(!require("corrplot")) install.packages("corrplot", dependencies = TRUE)
library(corrplot)

cor_mat <- cor(seed_traits[, traits])
rownames(cor_mat) <- colnames(cor_mat) <- c("Width","Length","Thickness",
                                             "Volume","Coat thick.","Wet wt.",
                                             "Dry wt.","Moisture")
corrplot(cor_mat, method = "color", type = "upper",
         tl.col = "black", tl.srt = 45, tl.cex = 0.85,
         addCoef.col = "black", number.cex = 0.70,
         col = colorRampPalette(c("white","steelblue","darkblue"))(200))
```

### Step 3: Fit the LDA Model

```{r fit-lda, echo=TRUE, message=FALSE, warning=FALSE}
lda_model <- lda(species ~ width_mm + long_mm + thickness_mm + volume_mm3 +
                   coat_thick_mm + wet_weight_g + dry_weight_g + moisture_percent,
                 data = seed_traits,
                 prior = rep(1/3, 3))   # equal prior probabilities

print(lda_model)
```

### Step 4: Discriminant Function Coefficients

The **standardised discriminant function coefficients** show the relative contribution of each variable to group separation, after accounting for differences in scale (Table \@ref(tab:std-coeffs)). Variables with large absolute standardised coefficients are the most important discriminators.

```{r std-coeffs, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# Raw coefficients
raw_coef <- as.data.frame(lda_model$scaling)

# Standardise by multiplying by within-group SD of each variable
pooled_sd <- seed_traits %>%
  group_by(species) %>%
  summarise(across(all_of(traits), sd), .groups = "drop") %>%
  summarise(across(all_of(traits), mean)) %>%
  unlist()

std_coef <- sweep(lda_model$scaling, 1, pooled_sd, `*`)
std_coef_df <- as.data.frame(round(std_coef, 3))
std_coef_df$Variable <- c("Width (mm)", "Length (mm)", "Thickness (mm)",
                           "Volume (mm³)", "Coat thickness (mm)",
                           "Wet weight (g)", "Dry weight (g)", "Moisture (%)")
std_coef_df <- std_coef_df %>% select(Variable, LD1, LD2)

std_coef_df %>%
  kbl(align = "c",
      col.names = c("Variable", "LD1", "LD2"),
      caption = "Standardised discriminant function coefficients for LD1 and LD2. Larger absolute values indicate greater contribution to group separation.") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE)
```

### Step 5: Proportion of Variance Explained

Each discriminant function's **eigenvalue** indicates how much between-group separation it achieves relative to within-group scatter. The proportion of variance attributed to each function shows its relative importance (Table \@ref(tab:eigenvalue-table)).

```{r eigenvalue-table, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
eig_vals <- lda_model$svd^2
pct_eig  <- round(eig_vals / sum(eig_vals) * 100, 1)

eig_df <- data.frame(
  Function  = c("LD1", "LD2"),
  Eigenvalue = round(eig_vals, 3),
  `Variance explained (%)` = pct_eig,
  `Cumulative (%)` = cumsum(pct_eig)
)

eig_df %>%
  kbl(align = "c",
      col.names = c("Function", "Eigenvalue", "Variance Explained (%)", "Cumulative (%)"),
      caption = "Eigenvalues and variance explained by each discriminant function") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE)
```

LD1 explains **`r pct_eig[1]`%** of the between-group variance and LD2 explains the remaining **`r pct_eig[2]`%**.

### Step 6: Plot Discriminant Scores

Figure \@ref(fig:lda-scores) shows the discriminant scores of all 60 seeds on LD1 and LD2. Well-separated clusters indicate that the discriminant functions have successfully differentiated the species.

```{r lda-scores, echo=FALSE, fig.cap="Scatter plot of discriminant scores on LD1 and LD2 for 60 seeds from three tropical tree species. Points coloured by species; ellipses show 95% confidence regions for each group centroid.", fig.height=5.5, fig.width=7, message=FALSE, warning=FALSE}
# Predict discriminant scores
lda_pred   <- predict(lda_model)
scores_df  <- data.frame(lda_pred$x, species = seed_traits$species)

# Group centroids
centroids <- scores_df %>%
  group_by(species) %>%
  summarise(LD1 = mean(LD1), LD2 = mean(LD2), .groups = "drop")

ggplot(scores_df, aes(x = LD1, y = LD2, colour = species, fill = species)) +
  stat_ellipse(geom = "polygon", alpha = 0.10, level = 0.95, linewidth = 0.8) +
  geom_point(size = 2.5, alpha = 0.8) +
  geom_point(data = centroids, aes(x = LD1, y = LD2),
             shape = 8, size = 5, stroke = 1.5, colour = "black",
             show.legend = FALSE) +
  geom_text(data = centroids, aes(x = LD1, y = LD2 + 0.35, label = species),
            fontface = "italic", size = 3.2, colour = "black") +
  geom_hline(yintercept = 0, linetype = "dashed", colour = "grey60") +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey60") +
  labs(
    x = paste0("LD1 (", pct_eig[1], "% of between-group variance)"),
    y = paste0("LD2 (", pct_eig[2], "% of between-group variance)")
  ) +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.title    = element_blank(),
    legend.text     = element_text(face = "italic", size = 10),
    axis.title      = element_text(size = 11)
  )
```

We can also display the distribution of discriminant scores on LD1 alone — the dominant axis — using density plots (Figure \@ref(fig:lda-density)).

```{r lda-density, echo=FALSE, fig.cap="Density plots of discriminant scores on LD1 by species. Non-overlapping distributions indicate good classification on this axis.", fig.height=4, fig.width=7, message=FALSE, warning=FALSE}
ggplot(scores_df, aes(x = LD1, fill = species, colour = species)) +
  geom_density(alpha = 0.35, linewidth = 0.8) +
  geom_rug(alpha = 0.6) +
  labs(x = paste0("LD1 (", pct_eig[1], "% of between-group variance)"),
       y = "Density") +
  theme_bw() +
  theme(
    legend.position = "bottom",
    legend.title    = element_blank(),
    legend.text     = element_text(face = "italic", size = 10)
  )
```

### Step 7: Evaluate Classification Accuracy

#### Confusion Matrix (Resubstitution)

The **confusion matrix** cross-tabulates predicted group membership against true group membership. The diagonal entries are correct classifications; off-diagonal entries are misclassifications (Table \@ref(tab:confusion-matrix)).

```{r confusion-matrix, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
pred_class <- lda_pred$class
conf_mat   <- table(True = seed_traits$species, Predicted = pred_class)
overall_acc <- round(sum(diag(conf_mat)) / sum(conf_mat) * 100, 1)

conf_df <- as.data.frame.matrix(conf_mat)
conf_df$`Correct (%)` <- round(diag(conf_mat) / rowSums(conf_mat) * 100, 1)
conf_df <- cbind(Species = rownames(conf_df), conf_df)
rownames(conf_df) <- NULL

conf_df %>%
  kbl(align = "c",
      caption = paste0("Confusion matrix from LDA resubstitution. Overall accuracy = ",
                       overall_acc, "%.")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE, italic = TRUE) %>%
  add_header_above(c(" " = 1, "Predicted species" = 3, " " = 1))
```

#### Cross-Validation (Leave-One-Out)

Resubstitution accuracy is **optimistic** because the model is evaluated on the same data used to fit it. A more honest estimate of classification performance comes from **leave-one-out cross-validation (LOOCV)**: each observation is removed, the LDA model is refitted on the remaining data, and the removed observation is then classified. This process is repeated for every observation.

```{r loocv, echo=TRUE, message=FALSE, warning=FALSE}
lda_cv <- lda(species ~ width_mm + long_mm + thickness_mm + volume_mm3 +
                coat_thick_mm + wet_weight_g + dry_weight_g + moisture_percent,
              data = seed_traits,
              prior = rep(1/3, 3),
              CV = TRUE)   # CV = TRUE triggers LOOCV
```

```{r loocv-matrix, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
conf_cv     <- table(True = seed_traits$species, Predicted = lda_cv$class)
cv_acc      <- round(sum(diag(conf_cv)) / sum(conf_cv) * 100, 1)

conf_cv_df <- as.data.frame.matrix(conf_cv)
conf_cv_df$`Correct (%)` <- round(diag(conf_cv) / rowSums(conf_cv) * 100, 1)
conf_cv_df <- cbind(Species = rownames(conf_cv_df), conf_cv_df)
rownames(conf_cv_df) <- NULL

conf_cv_df %>%
  kbl(align = "c",
      caption = paste0("Cross-validated (LOOCV) confusion matrix. Overall cross-validated accuracy = ",
                       cv_acc, "%.")) %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE, italic = TRUE) %>%
  add_header_above(c(" " = 1, "Predicted species" = 3, " " = 1))
```

The cross-validated overall accuracy is **`r cv_acc`%**, compared with the resubstitution accuracy of **`r overall_acc`%**. The difference between the two estimates indicates the degree to which the original model was overfit to the training data.

### Visualising Posterior Probabilities

For each observation, LDA computes the **posterior probability** of belonging to each group. Figure \@ref(fig:posterior-probs) displays these probabilities: stacked bars that reach confidently to one colour indicate high certainty; mixed colours indicate borderline cases.

```{r posterior-probs, echo=FALSE, fig.cap="Posterior probabilities of group membership for each seed, sorted by species. Stacked bars show the probability assigned to each of the three species. Seeds with mixed-colour bars are classified with lower certainty.", fig.height=5, fig.width=9, message=FALSE, warning=FALSE}
post_df <- as.data.frame(lda_pred$posterior)
post_df$true_species <- seed_traits$species
post_df$id           <- seq_len(nrow(post_df))

# Sort within each species by posterior probability of true class
post_df <- post_df %>%
  group_by(true_species) %>%
  arrange(true_species) %>%
  mutate(id = row_number()) %>%
  ungroup()

post_long <- post_df %>%
  pivot_longer(cols = -c(true_species, id),
               names_to = "Predicted", values_to = "Probability")

ggplot(post_long, aes(x = id, y = Probability, fill = Predicted)) +
  geom_col(width = 0.9) +
  facet_wrap(~ true_species, scales = "free_x",
             labeller = label_wrap_gen(width = 20)) +
  scale_fill_manual(values = c("#E69F00","#56B4E9","#009E73")) +
  labs(x = "Seed (within species)", y = "Posterior Probability", fill = "Assigned to") +
  theme_bw() +
  theme(
    legend.position  = "bottom",
    legend.text      = element_text(face = "italic", size = 9),
    strip.text       = element_text(face = "italic", size = 10),
    axis.text.x      = element_blank(),
    axis.ticks.x     = element_blank()
  )
```

### Interpreting the Discriminant Functions

To understand *which variables* drive group separation, we compare the standardised coefficients from Table \@ref(tab:std-coeffs) and inspect the **structure matrix** (Table \@ref(tab:structure-matrix)), which shows the correlation between each original variable and each discriminant function. The structure matrix is often easier to interpret than the raw or standardised coefficients.

```{r structure-matrix, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
# Compute structure matrix: correlations between variables and discriminant scores
ld_scores <- as.matrix(lda_pred$x)
X_scaled  <- scale(as.matrix(seed_traits[, traits]))
struct_mat <- round(cor(X_scaled, ld_scores), 3)
rownames(struct_mat) <- c("Width (mm)", "Length (mm)", "Thickness (mm)",
                           "Volume (mm³)", "Coat thickness (mm)",
                           "Wet weight (g)", "Dry weight (g)", "Moisture (%)")

as.data.frame(struct_mat) %>%
  tibble::rownames_to_column("Variable") %>%
  kbl(align = "c",
      col.names = c("Variable", "LD1", "LD2"),
      caption = "Structure matrix: correlations between each morphological variable and the two discriminant functions. Values above |0.30| are considered meaningful.") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE) %>%
  column_spec(1, bold = TRUE) %>%
  row_spec(which(abs(struct_mat[,"LD1"]) >= 0.30 | abs(struct_mat[,"LD2"]) >= 0.30),
           background = "#e8f4f8")
```

**Interpretation:**

-   **LD1** is most strongly correlated with overall size traits — volume, wet weight, dry weight, width, and length all have large positive correlations. This function separates the large-seeded *Garcinia cowa* from the other two species.
-   **LD2** is most strongly correlated with moisture content and coat thickness, providing additional separation between *Erythrina subumbrans* and *Cinnamomum caudatum* along a water-relations axis.

### Classifying a New (Unknown) Seed

A key practical use of LDA is **predicting the species of an unknown specimen**. Suppose a researcher finds a seed in the field with the following measurements:

```{r predict-new, echo=TRUE, message=FALSE, warning=FALSE}
new_seed <- data.frame(
  width_mm      = 10.5,
  long_mm       = 20.1,
  thickness_mm  = 9.8,
  volume_mm3    = 900,
  coat_thick_mm = 0.17,
  wet_weight_g  = 1.1,
  dry_weight_g  = 0.82,
  moisture_percent = 25.5
)

predict(lda_model, newdata = new_seed)
```

The model assigns the unknown seed to the species with the highest posterior probability. The researcher can use this classification, together with the posterior probability, to judge their confidence in the identification.

## Reporting LDA Results

> "Linear Discriminant Analysis was performed on eight seed morphological traits (width, length, thickness, volume, coat thickness, wet weight, dry weight, and moisture content) to evaluate the ability to discriminate among three tropical tree species (*Cinnamomum caudatum*, *Erythrina subumbrans*, and *Garcinia cowa*; n = 20 seeds per species). Two discriminant functions were extracted. LD1 accounted for `r pct_eig[1]`% of the between-group variance and was dominated by size-related traits (volume, wet weight, dry weight, and width), effectively separating *Garcinia cowa* (large seeds) from the other two species. LD2 accounted for the remaining `r pct_eig[2]`% and was most strongly associated with moisture content, further separating *Erythrina subumbrans* from *Cinnamomum caudatum*.
>
> Resubstitution classification accuracy was `r overall_acc`%. Leave-one-out cross-validation yielded an overall accuracy of `r cv_acc`%, demonstrating that the discriminant functions generalise well to new observations. These results indicate that seed morphology reliably distinguishes the three species in multivariate space."

## Summary

Table \@ref(tab:lda-summary) summarises the key decision points when performing LDA.

```{r lda-summary, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
sum_df <- data.frame(
  Step = c("Number of functions",
           "Most important step",
           "Key output 1",
           "Key output 2",
           "Key output 3",
           "Validation"),
  Guidance = c("Maximum = min(p, g − 1); often only the first 1–2 are meaningful",
               "Check assumptions — especially homogeneity of covariance matrices",
               "Standardised coefficients / structure matrix — identify key variables",
               "Discriminant score plot — visualize group separation",
               "Confusion matrix — assess classification accuracy",
               "Always use cross-validation (LOOCV) for a realistic accuracy estimate")
)

sum_df %>%
  kbl(align = "l",
      col.names = c("Step", "Guidance"),
      caption = "Key decision points in Linear Discriminant Analysis") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = TRUE) %>%
  column_spec(1, bold = TRUE)
```

## Exercise

**Scenario:** Use the seed morphology dataset (`seed3spp.csv`) to answer the following questions.

1.  Explain in your own words the difference between PCA and LDA. In what situation would you choose LDA over PCA?

2.  What is the maximum number of discriminant functions that can be extracted from this dataset (3 species, 8 variables)? Explain your reasoning.

3.  Look at the confusion matrix from resubstitution and from leave-one-out cross-validation.
    -   Which species is most accurately classified? Which is least accurately classified?
    -   Why is the cross-validated accuracy generally lower than the resubstitution accuracy?

4.  Using the structure matrix (Table \@ref(tab:structure-matrix)), name the **two variables** with the highest absolute correlation with LD1, and explain what biological characteristic of seeds this function represents.

5.  A researcher measures a newly collected seed with the following traits: width = 8.0 mm, length = 15.5 mm, thickness = 8.2 mm, volume = 560 mm³, coat thickness = 0.16 mm, wet weight = 0.95 g, dry weight = 0.72 g, moisture = 24.2%. Modify the `new_seed` code chunk to classify this seed and report (a) the predicted species and (b) the posterior probabilities for all three species.

6.  The homogeneity of covariance matrices is an important assumption of LDA.
    -   What test was used to evaluate this assumption in the example?
    -   If this assumption is violated, what alternative method could be used?

7.  **(Challenge)** Re-run the LDA using only the **three variables** with the highest absolute LD1 coefficients. Compare the cross-validated accuracy with the full eight-variable model. Does reducing the number of variables improve or worsen classification performance? Propose an explanation.

---

## Reference

-   Fisher, R. A. (1936). The use of multiple measurements in taxonomic problems. *Annals of Eugenics*, 7, 179–188.
-   Huberty, C. J. (1994). *Applied Discriminant Analysis*. Wiley.
-   McLachlan, G. J. (2004). *Discriminant Analysis and Statistical Pattern Recognition*. Wiley.
-   Tabachnick, B. G., & Fidell, L. S. (2013). *Using Multivariate Statistics* (6th ed.). Pearson.
-   Venables, W. N., & Ripley, B. D. (2002). *Modern Applied Statistics with S* (4th ed.). Springer. [`MASS` package]



## Exercise

1.    What are the two categories of multivariate analysis techniques? Give two examples each of those types.

2.    Use the seed morphology dataset (Table \@ref(tab:seed-data)) to answer the following questions.

  -   Why is the correlation matrix more appropriate than the covariance matrix for this dataset? List the units of at least three variables to justify your answer.
  -   What biological trait does PC1 represent? Name the three variables with the largest loadings on PC1 and explain what they have in common.

3.  The researcher plans to include a fourth species, which produces very small, lightweight seeds with high moisture content. Predict where this species would appear on the biplot (high or low PC1? high or low PC2?) and explain your reasoning.



---

## Reference

-   Anderson, E. (1935). The irises of the Gaspé Peninsula. *Bulletin of the American Iris Society*, 59, 2–5.
-   Fisher, R. A. (1936). The use of multiple measurements in taxonomic problems. *Annals of Eugenics*, 7, 179–188.
-   Jaadi, Z. Step-by-step explanation of Principal Component Analysis. https://builtin.com/data-science/step-step-explanation-principal-component-analysis
-   Pearson, K. (1901). On lines and planes of closest fit to systems of points in space. *Philosophical Magazine*, 2(11), 559–572.
-   Tabachnick, B. G., & Fidell, L. S. (1989). *Using Multivariate Statistics* (2nd ed.). Harper & Row.
-   Tabachnick, B. G., & Fidell, L. S. (1996). *Using Multivariate Statistics* (3rd ed.). Harper Collins.
