# Mixed Effects Models {#mixed-models}

**Duration:** 3-hour lab

## Learning outcomes

Students should be able to:

1. Demonstrate steps to run mixed models in R.
2. Explain the outputs of the mixed models.

## Data files for this lab

- root_data_chaiklang.csv [Click here to download](https://o365cmu-my.sharepoint.com/:x:/g/personal/pimonrat_t_cmu_ac_th/IQBYzegaBvK3R4Rn2HFDuFJnAfvfSC1WqiG5NPABDCzb1vo?e=fNLY3N)

## Introduction

In ecological and environmental studies, data are often collected from multiple locations, time points, or experimental units that are nested or grouped in some way. For example, when sampling species richness from different beaches, measurements from the same beach are likely to be more similar to each other than to measurements from different beaches. This structure in the data violates the independence assumption of traditional linear models.

Mixed effects models (also called mixed models or hierarchical models) provide a powerful solution by allowing us to account for both fixed effects (variables we are directly interested in testing) and random effects (grouping factors that introduce correlation in our data).

In this lab, we will learn how to build, compare, and interpret mixed effects models using the `lme4` package in R [@Bates2015].

## Download additional packages

We will use a number of packages in this lab, including `ggplot2` [@ggplot2], `dplyr` [@dplyr], `lme4` [@Bates2015], `lmerTest` [@lmerTest2017], `kableExtra` [@kableextra], `emmeans` [@emmeans], `multcomp`[@multcomp], and `effectsize` [@effectsize].

```{r setup-mixed, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
# Install and load required packages
if (!require("lme4"))     install.packages("lme4",     dependencies = TRUE)
if (!require("lmerTest")) install.packages("lmerTest", dependencies = TRUE)
if (!require("emmeans"))  install.packages("emmeans",  dependencies = TRUE)
if (!require("ggplot2"))  install.packages("ggplot2",  dependencies = TRUE)
if (!require("dplyr"))    install.packages("dplyr",    dependencies = TRUE)
if (!require("knitr"))    install.packages("knitr",    dependencies = TRUE)
if (!require("kableExtra")) install.packages("kableExtra", dependencies = TRUE)
if (!require("ggpattern")) install.packages("ggpattern", dependencies = TRUE)
if(!require("MuMIn")) install.packages("MuMIn", dependencies = TRUE)
if(!require("sfsmisc", character.only = TRUE)) {
  install.packages("sfsmisc", dependencies = TRUE)}
if(!require("formatdown", character.only = TRUE)) {
  install.packages("formatdown", dependencies = TRUE)}

library(lme4)
library(lmerTest)
library(emmeans)
library(ggplot2)
library(dplyr)
library(knitr)
library(kableExtra)
library(ggpattern)
library(MuMIn)
library(sfsmisc)
library(formatdown)
```

## The Data Set: Nursery Propagation Systems and Tree Performance

### Background

@Chaiklang2025 conducted an experiment testing whether growing tropical tree saplings in bottle crates improved seedling performance after planting out on a degraded restoration site in Krabi Province, southern Thailand.

The experiment used five framework tree species (*Saraca indica*, *Sandoricum koetjape*, *Cleistocalyx operculatus*, *Lepisanthes rubiginosa*, and *Garcinia hombroniana*) and compared three nursery production systems:

1.  **CON** (control): polybags standing directly on the ground
2.  **COB**: polybags in crates raised on wire benches
3.  **COG**: polybags in crates placed on the ground

The experiment used a **randomised complete block design (RCBD)** with three replicates (blocks) per species. Root length data are used as an example for the mixed model analysis.

```{r data-mixed, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
set.seed(213708)

chaiklang_data <- read.csv("data/root_data_chaiklang.csv", header = T)
chaiklang_data$Treatment <- factor(chaiklang_data$Treatment,
                                   levels = c("CON", "COB", "COG"))
chaiklang_data$Species   <- factor(chaiklang_data$Species,
                                   levels = c("C.operculatus", "G.hombroniana",
                   "L.rubiginosa",  "S.indica", "S.koetjape"))

head(chaiklang_data, 12)
unique(chaiklang_data$Species)
```

## Visualizing the Data

Let's explore the data set.

```{r chaiklang-boxplot, fig.height=5, fig.cap="Root length (cm) of five framework tree species under three nursery production systems. CON = polybags on ground (control); COB = crates on wire benches; COG = crates on ground. Points represent individual sapling measurements.", echo=FALSE, fig.align='center', message=FALSE, warning=FALSE}
ggplot(chaiklang_data,
       aes(x = Treatment, y = Rootlength, fill = Treatment)) +
  geom_boxplot_pattern(aes(pattern = Treatment), alpha = 0.6, outlier.shape = NA,
                       pattern_color = "black",
                       pattern_fill = "white",
                       pattern_density = 0.1) +
  #geom_jitter(width = 0.15, size = 0.8, alpha = 0.4) +
  facet_wrap(~ Species, nrow = 1) +
  #scale_fill_manual(values = c("CON" = "#4575b4",
  #                             "COB" = "#d73027",
  #                             "COG" = "#1a9850")) +
  labs(x = "Treatment",
       y = "Root length (cm)",
       fill = "Treatment") +
  theme_bw() +
  theme(axis.title = element_text(size = 14),
        axis.text.y  = element_text(size = 14),
    axis.text.x  = element_text(angle = 45, hjust = 1, size = 13),
    strip.text   = element_text(face = "italic", size = 13),
    legend.position = "none"
  )
```

Create a summary table of means and standard errors before modelling. 

```{r chaiklang-summary, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}
summary_ch <- chaiklang_data %>%
  group_by(Species, Treatment) %>%
  summarise(
    n    = n(),
    Mean = round(mean(Rootlength), 1),
    SD   = round(sd(Rootlength),   1),
    SE   = round(SD / sqrt(n), 1),
    .groups = "drop"
  )

summary_ch

## Make a nicer look table
kbl(summary_ch,
    col.names = c("Species", "Treatment", "n", "Mean root length (cm)",
                  "SD", "SE"),
    caption = "Summary statistics for root length (cm) by species and treatment") %>%
  kable_styling(full_width = TRUE, bootstrap_options = "striped") %>%
  column_spec(1, italic = TRUE)
```

## Fitting the Mixed Effects Model

First, let's start by running a **random intercept model** with 

- fixed effects = `Treatment`, `Species`, and their interaction
- a random effect = `Replicate`.

We use the `lmerTest` package [@lmerTest2017] which extends `lme4` [@Bates2015] to provide p-values in the summary table.

```{r fit-model1, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
# Fit the mixed effects model
# Fixed effects: Treatment + Species + Treatment:Species interaction
# Random effect: random intercept for Replicate (block)
mixed_model_IntOnly2 <- lmer(Rootlength ~ Treatment * Species + (1|Replicate),
                             REML = TRUE, data = chaiklang_data)
```

Then we do a **random slope and intercept model** and compare the two models and make a decision which one fit better according to AIC scores. 

```{r fit-model1, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
mixed_model_IntSlope2 <- lmer(Rootlength ~ Treatment * Species + 
                          (1 + Species|Replicate),
                          data = chaiklang_data, REML = TRUE)

summary(mixed_model_IntSlope2) ## it turns out to be too complex of a model

## We also take a look at the model without the fixed effect.

mixed_model_NoFix <- lmer(Rootlength ~ 1 + (1 | Replicate),
                          data = chaiklang_data, REML = TRUE)

summary(mixed_model_NoFix)

### model selection
AICc(mixed_model_IntOnly2, mixed_model_IntSlope2, mixed_model_NoFix) 
```

According to the AIC scores, the model with random intercept is better. Also, there is no interaction effect. We will cut the interaction from the model.

```{r fit-model-final, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}
## This model is for final results
model_final <- lmer(Rootlength ~ Treatment + Species + (1 | Replicate),
                 data = chaiklang_data,
                 REML = FALSE)

summary(model_final)
```

The output shows:

-   **Random effects**: the variance attributed to Replicate (between-block variation) and the residual variance (within-block variation)
-   **Fixed effects**: estimated coefficients for Treatment, Species, and their interaction, along with standard errors and p-values (from the Satterthwaite approximation in `lmerTest`)

## Model Validation

Before interpreting the model, we must check the residuals. The key assumptions of a linear mixed effects model are:

1.  Residuals are approximately normally distributed
2.  Residuals have roughly constant variance (homoscedasticity)
3.  Random effects are normally distributed

```{r model-validation, fig.height=4, fig.cap="Model diagnostics for the mixed effects model: (A) residuals vs. fitted values (checks homoscedasticity); (B) Q-Q plot of residuals (checks normality).", fig.align='center', echo=TRUE, message=FALSE, warning=FALSE}
# split the plotting area
par(mfrow = c(1, 2))

# Residuals vs fitted
plot(fitted(model_final), resid(model_final),
     xlab = "Fitted values", ylab = "Residuals",
     main = "(A) Residuals vs Fitted",
     pch = 16, col = "steelblue", cex = 0.6)
abline(h = 0, lty = 2, col = "red")

# Q-Q plot
qqnorm(resid(model_ch), main = "(B) Q-Q Plot of Residuals",
       pch = 16, col = "steelblue", cex = 0.6)
qqline(resid(model_ch), col = "red", lty = 2)

par(mfrow = c(1, 1))
```

No strong patterns in the residual plot; however, there are some outliers. Also, the Q-Q plot indicate that the model assumptions are reasonably satisfied.

## ANOVA Table for Fixed Effects

```{r anova-table-mixed, echo=FALSE, message=FALSE, warning=FALSE, results='asis'}

anova_final <- anova(model_final)

anova_final

## Make a nicer look table

kbl(round(as.data.frame(anova_ch), 2),
    caption = "ANOVA table (Type III Sums of Squares) for the mixed effects model of root length") %>%
  kable_styling(full_width = FALSE, bootstrap_options = "striped",font_size = 13)
```

**Interpretation of the results:** The interaction between the treatment and species variables are not significant. There are significant effects of Treatment (the crating treatments increasing root length) and Species (different species grow at different rates). 

Let's compare the treatments effects on root length, separately by species using estimated marginal means (EMMs) with Tukey's HSD correction (See @Chaiklang2025).

```{r posthoc-table-mixed, echo=TRUE, message=FALSE, warning=FALSE, include = FALSE}
# Estimated marginal means for Treatment within each Species
emm_ch <- emmeans(model_final, ~ Treatment | Species)

# Pairwise comparisons with Tukey adjustment
pairs_ch <- pairs(emm_ch, adjust = "tukey")
```

## Visualising Estimated Marginal Means

```{r emm-plot, fig.height=5, fig.cap="Estimated marginal means (±95% CI) for relative height growth rate (RGR-H) of five framework tree species under three nursery production systems from the mixed effects model.", echo=TRUE, message=FALSE, warning=FALSE, include = FALSE}

# I didn't use this in the lesson content. This might be useful for practical session
emm_df <- as.data.frame(emm_ch)

ggplot(emm_df, aes(x = Treatment, y = emmean,
                   colour = Treatment, group = Treatment)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2, linewidth = 0.8) +
  facet_wrap(~ Species, nrow = 1) +
  scale_colour_manual(values = c("CON" = "#4575b4",
                                 "COB" = "#d73027",
                                 "COG" = "#1a9850")) +
  labs(x = "Treatment",
       y = "Mean root length (cm)",
       colour = "Treatment") +
  theme_bw() +
  theme(
    axis.text.x  = element_text(angle = 45, hjust = 1),
    strip.text   = element_text(face = "italic", size = 9),
    legend.position = "bottom"
  )
```

## Reporting Results

The following paragraph presents the written report for a mixed effects model.

> A linear mixed effects model (lme4; Bates et al. 2015) was used to assess the effects of nursery production system (CON, COB, COG) and species on the root length of five framework tree species. Treatment and species were included as fixed effects along with their interaction. Replicate (block location in the nursery) was included as a random intercept to account for spatial variability. Significant effects of treatment (F = `r round(anova_final["Treatment","F value"], 2)`, P \< 0.05), species (F = `r round(anova_final["Species","F value"], 2)`, P \< 0.05) were detected. Post-hoc comparisons (Tukey HSD) showed that both crating treatments (COB and COG) significantly increased mean rooth length compared with the control for most species, consistent with the hypothesis that crating promotes air root-pruning and improves post-planting sapling performance.

## Exercise

**Submission:** A PDF file containing your answers.
Your answers must contain **R codes** that you wrote and **the outputs** you get.
Your codes and outputs can be text and/or screen capture.

**Full score:** 10 Submission:

**Data for this exercise:** rikz_data.txt. Download using codes below.

### Questions/directions

We will use the data - RIKZ dataset in @Zuur2007 and @Zuur2009. We will examine species richness at different beaches, to understand how environmental factors affect biodiversity while accounting for the nested structure of our data.
The data set includes:

1. Species richness
2. Beach: 9 beaches
3. NAP: Height compared to mean tidal level
4. Exposure: Index composed of the surf zone, slope, grain size, and depth of anaerobic layer
5. Site: 5 sites nested in the beaches

This exercise is modified from the lesson ***Linear mixed-effects models** and **Model selection** by James S. Santangelo [@santangelo1, @santangelo2]

```{r exercise-data, eval=FALSE, echo=TRUE, message=FALSE, warning=FALSE, results='hide'}

if(!require("readr"))install.packages("readr",dependencies = TRUE)
if(!require("plyr"))install.packages("plyr",dependencies = TRUE)

library(readr)
library(plyr)

rikz_data <- "https://uoftcoders.github.io/rcourse/data/rikz_data.txt"

download.file(rikz_data, "rikz_data.txt")

rikz_data <- read_delim("rikz_data.txt",
                        col_names = TRUE,
                        delim = "\t")

head(rikz_data)

## Explore the data
# Function to find polygons
find_hull <- function(df) df[chull(df$Richness, df$NAP), ]

# Identify polygons in data
hulls <- ddply(rikz_data, "Beach", find_hull)

# Plot
ggplot(rikz_data, aes(x = NAP, y = Richness, colour = as.factor(Beach))) +
  geom_point(size = 3) +
  theme_classic() +
  theme(legend.position = "none") +
  scale_colour_brewer(palette="Set1") +
  scale_fill_brewer(palette="Set1") +
  geom_polygon(data = hulls, aes(fill = as.factor(Beach), alpha = 0.2))

#### See the data separate by beach
ggplot(rikz_data, aes(x = NAP, y = Richness)) +
  geom_point() +
  geom_smooth(method = "lm") +
  facet_wrap(~ Beach) +
  xlab("NAP") + ylab("Richness") +
  theme_classic()
```

**1.** In `R`, run a mixed effect model with 

- fixed effects = NAP 
- random effect = Beach

show codes and other competing models that you run.

**2.**	In `R`, show the codes of how you select the best model and output of the best model

**3.**	In `R`, show code for model validation and explain in words if the model reach the assumption

**4.**	Write a paragraph to report the results of the model.

**5.**	In `R`, use this code to draw a graph and show the graph in the report. Insert the name of your best fit model where appropriate.

```{r eval=F}
rikz_data$fit_bestmodel <- predict(## name of your best fit model##
  ,type = 'response')

ggplot(rikz_data, aes(x = NAP, y = Richness, colour = Beach)) +
  # Add fitted values (i.e. regression) for each beach
  geom_line(aes(y = fit_bestmodel), size = 1) +
  geom_point(size = 3) +
  theme_classic() +
  theme(legend.position = "none") +
  scale_colour_brewer(palette="Set1")
```

### Grading Rubric

| Question | Score = 2 | Score = 1 | Score = 0 |
|----------|---------|---------|---------|
| 1 | Correct R code and outputs | One of R code or output is not included or incorrect | R code and output are not included or incorrect |
| 2 | Correct R code and outputs | One of R code or output is not included or incorrect | R code and output are not included or incorrect |
| 3 | Correct R code and explanation of the model checking | One of R code or the explanation is not included or incorrect | R code and the explanation are not included or incorrect |
| 4 | Correct results with statistical values reported | Partially correct results - missing important values | Incorrect results |
| 5 | Correct R code and a plot with proper axis labels | One of R code or output is not included or incorrect | R code and output are not included or incorrect |

**End of Lab 13**