---
title: "Lesson 13: Multiple Comparisons and Post-Hoc Tests"
author: "Pimonrat Tiansawat"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    toc_depth: 3
    number_sections: true
    theme: united
    highlight: tango
    code_folding: show
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE,
                      fig.width = 8, fig.height = 6, fig.align = 'center')
```

# Learning Objectives

By the end of this lesson, students should be able to:

- Understand why multiple comparisons require special statistical treatment
- Explain the concept of familywise error rate
- Choose appropriate post-hoc tests for different situations
- Conduct and interpret Tukey's HSD, Bonferroni, Dunnett's, and other tests in R
- Use planned contrasts vs. post-hoc comparisons appropriately
- Present multiple comparison results clearly

# The Multiple Comparisons Problem

## Why We Need Post-Hoc Tests

When ANOVA shows a significant F-statistic, we know that **at least one group mean differs** from the others. However, ANOVA doesn't tell us:

- Which specific groups are different?
- How many pairwise differences exist?
- The magnitude of these differences?

**This is where post-hoc tests come in!**

## The Problem with Multiple t-tests

### Example: Three Treatment Groups

In our fertilizer study, we have 3 groups and want to compare all pairs:

```{r calculate-comparisons}
# Number of groups
k <- 3

# Number of pairwise comparisons
n_comparisons <- k * (k - 1) / 2

cat("Number of groups:", k, "\n")
cat("Number of pairwise comparisons:", n_comparisons, "\n")

# List all pairs
groups <- c("No fertilizer", "A", "B")
pairs_list <- combn(groups, 2, simplify = FALSE)
cat("\nAll possible pairs:\n")
for (i in seq_along(pairs_list)) {
  cat(pairs_list[[i]][1], "vs", pairs_list[[i]][2], "\n")
}
```

### The Inflation of Type I Error

If we conduct each comparison at α = 0.05:

$$P(\text{at least one Type I error}) = 1 - (1 - \alpha)^c$$

where c = number of comparisons

```{r error-inflation}
library(ggplot2)

# Calculate familywise error rate
alpha <- 0.05
n_groups <- 2:10
n_comps <- n_groups * (n_groups - 1) / 2

# Familywise error rate (FWER)
fwer <- 1 - (1 - alpha)^n_comps

error_data <- data.frame(
  Groups = n_groups,
  Comparisons = n_comps,
  FWER = fwer,
  FWER_percent = fwer * 100
)

print(error_data)

# Visualize error inflation
ggplot(error_data, aes(x = Groups, y = FWER_percent)) +
  geom_line(color = "red", size = 1.5) +
  geom_point(size = 3, color = "darkred") +
  geom_hline(yintercept = 5, linetype = "dashed", color = "blue") +
  annotate("text", x = 8, y = 7, label = "Nominal α = 0.05", color = "blue") +
  scale_x_continuous(breaks = 2:10) +
  labs(title = "Inflation of Type I Error with Multiple Comparisons",
       subtitle = "Without correction for multiple testing",
       x = "Number of Groups",
       y = "Familywise Error Rate (%)") +
  theme_minimal()
```

**Key Finding:** With just 5 groups (10 comparisons), the probability of making at least one Type I error rises to **40%**, even though each individual test is conducted at α = 0.05!

## Familywise Error Rate (FWER)

**Definition:** The probability of making one or more Type I errors when conducting multiple hypothesis tests.

**Goal of post-hoc tests:** Control the FWER at a desired level (usually 0.05).

# Types of Multiple Comparison Procedures

## Planned Comparisons vs. Post-Hoc Tests

### Planned Comparisons (A Priori Contrasts)

- Specified **before** data collection
- Based on theoretical predictions
- More statistical power
- Fewer corrections needed

### Post-Hoc Tests (A Posteriori)

- Conducted **after** finding significant ANOVA
- Exploratory
- Compare all possible pairs or specific subsets
- Require stronger error rate control

# Common Post-Hoc Tests

## Overview of Methods

| Test | Use When | Conservativeness | Special Features |
|------|----------|------------------|------------------|
| **Tukey HSD** | All pairwise comparisons | Moderate | Most popular, good power |
| **Bonferroni** | Few planned comparisons | Conservative | Simple, very strict |
| **Holm** | Multiple comparisons | Less conservative than Bonferroni | Sequential method |
| **Scheffe** | Complex contrasts | Very conservative | Can test any contrast |
| **Dunnett** | Comparing to control | Moderate | Only vs. control group |
| **Fisher LSD** | Protected by significant ANOVA | Liberal | Use cautiously |
| **Games-Howell** | Unequal variances | Moderate | No homogeneity assumption |

# Example Dataset: Fertilizer Study

Let's use a simple fertilizer experiment with three treatments to demonstrate each test:

```{r create-dataset}
library(dplyr)
library(ggplot2)

set.seed(213708)

# Three fertilizer treatments: A, B, and No fertilizer (control)
fertilizer <- rep(c("A", "B", "No fertilizer"), each = 7)
height <- c(
  # Treatment A (lower heights)
  6, 3, 4, 5, 3, 6, 4,
  # Treatment B (higher heights)
  8, 5, 6, 7, 8, 6, 8,
  # Control (lowest heights)
  3, 2, 4, 5, 3, 3, 2
)

plant_data <- data.frame(
  Plant_ID = 1:21,
  Fertilizer = factor(fertilizer, levels = c("No fertilizer", "A", "B")),
  Height = height
)

# Display the data
cat("Dataset: Plant Height by Fertilizer Treatment\n")
cat("============================================\n\n")
print(plant_data)

# Summary statistics
summary_stats <- plant_data %>%
  group_by(Fertilizer) %>%
  summarise(
    n = n(),
    Mean = mean(Height),
    SD = sd(Height),
    SE = SD / sqrt(n),
    .groups = 'drop'
  )

cat("\n\nSummary Statistics:\n")
print(summary_stats)
```

## Visualize the Data

```{r visualize-data, fig.height=6}
# Boxplot with individual points
ggplot(plant_data, aes(x = Fertilizer, y = Height, fill = Fertilizer)) +
  geom_boxplot(alpha = 0.6, outlier.shape = NA) +
  geom_jitter(width = 0.2, alpha = 0.4, size = 2) +
  stat_summary(fun = mean, geom = "point", shape = 23, size = 3, 
               fill = "white", color = "black") +
  labs(title = "Plant Height by Fertilizer Treatment",
       subtitle = "Diamonds show group means",
       y = "Plant Height (cm)",
       x = "Fertilizer Treatment") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

# Mean plot with error bars
ggplot(summary_stats, aes(x = Fertilizer, y = Mean, fill = Fertilizer)) +
  geom_col(alpha = 0.7) +
  geom_errorbar(aes(ymin = Mean - SE, ymax = Mean + SE), 
                width = 0.3, size = 1) +
  geom_text(aes(label = round(Mean, 1)), vjust = -2.5, size = 3.5) +
  labs(title = "Mean Plant Height by Treatment",
       subtitle = "Error bars show ± 1 SE",
       y = "Mean Height (cm)") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

## Run ANOVA First

**Important:** Always run ANOVA before post-hoc tests. Post-hoc tests are only appropriate when ANOVA shows a significant result.

```{r anova-first}
# Always start with ANOVA
anova_model <- aov(Height ~ Fertilizer, data = plant_data)
anova_summary <- summary(anova_model)

print(anova_summary)

# Extract key statistics
f_stat <- anova_summary[[1]]["Fertilizer", "F value"]
p_value <- anova_summary[[1]]["Fertilizer", "Pr(>F)"]
df_treatment <- anova_summary[[1]]["Fertilizer", "Df"]
df_error <- anova_summary[[1]]["Residuals", "Df"]

cat("\nANOVA Results:\n")
cat("F-statistic: F(", df_treatment, ", ", df_error, ") = ", 
    round(f_stat, 3), "\n", sep = "")
cat("p-value:", format.pval(p_value, digits = 4), "\n")

if (p_value < 0.05) {
  cat("\n✓ ANOVA is significant (p < 0.05). Proceed with post-hoc tests.\n")
} else {
  cat("\n✗ ANOVA is not significant (p ≥ 0.05). Post-hoc tests not recommended.\n")
}

# Calculate effect size
library(effectsize)
eta_sq <- eta_squared(anova_model)
cat("\nEffect size (η²):", round(eta_sq$Eta2, 3), "\n")
```

# Tukey's Honestly Significant Difference (HSD)

## When to Use Tukey's HSD

- **All pairwise comparisons** among groups
- **Equal sample sizes** (works best, but robust to minor imbalances)
- **Homogeneity of variance** assumed
- **Most popular** post-hoc test

## How Tukey's HSD Works

Tukey's HSD compares all possible pairs of means using the **Studentized Range Distribution**.

### The Test Statistic

$$HSD = q_{\alpha, k, df_e} \times \sqrt{\frac{MS_E}{n}}$$

where:

- $q$ = critical value from studentized range distribution
- $k$ = number of groups
- $df_e$ = error degrees of freedom
- $MS_E$ = mean square error
- $n$ = sample size per group (assumes equal n)

## Running Tukey's HSD in R

```{r tukey-hsd}
# Perform Tukey's HSD test
tukey_results <- TukeyHSD(anova_model, conf.level = 0.95)

print(tukey_results)

# Access the table
tukey_table <- as.data.frame(tukey_results$Fertilizer)
tukey_table$Comparison <- rownames(tukey_table)
tukey_table <- tukey_table[, c("Comparison", "diff", "lwr", "upr", "p adj")]
colnames(tukey_table) <- c("Comparison", "Mean_Diff", "Lower_CI", "Upper_CI", "p_adj")

# Add significance indicator
tukey_table$Significant <- ifelse(tukey_table$p_adj < 0.05, "***", 
                                  ifelse(tukey_table$p_adj < 0.10, "*", "ns"))

print(tukey_table)
```

## Visualizing Tukey's HSD Results

```{r tukey-plot, fig.height=8}
# Plot confidence intervals
par(mar = c(5, 2, 4, 2))
plot(tukey_results, las = 1, col = "blue")
abline(v = 0, lty = 2, col = "red")
title(main = "Tukey's HSD: 95% Confidence Intervals for Pairwise Differences")

# Reset margins
par(mar = c(5, 4, 4, 2))
```

```{r tukey-plot-ggplot, fig.height=8}
# Using ggplot2 for better visualization
library(broom)

tukey_df <- tidy(tukey_results)

ggplot(tukey_df, aes(x = estimate, y = contrast)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", size = 1) +
  geom_errorbarh(aes(xmin = conf.low, xmax = conf.high), height = 0.3, size = 1) +
  geom_point(size = 3, color = "blue") +
  labs(title = "Tukey's HSD: Pairwise Comparisons",
       subtitle = "95% Confidence Intervals for Mean Differences",
       x = "Mean Difference in Height (cm)",
       y = "Comparison") +
  theme_minimal() +
  theme(axis.text.y = element_text(size = 10))
```

## Compact Letter Display (CLD)

A convenient way to show which groups differ:

```{r cld-tukey}
library(multcomp)

# Compact letter display
cld_model <- glht(anova_model, linfct = mcp(Fertilizer = "Tukey"))
cld_results <- cld(cld_model, level = 0.05, Letters = letters)

print(cld_results)

# Extract letters
cld_df <- data.frame(
  Fertilizer = names(cld_results$mcletters$Letters),
  Letter = cld_results$mcletters$Letters
)

# Merge with summary stats
summary_with_letters <- summary_stats %>%
  left_join(cld_df, by = "Fertilizer")

print(summary_with_letters)
```

```{r plot-cld, fig.height=6}
# Plot with significance letters
ggplot(summary_with_letters, aes(x = Fertilizer, y = Mean, fill = Fertilizer)) +
  geom_col(alpha = 0.7) +
  geom_errorbar(aes(ymin = Mean - SE, ymax = Mean + SE), 
                width = 0.3, size = 1) +
  geom_text(aes(label = Letter, y = Mean + SE + 2), 
            size = 6, fontface = "bold") +
  labs(title = "Plant Height by Fertilizer with Tukey's HSD Results",
       subtitle = "Groups sharing a letter are not significantly different (α = 0.05)",
       y = "Mean Height (cm)",
       x = "Fertilizer Treatment") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
```

# Bonferroni Correction

## When to Use Bonferroni

- **Small number** of planned comparisons
- **Maximum control** of Type I error
- Simple to understand and calculate
- Can be overly conservative with many comparisons

## How Bonferroni Works

The Bonferroni correction adjusts the significance level:

$$\alpha_{adjusted} = \frac{\alpha_{family}}{c}$$

where c = number of comparisons

**Example:** With 10 comparisons and α = 0.05:
- Adjusted α = 0.05 / 10 = 0.005
- Each test must have p < 0.005 to be significant

## Running Bonferroni in R

```{r bonferroni}
# Pairwise t-tests with Bonferroni correction
bonferroni_results <- pairwise.t.test(
  plant_data$Height,
  plant_data$Fertilizer,
  p.adjust.method = "bonferroni"
)

print(bonferroni_results)

# Create a more readable table
library(tidyr)

bonf_matrix <- bonferroni_results$p.value
bonf_df <- as.data.frame(bonf_matrix)
bonf_df$Group1 <- rownames(bonf_df)

bonf_long <- bonf_df %>%
  pivot_longer(cols = -Group1, names_to = "Group2", values_to = "p_value") %>%
  filter(!is.na(p_value)) %>%
  mutate(
    Comparison = paste(Group1, "vs", Group2),
    Significant = ifelse(p_value < 0.05, "Yes", "No"),
    Sig_code = ifelse(p_value < 0.001, "***",
                      ifelse(p_value < 0.01, "**",
                             ifelse(p_value < 0.05, "*", "ns")))
  ) %>%
  arrange(p_value)

print(bonf_long)
```

## Visualizing Bonferroni Results

```{r bonferroni-heatmap, fig.height=6, fig.width=7}
# Create a heatmap of p-values
bonf_matrix_full <- bonf_matrix
bonf_matrix_full[is.na(bonf_matrix_full)] <- 1  # Fill diagonal with 1

library(reshape2)
bonf_melted <- melt(as.matrix(bonf_matrix_full))

ggplot(bonf_melted, aes(x = Var2, y = Var1, fill = value)) +
  geom_tile(color = "white") +
  geom_text(aes(label = ifelse(value < 1, sprintf("%.3f", value), "")), 
            size = 3) +
  scale_fill_gradient2(low = "red", mid = "yellow", high = "white",
                       midpoint = 0.05, limit = c(0, 0.15),
                       name = "p-value") +
  labs(title = "Bonferroni Correction: Pairwise Comparisons",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

# Holm's Sequential Method

## When to Use Holm's Method

- Alternative to Bonferroni
- **Less conservative** while still controlling FWER
- Good for many comparisons

## How Holm's Method Works

1. Order p-values from smallest to largest
2. Compare each p-value to adjusted threshold:
   - 1st p-value: α/c
   - 2nd p-value: α/(c-1)
   - 3rd p-value: α/(c-2)
   - etc.
3. Stop at first non-significant result

```{r holm}
# Pairwise t-tests with Holm correction
holm_results <- pairwise.t.test(
  plant_data$Height,
  plant_data$Fertilizer,
  p.adjust.method = "holm"
)

print(holm_results)

# Compare Bonferroni vs. Holm
comparison <- data.frame(
  Comparison = bonf_long$Comparison,
  Raw_p = NA,  # Would need to calculate separately
  Bonferroni_p = bonf_long$p_value
)

# Get Holm p-values
holm_matrix <- holm_results$p.value
holm_df <- as.data.frame(holm_matrix)
holm_df$Group1 <- rownames(holm_df)

holm_long <- holm_df %>%
  pivot_longer(cols = -Group1, names_to = "Group2", values_to = "Holm_p") %>%
  filter(!is.na(Holm_p)) %>%
  mutate(Comparison = paste(Group1, "vs", Group2))

# Merge
comparison_full <- bonf_long %>%
  select(Comparison, p_value) %>%
  rename(Bonferroni_p = p_value) %>%
  left_join(holm_long %>% select(Comparison, Holm_p), by = "Comparison") %>%
  mutate(
    Bonf_Sig = ifelse(Bonferroni_p < 0.05, "Yes", "No"),
    Holm_Sig = ifelse(Holm_p < 0.05, "Yes", "No"),
    Agreement = ifelse(Bonf_Sig == Holm_Sig, "Same", "Different")
  )

print(comparison_full)
```

# Dunnett's Test

## When to Use Dunnett's Test

- **Comparing all treatments to a control group**
- Not interested in comparing treatments to each other
- More powerful than Tukey for this specific purpose
- In our example: Compare fertilizers A and B to "No fertilizer" control

## Running Dunnett's Test in R

```{r dunnett}
library(DescTools)

# Dunnett's test (comparing all to No fertilizer control)
dunnett_results <- DunnettTest(
  Height ~ Fertilizer,
  data = plant_data,
  control = "No fertilizer"  # Specify control group
)

print(dunnett_results)

cat("\nInterpretation:\n")
cat("This test compares Fertilizer A and Fertilizer B to the control (No fertilizer)\n")
cat("It does NOT compare A vs B (not needed for this research question)\n")
```

```{r dunnett-plot, fig.height=5}
# Visualize Dunnett's results
dunnett_df <- as.data.frame(dunnett_results$`No fertilizer`)
dunnett_df$Comparison <- rownames(dunnett_df)

ggplot(dunnett_df, aes(x = Comparison, y = diff)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_errorbar(aes(ymin = lwr.ci, ymax = upr.ci), width = 0.3, size = 1) +
  geom_point(size = 4, color = "blue") +
  geom_text(aes(label = ifelse(pval < 0.05, "*", "ns")), 
            vjust = -1, size = 5) +
  labs(title = "Dunnett's Test: Comparisons to Control",
       subtitle = "95% Confidence Intervals",
       x = "Comparison vs. Control",
       y = "Mean Difference") +
  theme_minimal() +
  coord_flip()
```

# Games-Howell Test

## When to Use Games-Howell

- **Variances are NOT equal** across groups
- Unequal sample sizes
- Alternative to Tukey's HSD when assumptions violated

## Running Games-Howell Test in R

```{r games-howell}
library(rstatix)

# Games-Howell test
gh_results <- games_howell_test(plant_data, Height ~ Fertilizer)

print(gh_results)

# Add significance stars
gh_results <- gh_results %>%
  mutate(
    sig = case_when(
      p.adj < 0.001 ~ "***",
      p.adj < 0.01 ~ "**",
      p.adj < 0.05 ~ "*",
      TRUE ~ "ns"
    )
  )

print(gh_results)
```

# Fisher's Least Significant Difference (LSD)

## When to Use Fisher's LSD

- **Only after significant ANOVA** (Fisher's Protected LSD)
- Comparing just a few specific pairs
- Understanding: more liberal, less conservative

## Running Fisher's LSD in R

```{r fisher-lsd}
library(agricolae)

# Fisher's LSD test
lsd_results <- LSD.test(anova_model, "Fertilizer", p.adj = "none")

print(lsd_results)

# Extract groups
lsd_groups <- lsd_results$groups
print(lsd_groups)
```

# Scheffé's Test

## When to Use Scheffé's Test

- Testing **complex contrasts** (not just pairwise)
- Most conservative method
- Can test any contrast, even after looking at data

## Running Scheffé's Test in R

```{r scheffe}
library(DescTools)

# Scheffé test for all pairwise comparisons
scheffe_results <- ScheffeTest(anova_model)

print(scheffe_results)
```

# Comparison of All Methods

Let's compare all methods side by side:

```{r compare-methods, fig.height=10}
# Function to get significant pairs from each method
get_sig_pairs <- function(results_list, method_name) {
  # This is simplified - actual implementation depends on result structure
  tibble(Method = method_name)
}

# Create comparison summary
library(knitr)
library(kableExtra)

comparison_summary <- data.frame(
  Method = c("Tukey HSD", "Bonferroni", "Holm", "Dunnett", 
             "Games-Howell", "Fisher LSD", "Scheffé"),
  Conservativeness = c("Moderate", "High", "Moderate-High", "Moderate",
                       "Moderate", "Low", "Very High"),
  Use_Case = c(
    "All pairwise, equal n",
    "Few planned comparisons",
    "Many comparisons, less conservative",
    "Compare to control only",
    "Unequal variances",
    "Post-hoc, few comparisons",
    "Complex contrasts"
  ),
  Assumptions = c(
    "Homogeneity of variance",
    "None (but uses t-tests)",
    "None",
    "Homogeneity of variance",
    "Unequal variances OK",
    "Homogeneity of variance",
    "Homogeneity of variance"
  )
)

kable(comparison_summary, 
      caption = "Comparison of Post-Hoc Test Methods") %>%
  kable_styling(bootstrap_options = c("striped", "hover"),
                full_width = FALSE) %>%
  column_spec(1, bold = TRUE)
```

## Comparing Power and Type I Error

```{r power-comparison, fig.height=6}
# Conceptual visualization of conservativeness
methods_df <- data.frame(
  Method = factor(c("Fisher LSD", "Tukey HSD", "Holm", "Bonferroni", "Scheffé"),
                  levels = c("Fisher LSD", "Tukey HSD", "Holm", "Bonferroni", "Scheffé")),
  Conservativeness = c(2, 5, 6, 8, 9),
  Power = c(9, 7, 6.5, 4, 3)
)

library(patchwork)

p1 <- ggplot(methods_df, aes(x = Method, y = Conservativeness, fill = Method)) +
  geom_col(alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Conservativeness",
       subtitle = "Higher = stricter control of Type I error",
       y = "Relative Conservativeness") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- ggplot(methods_df, aes(x = Method, y = Power, fill = Method)) +
  geom_col(alpha = 0.7) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Statistical Power",
       subtitle = "Higher = better at detecting true differences",
       y = "Relative Power") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))

p1 / p2 + plot_annotation(
  title = "Trade-off Between Type I Error Control and Statistical Power",
  theme = theme(plot.title = element_text(face = "bold", size = 13))
)
```

**Key Insight:** There is always a trade-off between:
- **Type I error control** (avoiding false positives) 
- **Statistical power** (detecting real differences)

Choose your method based on what's more important for your research question!

# Planned Contrasts

## What Are Contrasts?

**Contrasts** are specific comparisons planned before data collection, based on theory or research questions.

### Example Research Questions:

1. Do fertilizer treatments differ from control? (Treatment vs. Control)
2. Do organic methods differ from synthetic methods?
3. Is there a linear trend across fertilizer concentrations?

## Types of Contrasts

### 1. Simple Contrasts

Compare each group to a reference (e.g., control):

```{r simple-contrasts}
# Set "No fertilizer" as reference
plant_data$Fertilizer <- relevel(plant_data$Fertilizer, ref = "No fertilizer")

# Fit model
contrast_model <- lm(Height ~ Fertilizer, data = plant_data)
summary(contrast_model)

cat("\nInterpretation:\n")
cat("(Intercept): Mean height for No fertilizer group\n")
cat("FertilizerA: Difference between Fertilizer A and No fertilizer\n")
cat("FertilizerB: Difference between Fertilizer B and No fertilizer\n")
```

### 2. Orthogonal Contrasts

Independent contrasts that don't overlap. For our 3 groups, we can create 2 orthogonal contrasts:

```{r orthogonal-contrasts}
# Define orthogonal contrasts for 3 groups
# Contrast 1: Control vs. Both Fertilizers (A and B combined)
# Contrast 2: Fertilizer A vs. Fertilizer B

library(emmeans)

# Get estimated marginal means
emm <- emmeans(anova_model, ~ Fertilizer)

# Define custom contrasts
contrast_list <- list(
  "Control vs Fertilizers" = c(2, -1, -1),  # No fert vs (A+B)/2
  "Fertilizer A vs B" = c(0, 1, -1)         # A vs B
)

# Test contrasts
contrast_results <- contrast(emm, contrast_list, adjust = "none")
print(contrast_results)
confint(contrast_results)

cat("\n\nVerifying Orthogonality:\n")
cat("Two contrasts are orthogonal if sum of products = 0\n")
c1 <- contrast_list[[1]]
c2 <- contrast_list[[2]]
cat("Contrast 1:", c1, "\n")
cat("Contrast 2:", c2, "\n")
cat("Sum of products:", sum(c1 * c2), "\n")
if (sum(c1 * c2) == 0) {
  cat("✓ Contrasts are orthogonal!\n")
}
```

### 3. Visualizing Contrasts

```{r visualize-contrasts, fig.height=6}
# Visualize the contrasts
contrast_df <- as.data.frame(contrast_results)
contrast_df$Contrast <- rownames(contrast_df)

ggplot(contrast_df, aes(x = Contrast, y = estimate)) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), 
                width = 0.3, size = 1.2) +
  geom_point(size = 4, color = "blue") +
  geom_text(aes(label = ifelse(p.value < 0.05, "*", "ns")), 
            vjust = -1.5, size = 6, fontface = "bold") +
  labs(title = "Planned Orthogonal Contrasts",
       subtitle = "95% Confidence Intervals\n* indicates p < 0.05",
       x = "Contrast",
       y = "Estimated Difference") +
  theme_minimal() +
  coord_flip()
```

# Visualization Best Practices

## 1. Significance Letters on Bar Plots

```{r vis-letters, fig.height=6}
# Using Tukey results with compact letter display
summary_final <- summary_with_letters

ggplot(summary_final, aes(x = Fertilizer, y = Mean, fill = Fertilizer)) +
  geom_col(alpha = 0.8, color = "black", size = 0.5) +
  geom_errorbar(aes(ymin = Mean - SE, ymax = Mean + SE), 
                width = 0.25, size = 0.8) +
  geom_text(aes(label = Letter, y = Mean + SE + 1.5), 
            size = 8, fontface = "bold") +
  geom_text(aes(label = round(Mean, 1), y = Mean/2),
            color = "white", fontface = "bold", size = 4) +
  scale_fill_brewer(palette = "Set2") +
  labs(title = "Plant Height by Fertilizer Treatment",
       subtitle = "Groups with different letters are significantly different (Tukey's HSD, α = 0.05)",
       y = "Mean Height (cm)",
       x = "Fertilizer Treatment") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
        plot.title = element_text(size = 14, face = "bold"))
```

## 2. Pairwise Comparison Matrix

```{r vis-matrix, fig.height=7, fig.width=8}
# Create significance matrix from Tukey results
sig_matrix <- matrix(NA, nrow = 5, ncol = 5)
rownames(sig_matrix) <- colnames(sig_matrix) <- levels(plant_data$Fertilizer)

# Fill matrix with p-values
tukey_table_matrix <- tukey_results$Fertilizer
for (i in 1:nrow(tukey_table_matrix)) {
  comparison <- strsplit(rownames(tukey_table_matrix)[i], "-")[[1]]
  group1 <- comparison[1]
  group2 <- comparison[2]
  p_val <- tukey_table_matrix[i, "p adj"]
  
  sig_matrix[group2, group1] <- p_val
}

# Create heatmap
library(reshape2)
sig_melted <- melt(sig_matrix)
sig_melted$Significance <- cut(sig_melted$value,
                               breaks = c(0, 0.001, 0.01, 0.05, 1),
                               labels = c("***", "**", "*", "ns"),
                               include.lowest = TRUE)

ggplot(sig_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white", size = 1) +
  geom_text(aes(label = Significance), size = 6, fontface = "bold") +
  scale_fill_gradient2(low = "darkred", mid = "yellow", high = "lightgreen",
                       midpoint = 0.05, na.value = "grey90",
                       name = "p-value",
                       limits = c(0, 0.15)) +
  labs(title = "Pairwise Comparison Matrix (Tukey's HSD)",
       subtitle = "*** p<0.001  ** p<0.01  * p<0.05  ns = not significant",
       x = NULL, y = NULL) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 11),
        axis.text.y = element_text(size = 11))
```

## 3. Forest Plot for Effect Sizes

```{r forest-plot, fig.height=6}
# Calculate effect sizes (Cohen's d) for each comparison vs. control
control_mean <- summary_stats$Mean[summary_stats$Fertilizer == "Control"]
control_sd <- summary_stats$SD[summary_stats$Fertilizer == "Control"]

effect_sizes <- summary_stats %>%
  filter(Fertilizer != "Control") %>%
  mutate(
    Cohens_d = (Mean - control_mean) / 
      sqrt((SD^2 + control_sd^2) / 2),
    Lower_CI = Cohens_d - 1.96 * sqrt(2/n),
    Upper_CI = Cohens_d + 1.96 * sqrt(2/n)
  )

ggplot(effect_sizes, aes(x = Cohens_d, y = Fertilizer)) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "gray50") +
  geom_vline(xintercept = c(0.2, 0.5, 0.8), linetype = "dotted", 
             color = "gray70", alpha = 0.5) +
  geom_errorbarh(aes(xmin = Lower_CI, xmax = Upper_CI), 
                 height = 0.2, size = 1) +
  geom_point(size = 4, color = "darkblue") +
  annotate("text", x = 0.2, y = 4.5, label = "Small", angle = 90, 
           color = "gray50", size = 3) +
  annotate("text", x = 0.5, y = 4.5, label = "Medium", angle = 90, 
           color = "gray50", size = 3) +
  annotate("text", x = 0.8, y = 4.5, label = "Large", angle = 90, 
           color = "gray50", size = 3) +
  labs(title = "Effect Sizes: Treatments vs. Control",
       subtitle = "Cohen's d with 95% confidence intervals",
       x = "Effect Size (Cohen's d)",
       y = "Treatment") +
  theme_minimal()
```

# Decision Tree: Choosing a Post-Hoc Test

```{r decision-tree, echo=FALSE, fig.height=8}
library(DiagrammeR)

grViz("
digraph post_hoc_decision {
  
  # Graph attributes
  graph [layout = dot, rankdir = TB]
  
  # Node definitions
  node [shape = box, style = filled, fillcolor = lightblue, fontname = Helvetica]
  
  start [label = 'ANOVA Significant?', fillcolor = lightgreen]
  stop [label = 'STOP\nNo post-hoc tests needed', fillcolor = pink, shape = oval]
  
  q1 [label = 'What comparisons?']
  
  all_pairs [label = 'All pairwise\ncomparisons?']
  vs_control [label = 'All vs.\nControl only?']
  planned [label = 'Planned\ncontrasts?']
  
  equal_var [label = 'Equal variances?']
  
  tukey [label = 'Use Tukey HSD', fillcolor = yellow, shape = oval]
  gh [label = 'Use Games-Howell', fillcolor = yellow, shape = oval]
  dunnett [label = 'Use Dunnett', fillcolor = yellow, shape = oval]
  contrasts [label = 'Use Planned Contrasts\n(no correction needed)', fillcolor = yellow, shape = oval]
  
  few [label = 'Few comparisons?']
  bonf [label = 'Use Bonferroni', fillcolor = yellow, shape = oval]
  holm [label = 'Use Holm', fillcolor = yellow, shape = oval]
  
  # Edges
  start -> stop [label = 'No']
  start -> q1 [label = 'Yes']
  
  q1 -> all_pairs [label = 'All pairs']
  q1 -> vs_control [label = 'vs Control']
  q1 -> planned [label = 'Planned']
  q1 -> few [label = 'Few specific']
  
  all_pairs -> equal_var
  equal_var -> tukey [label = 'Yes']
  equal_var -> gh [label = 'No']
  
  vs_control -> dunnett
  planned -> contrasts
  
  few -> bonf [label = 'Yes (< 5)']
  few -> holm [label = 'No (≥ 5)']
}
", width = 800, height = 700)
```

# Reporting Multiple Comparison Results

## Template for Writing Results

### Example 1: Tukey's HSD

> "A one-way ANOVA revealed a significant effect of fertilizer treatment on plant height, F(2, 18) = 8.52, p = 0.003, η² = 0.49. Post-hoc comparisons using Tukey's HSD test indicated that plants receiving Fertilizer B (M = 6.86, SD = 1.35) grew significantly taller than those in the control group with no fertilizer (M = 3.14, SD = 1.07, p = 0.002). Fertilizer A (M = 4.43, SD = 1.27) showed an intermediate height that was not significantly different from either the control (p = 0.16) or Fertilizer B (p = 0.053)."

### Example 2: With Compact Letter Display

> "Tukey's HSD post-hoc test revealed that Fertilizer B produced the greatest mean plant height (6.86 cm, group 'b'), which was significantly different from the control group with no fertilizer (3.14 cm, group 'a'). Fertilizer A (4.43 cm, group 'ab') showed intermediate growth that did not differ significantly from either group at α = 0.05."

### Example 3: Dunnett's Test

> "Dunnett's test comparing both fertilizer treatments to the control group showed that Fertilizer B produced significantly taller plants than the control (mean difference = 3.71 cm, 95% CI [1.53, 5.90], p = 0.002). Fertilizer A showed a non-significant trend toward greater height compared to control (mean difference = 1.29 cm, 95% CI [-0.90, 3.47], p = 0.20)."

### Example 4: Planned Contrasts

> "Two planned orthogonal contrasts were conducted. The first contrast, comparing the control group to both fertilizer treatments combined, was significant (t(18) = 3.85, p = 0.001), indicating that fertilization increased plant height. The second contrast, comparing Fertilizer A to Fertilizer B, approached significance (t(18) = 2.04, p = 0.056), suggesting that Fertilizer B may be more effective than Fertilizer A."

## Creating Publication-Ready Tables

```{r publication-table}
library(knitr)
library(kableExtra)

# Summary table with post-hoc results
pub_table <- summary_final %>%
  mutate(
    Mean_SD = sprintf("%.2f (%.2f)", Mean, SD),
    SE = sprintf("%.2f", SE),
    Group = Letter
  ) %>%
  select(Fertilizer, n, Mean_SD, SE, Group)

kable(pub_table,
      col.names = c("Treatment", "n", "Mean (SD)", "SE", "Tukey Group"),
      caption = "Table 1. Plant Height (cm) by Fertilizer Treatment with Post-Hoc Groupings",
      align = c("l", "c", "c", "c", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE) %>%
  footnote(general = "Groups with different letters are significantly different at α = 0.05 (Tukey's HSD test). Data represent mean plant height from 7 plants per treatment.",
           general_title = "Note:")
```

```{r anova-table}
# ANOVA table
anova_table <- tidy(anova_model)

kable(anova_table,
      col.names = c("Source", "df", "Sum Sq", "Mean Sq", "F", "p-value"),
      caption = "Table 2. Analysis of Variance for Plant Height",
      digits = c(0, 0, 2, 2, 2, 4),
      align = c("l", "c", "c", "c", "c", "c")) %>%
  kable_styling(bootstrap_options = c("striped", "hover", "condensed"),
                full_width = FALSE) %>%
  row_spec(0, bold = TRUE)
```

# Common Mistakes to Avoid

## Mistake 1: Doing Post-Hoc Tests Without Significant ANOVA

```{r mistake1, eval=FALSE}
# ❌ WRONG: Doing post-hoc tests when ANOVA is not significant
# If p > 0.05 from ANOVA, STOP. Don't do post-hoc tests.

# ✓ RIGHT: Check ANOVA first
if (anova_p < 0.05) {
  # Proceed with post-hoc tests
  tukey_results <- TukeyHSD(anova_model)
} else {
  cat("ANOVA not significant. Post-hoc tests not appropriate.\n")
}
```

## Mistake 2: Not Correcting for Multiple Comparisons

```{r mistake2, eval=FALSE}
# ❌ WRONG: Multiple t-tests without correction
# This inflates Type I error!
t.test(height[group == "A"], height[group == "B"])
t.test(height[group == "A"], height[group == "C"])
t.test(height[group == "B"], height[group == "C"])
# ... etc.

# ✓ RIGHT: Use post-hoc tests with correction
TukeyHSD(anova_model)
# OR
pairwise.t.test(height, group, p.adjust.method = "bonferroni")
```

## Mistake 3: Using the Wrong Test

```{r mistake3, eval=FALSE}
# ❌ WRONG: Using Tukey when variances are unequal
# Check assumptions first!

# ✓ RIGHT: Check assumptions, use appropriate test
leveneTest(Height ~ Fertilizer, data = plant_data)

if (levene_p < 0.05) {
  # Variances unequal - use Games-Howell
  games_howell_test(plant_data, Height ~ Fertilizer)
} else {
  # Variances equal - use Tukey
  TukeyHSD(anova_model)
}
```

## Mistake 4: Over-interpreting Non-Significant Results

```{r mistake4, eval=FALSE}
# ❌ WRONG: "Groups A and B are the same" (when p > 0.05)

# ✓ RIGHT: "No significant difference was found between groups A and B"
# Absence of evidence ≠ Evidence of absence
```

# Practice Exercises

## Exercise 1: Basic Post-Hoc Testing

Analyze this dataset and conduct appropriate post-hoc tests:

```{r exercise1}
set.seed(999)
n <- 10

study_data <- data.frame(
  Method = factor(rep(c("Method_A", "Method_B", "Method_C", "Method_D"), each = n)),
  Score = c(
    rnorm(n, 75, 8),
    rnorm(n, 82, 7),
    rnorm(n, 78, 9),
    rnorm(n, 85, 7)
  )
)

# Your tasks:
# 1. Run ANOVA
# 2. Check assumptions
# 3. Conduct Tukey's HSD
# 4. Create visualization with significance letters
# 5. Write results paragraph
```

## Exercise 2: Choosing the Right Test

For each scenario, choose the most appropriate post-hoc test:

```{r exercise2-scenarios, eval=FALSE}
# Scenario A: Testing 5 drug treatments vs. placebo
# Your answer: _______________

# Scenario B: Comparing 6 groups, variances very different
# Your answer: _______________

# Scenario C: 4 specific planned comparisons based on theory
# Your answer: _______________

# Scenario D: Exploratory study with 8 groups, all pairwise comparisons
# Your answer: _______________
```

## Exercise 3: Complex Contrasts

```{r exercise3}
# Three crop varieties grown under three water conditions
crop_data <- expand.grid(
  Variety = factor(c("V1", "V2", "V3")),
  Water = factor(c("Low", "Medium", "High"))
)

set.seed(111)
crop_data$Yield <- with(crop_data,
  50 + 
  10 * as.numeric(Variety) +
  8 * as.numeric(Water) +
  3 * as.numeric(Variety) * as.numeric(Water) +
  rnorm(nrow(crop_data), 0, 5)
)

# Replicate each combination
crop_data <- crop_data[rep(1:nrow(crop_data), each = 5), ]
crop_data$Yield <- crop_data$Yield + rnorm(nrow(crop_data), 0, 3)

# Your tasks:
# 1. Run two-way ANOVA
# 2. If no interaction: test main effects with appropriate post-hoc
# 3. If interaction: conduct simple effects analysis
# 4. Plan and test 2 orthogonal contrasts
```

# Summary

**Key Concepts:**

1. **Multiple comparisons inflate Type I error** - must be controlled
2. **Choose post-hoc test based on:**
   - Type of comparisons (all pairs, vs. control, specific)
   - Sample size and equality
   - Variance homogeneity
   - Conservativeness desired

3. **Most common choices:**
   - Tukey's HSD for all pairwise comparisons
   - Bonferroni for few planned comparisons
   - Dunnett for comparisons to control
   - Games-Howell for unequal variances

4. **Planned contrasts** are more powerful than post-hoc tests

5. **Always check ANOVA significance** before post-hoc tests

**R Functions Summary:**

```{r summary-functions, eval=FALSE}
# Post-hoc tests
TukeyHSD(model)                                    # Tukey's HSD
pairwise.t.test(y, group, p.adjust = "bonferroni") # Bonferroni
pairwise.t.test(y, group, p.adjust = "holm")       # Holm
DunnettTest(y ~ group, control = "Control")        # Dunnett
games_howell_test(data, y ~ group)                 # Games-Howell
LSD.test(model, "factor")                          # Fisher's LSD
ScheffeTest(model)                                 # Scheffé

# Visualization
cld(glht(model, linfct = mcp(factor = "Tukey")))  # Compact letter display

# Planned contrasts
contrast(emmeans(model, ~ factor), list)           # Custom contrasts
```

# References

- Maxwell, S. E., & Delaney, H. D. (2004). *Designing Experiments and Analyzing Data: A Model Comparison Perspective* (2nd ed.). Lawrence Erlbaum Associates.
- Hochberg, Y., & Tamhane, A. C. (1987). *Multiple Comparison Procedures*. Wiley.
- Toothaker, L. E. (1993). *Multiple Comparison Procedures*. SAGE Publications.
- Hsu, J. C. (1996). *Multiple Comparisons: Theory and Methods*. Chapman & Hall/CRC.

---

**Next Lesson:** Lesson 14 - Repeated Measures and Mixed Models
